---
title: "venn_diagrams"
output: html_document
date: "2025-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacHumanPreSuspension)
library(MotrpacHumanPreSuspensionData)
library(dplyr)
library(tidyr)
library(ggVennDiagram)
library(conflicted)

config = jsonlite::fromJSON("~/config.json")
repo_local_dir = config$precovid_repo_path
devtools::load_all(config$r_package_path)
```

```{r}
full_da_results_all_tissues = MotrpacHumanPreSuspensionData::load_differential_analysis(single_matrix = TRUE,
                                                                                        combine_with_featgene = TRUE)
qc_norm_all_tissues = MotrpacHumanPreSuspensionData::load_qc(repo_local_dir)
pheno_loading = load_pheno()
```


```{r}
specific_tissue = "muscle"
full_da_results = full_da_results_all_tissues %>% 
  filter(tissue == specific_tissue) %>%
  mutate(contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
         # Split the left part on the period to get the group and timepoint
         Group = stringr::str_split_fixed(contrast_left, "\\.", 2)[,1],
         Timepoint = stringr::str_split_fixed(contrast_left, "\\.", 2)[,2]) %>%
  filter(!contrast_type == "baseline") %>%
  select(tissue, assay, Group, Timepoint, contrast_type, adj_p_value, feature_id) %>%
  pivot_wider(., names_from = c("contrast_type"), #aaaa i could also just split by group here? will do both
              values_from = adj_p_value) %>%
  select(-c(Endur_vs_Resist)) %>% #useless, will remove control only later
  group_by(tissue, assay, Timepoint, feature_id) %>%
  mutate(control_only = ifelse(is.na(control_only), first(na.omit(control_only)), control_only)) %>% #just filling in control values for both endur and resist 
  ungroup() %>%
  filter(!Group == "Control")
head(full_da_results[5:8])
```
```{r}
both_re_endur = full_da_results %>%
  mutate(
    exercise_with_controls = ifelse(exercise_with_controls <= 0.05, "Sig", "Non-sig"),
    exercise_no_controls = ifelse(exercise_no_controls <= 0.05, "Sig", "Non-sig"),
    control_only = ifelse(control_only <= 0.05, "Sig", "Non-sig")
  ) %>%
  pivot_wider(., names_from = c("Group"), 
              values_from = c("exercise_with_controls", "exercise_no_controls", "control_only"))
venn_data = list(
  "Exercise w/ Controls (Endur)" = both_re_endur %>% filter(exercise_with_controls_Endur == "Sig") %>% pull(feature_id),
  "Exercise w/ Controls (Resist)" = both_re_endur %>% filter(exercise_with_controls_Resist == "Sig") %>% pull(feature_id),
  "Exercise No Controls (Endur)" = both_re_endur %>% filter(exercise_no_controls_Endur == "Sig") %>% pull(feature_id),
  "Exercise No Controls (Resist)" = both_re_endur %>% filter(exercise_no_controls_Resist == "Sig") %>% pull(feature_id),
  "Control Only" = both_re_endur %>% filter(control_only_Endur == "Sig") %>% pull(feature_id) #so only 1 needed bc its the same from above
)
p = ggVennDiagram(venn_data, 
                  label = "count",
                  label_size = 4) +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() + 
  scale_x_continuous(expand = expansion(mult = .2)) + 
  ggtitle("ok this was worth a try but unreadable") +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),  
    text = element_text(size = 8)  
  )
print(p)
#so overlaps are based on multiple timepoints, if a feature is mediated by some mix of the categories at diff timepoints
```

```{r}
for(randomGroupCode in c("Resist", "Endur")){
  possible_combinations = full_da_results %>%
    filter(Group == randomGroupCode) %>%
    mutate(
      exercise_with_controls = ifelse(exercise_with_controls <= 0.05, "Sig", "Non-sig"),
      exercise_no_controls = ifelse(exercise_no_controls <= 0.05, "Sig", "Non-sig"),
      control_only = ifelse(control_only <= 0.05, "Sig", "Non-sig")
    ) 
  venn_data = list(
    "Exercise w/ Controls" = possible_combinations %>% filter(exercise_with_controls == "Sig") %>% pull(feature_id),
    "Exercise No Controls" = possible_combinations %>% filter(exercise_no_controls == "Sig") %>% pull(feature_id),
    "Control Only" = possible_combinations %>% filter(control_only == "Sig") %>% pull(feature_id)
  )
  p = ggVennDiagram(venn_data, 
                    label = "count",
                    label_size = 4) +
    scale_fill_gradient(low = "white", high = "blue") +
    theme_minimal() + 
    scale_x_continuous(expand = expansion(mult = .2)) + 
    ggtitle(paste("Significance in categories: ", randomGroupCode)) +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),  
      text = element_text(size = 8)  
    )
  print(p)
}
#so overlaps are based on multiple timepoints, if a feature is mediated by some mix of the categories at diff timepoints
```

Ok so now we can be a bit more selective and choose more meaningful stuff

```{r}
#specific lines of interest (note these are available in the above venns too, just highlighting categories)
# 1: YES in controls, NOT in exercise no controls, YES in exercise with controls "control mediated"  
# -> so this represents something exclusive to the controls (aka probably not circadian)
# Not actually sure if this is one of the "high importance" categories

# 2: NOT in controls, YES in exercise no controls, YES in exercise with controls "standard exercise effect"
# This one is probably the most standard category

# 3: NOT in controls, NOT in exercise no controls, YES in exercise with controls "dependent on combination"
# This one is situation where the controls and exercise go in opposite directions

# 4: YES in controls, YES in exercise no controls, NO in exercise with controls "probably circadian/biopsy/other mediated"
# This one is situations where controls and exercise go in same dir (more likely same at least)

for(randomGroupCode in c("Resist", "Endur")){
  selective_combos = full_da_results %>%
    filter(Group == randomGroupCode) %>%
    mutate(
      exercise_with_controls = ifelse(exercise_with_controls <= 0.05, "Sig", "Non-sig"),
      exercise_no_controls = ifelse(exercise_no_controls <= 0.05, "Sig", "Non-sig"),
      control_only = ifelse(control_only <= 0.05, "Sig", "Non-sig")
    ) %>%
    mutate(
      control_mediated = ifelse(control_only == "Sig" & exercise_no_controls == "Non-sig" & exercise_with_controls == "Sig", "Yes", "No"), #see cat 1
      standard_exercise_effect = ifelse(control_only == "Non-sig" & exercise_no_controls == "Sig" & exercise_with_controls == "Sig", "Yes", "No"), #see cat 2
      controls_ex_opposite_contribute = ifelse(control_only == "Non-sig" & exercise_no_controls == "Non-sig" & exercise_with_controls == "Sig", "Yes", "No"), #see cat 3
      probably_circadian = ifelse(control_only == "Sig" & exercise_no_controls == "Sig" & exercise_with_controls == "Non-sig", "Yes", "No") #see cat 4
    ) 
  venn_data = list(
    "Control Effect" = selective_combos %>% filter(control_mediated == "Yes") %>% pull(feature_id),
    "Exercise Effect" = selective_combos %>% filter(standard_exercise_effect == "Yes") %>% pull(feature_id),
    "Combined Effect" = selective_combos %>% filter(controls_ex_opposite_contribute == "Yes") %>% pull(feature_id),
    "Circadian Effect" = selective_combos %>% filter(probably_circadian == "Yes") %>% pull(feature_id)
  )
  p = ggVennDiagram(venn_data, 
                    label = "count", 
                    label_size = 4) +
    scale_fill_gradient(low = "white", high = "blue") +
    theme_minimal() + 
    scale_x_continuous(expand = expansion(mult = .2))
  ggtitle(paste("Overlaps bc of multiple timepoints - Mutually exclusive @ 1 tp"), randomGroupCode) +
    theme(
      plot.title = element_text(size = 10, hjust = 0.5),
      text = element_text(size = 8)
    )
  
  print(p)
}
#so overlaps are based on multiple timepoints, if a feature is mediated by some mix of the categories at diff timepoints
```
Based on some feedback, I don't think venn diagrams are really the best way to 
present some of the info. I'm going to try and expand to some barplots to possibly
jitter some different combos. 
```{r}
wide_by_contrasts = full_da_results_all_tissues %>% 
  mutate(contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
         # Split the left part on the period to get the group and timepoint
         Group = stringr::str_split_fixed(contrast_left, "\\.", 2)[,1],
         Timepoint = stringr::str_split_fixed(contrast_left, "\\.", 2)[,2]) %>% #add easy way to filter into group, tp
  filter(!contrast_type == "baseline") %>%
  select(tissue, assay, Group, Timepoint, contrast_type, adj_p_value, feature_id) %>%
  pivot_wider(., names_from = c("contrast_type"), #aaaa i could also just split by group here? will do both
              values_from = adj_p_value) %>%
  select(-c(Endur_vs_Resist)) %>% #useless, will remove control only later
  group_by(tissue, assay, Timepoint, feature_id) %>%
  mutate(control_only = ifelse(is.na(control_only), first(na.omit(control_only)), control_only)) %>% #just filling in control values for both endur and resist 
  ungroup() %>%
  filter(!Group == "Control") %>%
  mutate(assay = case_when(grepl("metab", assay) ~ "metab", 
                           grepl("prot-ol|prot-pr", assay) ~ "prot", 
                           TRUE ~ assay))  %>%
  filter(!(grepl("during", Timepoint) | grepl("10_min", Timepoint))) %>%
  mutate(Timepoint = factor(Timepoint,
                            levels = c("post_15_30_45_min", "post_3.5_4_hr", "post_24_hr")))
#remove timepoint for blood samples
#now we actually categorize things similar to before. 
wide_by_contrasts = wide_by_contrasts %>%
  # filter(!(exercise_with_controls > 0.05 & exercise_no_controls > 0.05 & control_only > 0.05)) %>%
  mutate(
    exercise_with_controls_cat = ifelse(exercise_with_controls <= 0.05, "Sig", "Non-sig"),
    exercise_no_controls_cat = ifelse(exercise_no_controls <= 0.05, "Sig", "Non-sig"),
    control_only_cat = ifelse(control_only <= 0.05, "Sig", "Non-sig")
  ) %>%
  mutate(
    category = case_when(
      (control_only_cat == "Sig" & exercise_no_controls_cat == "Non-sig" & exercise_with_controls_cat == "Sig") ~ "Control Effect",
      (exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Sig") ~ "Exercise Effect", #we're deciding to include regardless of control_only_cat
      #which could be a bit arbitrary, but i 
      (control_only_cat == "Non-sig" & exercise_no_controls_cat == "Non-sig" & exercise_with_controls_cat == "Sig") ~ "Combined group effect",
      (control_only_cat == "Sig" & exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Non-sig") ~ "Circadian/Non exercise",
      TRUE ~ "Other"
    )
  ) 

wide_all_save = wide_by_contrasts #saving for false negatives analysis
wide_by_contrasts = wide_by_contrasts %>%
  filter(!category == "Other")
```

finding good examples of items in ea category

Starting with Standard Exercise Effect: 
```{r}
#this one is our generic example, just going to keep using PGC1a, just using muscle
plot_single_feature(feature = "PPARGC1A", 
                    selected_tissues = "muscle",
                    repo_local_dir = repo_local_dir,
                    da_object = full_da_results_all_tissues,
                    qc_object = qc_norm_all_tissues,
                    pheno_object = pheno_loading)
```

```{r}
control_med_examples = wide_by_contrasts %>% 
  filter(category == "Control Effect") %>%
  group_by(feature_id) %>%
  filter(all(c("Resist", "Endur") %in% Group)) %>%
  ungroup()

temp_da_blood = full_da_results_all_tissues %>% filter(tissue == "blood")
temp_qc_blood = list(); temp_qc_blood$blood = qc_norm_all_tissues$blood

# 
# plot_single_feature(feature = "PER1", 
#                     repo_local_dir = repo_local_dir,
#                     da_object = full_da_results_all_tissues,
#                     qc_object = qc_norm_all_tissues,
#                     pheno_object = pheno_loading,
#                     selected_tissues = "blood")

plot_single_feature(feature = "PER1",
                    repo_local_dir = repo_local_dir,
                    da_object = temp_da_blood,
                    qc_object = temp_qc_blood,
                    pheno_object = pheno_loading,
                    selected_tissues = "blood")
#maybe not the greatest example?
```

```{r}
circadian_type = wide_by_contrasts %>% 
  filter(category == "Circadian/Non exercise") %>%
  group_by(feature_id) %>%
  filter(all(c("Resist", "Endur") %in% Group)) %>%
  ungroup()

plot_single_feature(feature = "Alloisoleucine", 
                    repo_local_dir = repo_local_dir,
                    da_object = full_da_results_all_tissues,
                    qc_object = qc_norm_all_tissues,
                    selected_tissues = "adipose",
                    pheno_object = pheno_loading)
#adipose is a good example here, I think
```
```{r}
plot_single_feature(feature = "ENSG00000168209.6", 
                    repo_local_dir = repo_local_dir,
                    da_object = full_da_results_all_tissues,
                    qc_object = qc_norm_all_tissues,
                    selected_tissues = "adipose",
                    pheno_object = pheno_loading)
#PER2 in adipose is another good example
```
```{r}
synergistic = wide_by_contrasts %>% 
  filter(category == "Significant together with controls") %>%
  group_by(feature_id) %>%
  filter(all(c("Resist", "Endur") %in% Group)) %>%
  ungroup() %>%
  arrange(exercise_with_controls)

plot_single_feature(feature = "Q86TB9_S177s", 
                    repo_local_dir = repo_local_dir,
                    da_object = full_da_results_all_tissues,
                    qc_object = qc_norm_all_tissues,
                    pheno_object = pheno_loading)

#slightly weaker example but I think its fine. This group is kinda the weakest
#or least exciting example bc it just helps with power a bit
```

```{r}
ggplot(wide_by_contrasts, aes(x = assay, fill = category)) +
  geom_bar(stat = "count", position = "stack") + #making the y axis counts
  facet_grid(Group ~ tissue, scales = "free_x") +
  theme_minimal() +
  labs(x = "Assay", y = "Count", title = "Assay + Tissue, still combining all tp") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#ok attempt at different timepoint inclusion
wide_by_contrasts <- wide_by_contrasts %>%
  mutate(assay_timepoint = factor(paste(assay, Timepoint, sep = " - "),
                                  levels = unique(paste(rep(levels(factor(assay)), 
                                                            each = length(levels(factor(Timepoint)))),
                                                        rep(levels(factor(Timepoint)), 
                                                            times = length(levels(factor(assay)))),
                                                        sep = " - ")))) %>%
  mutate(tissue_group = factor(paste(tissue, Group, sep = " - "),
                               levels = unique(paste(rep(levels(factor(tissue)), 
                                                         each = length(levels(factor(Group)))),
                                                     rep(levels(factor(Group)), 
                                                         times = length(levels(factor(tissue)))),
                                                     sep = " - ")))) 


ggplot(wide_by_contrasts, aes(x = Timepoint, fill = category)) +
  geom_bar(stat = "count", position = "stack") +
  facet_grid(assay ~ tissue_group, scales = "free_y") +  # Facet by ome (rows) and tissue-group (columns)
  theme_minimal() +
  labs(x = "Timepoint", y = "Count", title = "Selected Categories by Ome/Timepoint") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Keep x-axis readable

#notably: i can't really label all the timepoints + assays very easily
```

So the above all describes 4 specific patterns of interest that we can gain information
about by using the controls.

One other way to describe the dataset is through analysis of inference results
that would otherwise be obtained if an experiment exactly like pre-cawg 
was done but without the controls. 

So we have the categories:
1: True Positives - This includes all of the "standard exercise effect" category,
essentially any feature that matches 

2: False Positives - This includes all of the "Non exercise induced"

3: False Negatives - This includes "Control mediated" category,
also includes "significant together with controls" category

4: True Negatives - This is not described in any of the above categories,
but it's just any situation where a feature is considered non-significant in 
the delta-delta and also in the exercise without controls section. 

The 
```{r}
numbers = wide_all_save %>%
  group_by(exercise_no_controls_cat, exercise_with_controls_cat, control_only_cat) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  distinct(exercise_no_controls_cat, exercise_with_controls_cat, control_only_cat, n) %>%
  select(exercise_with_controls_cat, everything())
```

```{r}
true_positives = wide_all_save %>%
  filter(exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Sig")
nrow(true_positives)
```

```{r}
false_negatives = wide_all_save %>%
  filter(exercise_no_controls_cat == "Non-sig" & exercise_with_controls_cat == "Sig")
nrow(false_negatives)
```

```{r}
false_positives = wide_all_save %>%
  filter(exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Non-sig")
nrow(false_positives)
```
And so it may seem alarming that "false-positives" is the largest category, but
in most cases, this is just a power/threshold issue--the estimation needed to 
evaluate the controls, and the relatively more limited number of controls makes 
many features simply not achieve the `adj_p < 0.05` threshold. So even if the
features are in the same 'direction' but one doesn't reach a threshold, it 
comes out to look like its a "false negative"

```{r}
true_negatives = wide_all_save %>%
  filter(exercise_no_controls_cat == "Non-sig" & exercise_with_controls_cat == "Non-sig")
nrow(true_negatives)
```

Finding examples of RE vs EE
```{r}
re_v_ee_index = full_da_results_all_tissues %>%
  mutate(contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
         # Split the left part on the period to get the group and timepoint
         Group = stringr::str_split_fixed(contrast_left, "\\.", 2)[,1],
         Timepoint = stringr::str_split_fixed(contrast_left, "\\.", 2)[,2]) %>%
  filter(contrast_type == "Endur_vs_Resist") %>%
  select(assay, feature_id, z.std, adj_p_value, tissue, contrast_type, Timepoint) %>%
  filter(!Timepoint %in% c("during_20_min", "during_40_min", "post_10_min")) %>%
  mutate(assay = case_when(grepl("metab", assay) ~ "metab",
                           TRUE ~ assay)) %>%
  filter(adj_p_value < 0.05)

# pdf(file = "~/Downloads/tempreee.pdf", width = 10)
re_v_ee_index %>%
  count(assay, tissue, Timepoint, name = "n_features") %>%
  mutate(Timepoint = factor(Timepoint, levels =c("post_15_30_45_min", "post_3.5_4_hr","post_24_hr"))) %>%
  ggplot(aes(x = tissue, y = assay, label = n_features)) +
  geom_tile(fill = "white", color = "grey70") +
  geom_text(size = 4) +
  facet_wrap(~ Timepoint, ncol = 1) +
  labs(x = "Tissue", y = "Assay", title = "Significant Differences for RE vs EE") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
dev.off()


```

```{r}
re_v_ee_index_tests  = re_v_ee_index 

for(feat in re_v_ee_index_tests$feature_id[1:10]){
  p = plot_single_feature(feature = feat, 
                          repo_local_dir = repo_local_dir,
                          da_object = full_da_results_all_tissues,
                          qc_object = qc_norm_all_tissues,
                          pheno_object = pheno_loading)
  print(p)
}

# pdf(file = "~/Downloads/tempmyog.pdf", width = 10)
plot_single_feature(feature = "MYOG", 
                    repo_local_dir = repo_local_dir,
                    da_object = full_da_results_all_tissues,
                    qc_object = qc_norm_all_tissues,
                    pheno_object = pheno_loading)
dev.off()
```




