---
title: "mofa_exploration"
output: html_document
date: "2026-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MOFA2)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(here)
library(dplyr)
library(tidyr)

mofa_out = file.path(here(), "revisions", "landscape", "mofa")
dir.create(mofa_out, recursive = TRUE, showWarnings = FALSE)
```

The goal of this is to run MOFA on the SCION input matrices to identify latent factors
that jointly explain variation across transcriptomics and phosphoproteomics in muscle.
MOFA is run separately for ADUResist and ADUControl samples.

# Overview

Multi-Omics Factor Analysis (MOFA) omic layers: transcriptomics, phosphoproteomics.

## Load SCION matrices

Phosphoproteomics (regulators) and transcriptomics (targets) matrices from the muscle
SCION network.

```{r load-data}
phospho_mat = .load_scion_matrixes(desired_matrixes = "muscle.prot-ph",
                                   subset_TFs = TRUE,
                                   subset_DE = FALSE) 

trnscrpt_mat = .load_scion_matrixes(desired_matrixes = "muscle.transcript-rna-seq",
                                    subset_TFs = FALSE,
                                    subset_DE = TRUE) 

rownames(phospho_mat)  = sub("\\.\\..*", "", rownames(phospho_mat))
rownames(trnscrpt_mat) = sub("\\.\\..*", "", rownames(trnscrpt_mat))

cat("Phospho dimensions:", dim(phospho_mat), "\n")
cat("Transcript dimensions:", dim(trnscrpt_mat), "\n")
```


Per MOFA instructions, we select the top features by variability. 

The number of features we select is somewhat arbitrary, but I've done used 1000 per the instructions.

This actually does NOT include the TFEB transcript. 
```{r}
select_top_var = function(mat, n) {
  vars = apply(mat, 1, var)
  top_idx = order(vars, decreasing = TRUE)[seq_len(n)]
  mat[top_idx, ]
}
phospho_mat = select_top_var(phospho_mat, 1000)
trnscrpt_mat = select_top_var(trnscrpt_mat, 1000)
```

# Prepare shared samples and metadata

```{r build-df}
shared_cols = intersect(colnames(phospho_mat), colnames(trnscrpt_mat))
cat("Shared samples:", length(shared_cols), "\n")

# Load metadata and assign group from randomGroupCode
sample_metadata = load_qc(selected_tissues = "muscle",
                          selected_omes = "transcript-rna-seq")
sample_metadata = sample_metadata$muscle$`transcript-rna-seq`$sample_metadata %>%
  mutate(sample = as.character(interaction(pid, Timepoint, sep = ".."))) %>%
  select(sample, Sex, Timepoint, randomGroupCode) %>%
  mutate(group = "single_group") %>%
  filter(sample %in% shared_cols)
rownames(sample_metadata) = NULL

cat("Group counts:\n")
print(table(sample_metadata$group))

# Helper: build long-format df for a subset of samples
mat_to_long = function(mat, view_name, samples) {
  mat[, samples, drop = FALSE] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature") %>%
    pivot_longer(-feature, names_to = "sample", values_to = "value") %>%
    mutate(view = view_name)
}

run_mofa_for_group = function(grp, outfile) {
  meta = sample_metadata %>% filter(randomGroupCode == grp)
  samps = meta$sample

  mofa_df = bind_rows(
    mat_to_long(phospho_mat,  "Phosphoproteomics", samps),
    mat_to_long(trnscrpt_mat, "Transcriptomics",   samps)
  )
  
  obj = create_mofa_from_df(mofa_df)
  samples_metadata(obj) = meta
  
  data_opts = MOFA2::get_default_data_options(obj)
  data_opts$scale_views = TRUE
  
  model_opts = MOFA2::get_default_model_options(obj)
  model_opts$num_factors = 10
  
  train_opts = MOFA2::get_default_training_options(obj)
  
  obj = MOFA2::prepare_mofa(obj,
                            data_options = data_opts,
                            model_options = model_opts,
                            training_options = train_opts)
  
  MOFA2::run_mofa(obj, outfile = outfile, use_basilisk = TRUE)
}
```

# Run MOFA per group

```{r train-resist, results='hide', message=FALSE}
mofa_resist = run_mofa_for_group(
  "ADUResist",
  outfile = file.path(mofa_out, "mofa_ADUResist.out")
)
```

```{r train-control, results='hide', message=FALSE}
mofa_endur = run_mofa_for_group(
  "ADUEndur",
  outfile = file.path(mofa_out, "mofa_ADUEndur.out")
)
```

# Downstream exploration

## Variance explained

```{r var-explained-resist}
cat("--- ADUResist ---\n")
plot_variance_explained(mofa_resist, plot_total = TRUE)
```

```{r var-explained-control}
cat("--- ADUEndur ---\n")
plot_variance_explained(mofa_endur, plot_total = TRUE)
```

## Factor plots

```{r factor-timepoint-resist, fig.width=8, fig.height=6}
plot_factor(mofa_resist,  factors = 1:5, color_by = "Timepoint", shape_by = "Sex")
```

```{r factor-timepoint-control, fig.width=8, fig.height=6}
plot_factor(mofa_endur, factors = 1:5, color_by = "Timepoint", shape_by = "Sex")
```

```{r factor-sex-resist, fig.width=8, fig.height=6}
plot_factor(mofa_resist,  factors = 1:5, color_by = "Sex", shape_by = "Timepoint")
```

```{r factor-sex-control, fig.width=8, fig.height=6}
plot_factor(mofa_endur, factors = 1:5, color_by = "Sex", shape_by = "Timepoint")
```

## Factor correlations

```{r factor-cor-resist}
plot_factor_cor(mofa_resist)
```

```{r factor-cor-control}
plot_factor_cor(mofa_endur)
```

Next we want to find the weighting of genes of interest to see how some of the methods align.

```{r gene-weights}
genes_of_interest = c("TFEB", "MEF2A", "NFIC", "BAG3", "HIPK3", "HIPK2", "ZEB1", "FOXO1", "FOXO3")

target_features = HUMAN_FEATURE_TO_GENE %>%
  filter(assay %in% c("transcript-rna-seq", "prot-ph")) %>%
  select(feature_id, gene_symbol)

get_gene_weights = function(mofa_trained, grp) {
  var_exp = get_variance_explained(mofa_trained)$r2_per_factor[[1]] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("factor") %>%
    pivot_longer(-factor, names_to = "view", values_to = "factor_var_pct_exp")

  get_weights(mofa_trained, as.data.frame = TRUE) %>%
    group_by(factor, view) %>%
    mutate(rank = rank(-abs(value))) %>%
    ungroup() %>%
    left_join(target_features, by = c("feature" = "feature_id")) %>%
    left_join(var_exp, by = c("factor", "view")) %>%
    mutate(group = grp) %>%
    arrange(rank)
}

full_gene_weight_results = bind_rows(
  get_gene_weights(mofa_resist, "ADUResist"),
  get_gene_weights(mofa_endur,  "ADUEndur")
)

write.csv(full_gene_weight_results, file = file.path(here(), "revisions", "landscape", "mofa", "full_revision_gene_weight_results.csv"),
          row.names = FALSE)

interested_gene_weights = full_gene_weight_results %>% 
  filter(gene_symbol %in% genes_of_interest) %>%
  arrange(gene_symbol, rank)

write.csv(interested_gene_weights, file = file.path(here(), "revisions", "landscape", "mofa", "scion_revision_gene_weight_results.csv"),
          row.names = FALSE)

interested_gene_weights
```
