---
title: "sex_effects_acute"
output: html_document
date: "2026-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(dplyr)
library(ggplot2)

source(here::here("revisions/landscape/sex_sensitivity_analysis/sex_differences_single_feat.R"))
source(here::here("revisions/landscape/sex_sensitivity_analysis/plot_mvsf_da_matrix.R"))
```

## Review Comment: 

The cohort is 72% female, yet the manuscript fails to analyze whether molecular responses differ by sex. Given the well-established biological sexual dimorphism in exercise responses, this is a major oversight. Consequently, it is unclear whether the results are generalizable or primarily reflect the physiology of the female majority. The authors should perform and report a sex-stratified analysis or, at minimum, discuss this potential bias as a significant limitation.


## How to address this comment

The way we currently model the DA, we are essentially averaging any changes across male and female participants. We do have a covariate for sex, which is estimated across all groups + timepoints, but this does not really answer if we see a sexual dimorphism for the response for a specific group and timepoint, and if this dimorphism is being "averaged-out" by our modeling stratey.  

Instead, we want to try to estimate the effect of exercise in a specific sex and specific exercise group independently, and see if the directions of change are similar between the sex-group-timepoint combinations. 

### **Caveats** 

1. The major reason we avoided this is because in most cases, the number of samples for a sex-group-timepoint can be very limited (especially for smaller male groups). Any conclusions have to be made with the understanding some of the potential differences (or lack of differences) may be due to sampling limitations

2. The MoTrPAC exec has limited sex-difference analysis (outside of baseline comparisons) for this above reason.

Therefore, I am proposing we describe these to reviewers + clarify that any results are descriptive in nature only and a point of possible comparison. 

The sex-based differential analysis is described in `MotrpacHumanPreSuspensionAnalysis/data-raw/generate_differential_analysis.R`. Because there are many more parameters to estimate, I am no longer adjusting for initial clinical site, bc the model matrix will not be full rank. 


### What does the data look like in the first place?

So I'm going to start with the same initial filtering criteria we use in the landscape. Each cell must have at least 3 observations, e.g. every sex-group-timepoint must have 3 male participants + 3 female. 

```{r, include = FALSE}
all_dataset = load_qc()

split_tissues = unlist(all_dataset, recursive = FALSE)
split_tissue_assay = unlist(split_tissues, recursive = FALSE)
qc_norm_only = split_tissue_assay[grep("qc_norm", names(split_tissue_assay))]

#makes a big list of tissues, assays
samples_per_plat = data.frame(
  list_name = names(qc_norm_only),
  vialLabels = sapply(qc_norm_only, function(x) paste(colnames(x), collapse = ";"))
) %>%
  mutate(
    tissue = sapply(strsplit(list_name, "\\."), `[`, 1),
    assay = sapply(strsplit(list_name, "\\."), `[`, 2)
  ) %>%
  select(-list_name)

#then uses the pheno file to annotate group and timepoint info and split the vector
qc_norm_counts = lapply(seq_len(nrow(samples_per_plat)), function(row) {
  vials_vec = strsplit(samples_per_plat[row, "vialLabels"], ";")[[1]] %>%
    base::trimws()
  
  pheno$data %>%
    filter(vialLabel %in% vials_vec) %>%
    group_by(randomGroupCode, Timepoint, Sex) %>%
    summarise(
      n = n(),
      vialLabels = paste(vialLabel, collapse = ","),
      .groups = "drop"
    ) %>%
    mutate(
      tissue = samples_per_plat$tissue[row],
      assay = samples_per_plat$assay[row]
    )
}) %>%
  bind_rows() %>%
  select(tissue, assay, randomGroupCode, Sex, Timepoint, n, vialLabels)

qualifying_omes_tissues = qc_norm_counts %>%
  group_by(tissue, assay) %>%
  mutate(flag_low_n = any(n < 3)) %>%
  filter(!flag_low_n) %>%
  distinct(tissue, assay)
```

```{r}
qualifying_omes_tissues %>%
  mutate(assay = case_when(
    grepl("metab", assay) ~ "metab",
    TRUE ~ assay
  )) %>%
  distinct() 
```

Note: I am going to be focusing on metabolomics for this initial report. 

```{r, include = FALSE}
sex_differential_analysis = list.files(file.path("~/Downloads", "data", "tmp", "freeze_DA"),
                                       full.names = TRUE)

read_sex_diff = lapply(sex_differential_analysis, function(f) {
  tissue = dplyr::case_when(
    grepl("muscle", f) ~ "muscle",
    grepl("blood|plasma", f) ~ "blood"
  )
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    dplyr::mutate(tissue = tissue)
}) %>%
  bind_rows() 
```

```{r, fig.width = 14, fig.height = 3}
direct_m_vs_f = read_sex_diff %>%
    filter((grepl("Male", contrast) & grepl("Female", contrast)))

plot_mvsf_da_matrix(direct_m_vs_f)
```

**Note:** The lower % of DA features relative to the main analysis is also relfecting reduced statistical power due to lower n and ability to estimate proper point estimates of sex-group-timepoint. 

```{r}
sex_da_combined = read_sex_diff %>%
  filter(!(grepl("Male", contrast) & grepl("Female", contrast))) 

sex_da_annotated = sex_da_combined %>%
  dplyr::mutate(contrast_short = gsub("sex_group_timepoint|[()|]", "", contrast),
                contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
                sex_comparison = stringr::str_split_fixed(contrast_left, "\\.", 3)[,1],
                randomGroupCode = stringr::str_split_fixed(contrast_left, "\\.", 3)[,2],
                Timepoint = sub("^[^.]+\\.[^.]+\\.", "", contrast_left)) %>%
  dplyr::select(-contrast_left) %>%
  mutate(Timepoint = factor(Timepoint,
                            levels = c("pre_exercise",
                                       "during_20_min",
                                       "during_40_min",
                                       "post_10_min",
                                       "post_15_30_45_min",
                                       "post_3.5_4_hr",
                                       "post_24_hr"))) %>%
  rename(platform = assay) %>%
  select(tissue, platform, feature_id, logFC, CI.L, CI.R, randomGroupCode, sex_comparison, Timepoint) 

precawg_analysis = load_differential_analysis(single_matrix = TRUE) %>%
  filter(contrast_type == "exercise_with_controls") %>%
  select(tissue, platform, feature_id, logFC, CI.L, CI.R, randomGroupCode, Timepoint)

precawg_and_sex_diff = precawg_analysis %>%
  right_join(sex_da_annotated,
             by = c("tissue", "platform", "feature_id", "randomGroupCode", "Timepoint"),
             suffix = c("_precawg", "_sex_diff"))
```


# Metabs included in PreCAWG analysis

1. Carnitines, Glutamic acid, Citrate, Leucine + isoleucine, lactate

```{r, results = "asis"}
manuscript_metabs = c("Lactic acid", "Leucine", "Isoleucine", "Glutamic acid")

for (feat in manuscript_metabs) {
  tryCatch({
    cat("\n\n###", feat, "\n\n")
    print(sex_differences_single_feature(feat))
    print(sex_differences_logfc_feature(feat))
  }, error = function(e) {
    message("Skipping ", feat, ": ", conditionMessage(e))
  })
}
```


## Top M vs F DA features

```{r, include = FALSE}
top_mvsf_features = direct_m_vs_f %>%
  group_by(feature_id) %>%
  slice_min(adj_p_value, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(adj_p_value < 0.05) %>%
  arrange(adj_p_value) 
```

```{r, results = "asis"}
for (feat in top_mvsf_features$feature_id[1:5]) {
  tryCatch({
    cat("\n\n###", feat, "\n\n")
    print(sex_differences_single_feature(feat))
    print(sex_differences_logfc_feature(feat))
  }, error = function(e) {
    message("Skipping ", feat, ": ", conditionMessage(e))
  })
}
```


