---
title: "sex_effects_acute"
output: html_document
date: "2026-03-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(tibble)
library(biomaRt)

source(here::here("revisions/landscape/sex_sensitivity_analysis/sex_sensitivity_helpers.R"))
```

## Review Comment:

The cohort is 72% female, yet the manuscript fails to analyze whether molecular responses differ by sex. Given the well-established biological sexual dimorphism in exercise responses, this is a major oversight. Consequently, it is unclear whether the results are generalizable or primarily reflect the physiology of the female majority. The authors should perform and report a sex-stratified analysis or, at minimum, discuss this potential bias as a significant limitation.

## Draft Reviewer Response

We thank the reviewer for raising this important point. We acknowledge that the cohort composition (72% female) warrants careful consideration of potential sex-specific effects on the molecular responses we report. In our primary analysis, sex was included as a covariate, which does not directly test whether the exercise-induced response differs between males and females. To address this concern, we performed a supplementary sex-stratified sensitivity analysis in which we estimated exercise-induced log fold-changes in male and female group-timepoint combinations. This analysis was restricted to tissue-assay platforms where at least three participants of each sex were present in every group-timepoint cell, to ensure estimates were not unduly driven by individual observations. This limits our sensitivity analysis to metabolomics and transcriptomics for every tissue. 

For the specific molecular features highlighted in the manuscript — including key transcripts (e.g., NR4A1, PPARGC1A, MYC, FOXO1/3) and metabolites (e.g., lactate, leucine, isoleucine, glutamic acid) — the direction and approximate magnitude of exercise-induced change were consistent between male and female participants, supporting the generalizability of our primary findings. We note that the reduced per-sex sample sizes yield wider confidence intervals and lower statistical power than the full-cohort analysis, so this sex-stratified analysis should be interpreted as descriptive and hypothesis-generating rather than confirmatory. We have added a dedicated section to the manuscript describing this analysis and its limitations, and have updated the limitations section to explicitly discuss the unbalanced sex composition of the cohort.


## How to address this comment

The way we currently model the DA, we are essentially averaging any changes across male and female participants. We do have a covariate for sex, which is estimated across all groups + timepoints, but this does not really answer if we see a sexual dimorphism for the response for a specific group and timepoint, and if this dimorphism is being "averaged-out" by our modeling stratey.  

Instead, we want to try to estimate the effect of exercise in a specific sex and specific exercise group independently, and see if the directions of change are similar between the sex-group-timepoint combinations. 

### **Caveats** 

1. The major reason we avoided this is because in most cases, the number of samples for a sex-group-timepoint can be very limited (especially for smaller male groups). Any conclusions have to be made with the understanding some of the potential differences (or lack of differences) may be due to sampling limitations

Therefore, I am proposing we describe these to reviewers + clarify that any results are descriptive in nature only and a point of possible comparison. 

The sex-based differential analysis is described in `MotrpacHumanPreSuspensionAnalysis/data-raw/generate_differential_analysis.R`. Because there are many more parameters to estimate, I am no longer adjusting for initial clinical site, bc the model matrix will not be full rank. 

### What does the data look like in the first place?

Let's start by visualizing the distribution of Male vs Female participants per group and timepoint first. 

So I'm going to start with the same initial filtering criteria we use in the landscape. Each cell must have at least 3 observations, e.g. every sex-group-timepoint must have 3 male participants + 3 female. 

```{r, include = FALSE}
all_dataset = load_qc()

split_tissues = unlist(all_dataset, recursive = FALSE)
split_tissue_assay = unlist(split_tissues, recursive = FALSE)
qc_norm_only = split_tissue_assay[grep("qc_norm", names(split_tissue_assay))]

#makes a big list of tissues, assays
samples_per_plat = data.frame(
  list_name = names(qc_norm_only),
  vialLabels = sapply(qc_norm_only, function(x) paste(colnames(x), collapse = ";"))
) %>%
  dplyr::mutate(
    tissue = sapply(strsplit(list_name, "\\."), `[`, 1),
    assay = sapply(strsplit(list_name, "\\."), `[`, 2)
  ) %>%
  dplyr::select(-list_name)

#then uses the pheno file to annotate group and timepoint info and split the vector
qc_norm_counts = lapply(seq_len(nrow(samples_per_plat)), function(row) {
  vials_vec = strsplit(samples_per_plat[row, "vialLabels"], ";")[[1]] %>%
    base::trimws()
  
  pheno$data %>%
    dplyr::filter(vialLabel %in% vials_vec) %>%
    dplyr::group_by(randomGroupCode, Timepoint, Sex) %>%
    dplyr::summarise(
      n = dplyr::n(),
      vialLabels = paste(vialLabel, collapse = ","),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      tissue = samples_per_plat$tissue[row],
      assay = samples_per_plat$assay[row]
    )
}) %>%
  dplyr::bind_rows() %>%
  dplyr::select(tissue, assay, randomGroupCode, Sex, Timepoint, n, vialLabels) %>%
  dplyr::filter(!grepl("metab", assay) | assay == "metab-u-hilicpos") 


qualifying_omes_tissues = qc_norm_counts %>%
  dplyr::filter(randomGroupCode != "ADUControl") %>%
  dplyr::group_by(tissue, assay) %>%
  dplyr::mutate(flag_low_n = any(n < 3)) %>%
  dplyr::filter(!flag_low_n) %>%
  dplyr::distinct(tissue, assay)
```

## Discusison for metab platform to include. 

Because there are multiple metabolomics platforms, I choose to visualize just Hilic Positive, although the metab platforms generally are very similar in terms of n. 


# Sex Distribution by Timepoint and Group

Visualize # of participants per sex-group-timepoint. 


```{r, include = FALSE}
timepoint_short = c(
  "pre_exercise"      = "Pre",
  "during_20_min"     = "D20",
  "during_40_min"     = "D40",
  "post_10_min"       = "P10",
  "post_15_30_45_min" = "P1545",
  "post_3.5_4_hr"     = "P35",
  "post_24_hr"        = "P24"
)

group_colors = c(EE = "#d95f02", RE = "#1b9e77", CON = "#7570b3")
sex_colors   = c(Female = "#E87461", Male = "#5B9BD5")

sex_dist_plot = qc_norm_counts %>%
  dplyr::mutate(
    Group = dplyr::recode(randomGroupCode,
                          "ADUEndur" = "EE", "ADUResist" = "RE", "ADUControl" = "CON"),
    Group = factor(Group, levels = c("EE", "RE", "CON")),
    tp = dplyr::recode(as.character(Timepoint), !!!timepoint_short),
    tp = factor(tp, levels = unname(timepoint_short)),
    Ome = dplyr::case_when(
      grepl("metab", assay)  ~ "Metabolomics",
      assay == "prot-ph"     ~ "Phosphoproteomics",
      assay %in% c("prot-pr", "prot-ol") ~ "Proteomics",
      grepl("rna", assay)    ~ "Transcriptomics",
      grepl("atac", assay)   ~ "ATAC",
      grepl("methyl", assay) ~ "Methylation",
      TRUE ~ assay
    )
  )

```

```{r sex-dist-plot, fig.width = 14, fig.height = 5}
ggplot(sex_dist_plot, aes(x = tp, y = n, fill = Sex)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.25, width = 0.75) +
  geom_text(
    aes(label = n),
    position = position_dodge(width = 0.75),
    vjust = -0.35,
    size = 2.5
  ) +
  facet_grid(Ome ~ tissue + Group, scales = "free_y") +
  scale_fill_manual(values = sex_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(
    x = "Timepoint", y = "N samples",
    title = "Sample count by sex, tissue, exercise group, and timepoint"
  ) +
  theme_bw(base_size = 10) +
  theme(
    strip.text       = element_text(size = 7, face = "bold"),
    axis.text.x      = element_text(size = 8, angle = 40, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```


# Targeted Data Overview: Sample-Level PCA by Sex

```{r pca-sex, fig.width = 14, fig.height = 5, message = FALSE, warning = FALSE}
tp_levels = c("pre_exercise", "during_20_min", "during_40_min",
              "post_10_min", "post_15_30_45_min", "post_3.5_4_hr", "post_24_hr")
tp_colors_pca = MotrpacHumanPreSuspensionAnalysis::HUMAN_ACUTE_TIMEPOINT_COLORS
tp_colors_pca = tp_colors_pca[tp_levels]

pca_plots = list()

for (i in seq_len(nrow(qualifying_omes_tissues))) {
  tiss = qualifying_omes_tissues$tissue[i]
  ome  = qualifying_omes_tissues$assay[i]
  
  ome_data = all_dataset[[tiss]][[ome]]
  if (is.null(ome_data)) next
  
  p = make_pca_panel(tiss, ome, ome_data, tp_levels, tp_colors_pca)
  if (!is.null(p)) {
    if (is.null(pca_plots[[tiss]])) pca_plots[[tiss]] = list()
    pca_plots[[tiss]][[ome]] = p
  }
}

for (tiss in names(pca_plots)) {
  panels = pca_plots[[tiss]]
  print(
    patchwork::wrap_plots(panels, ncol = min(5, length(panels))) +
      patchwork::plot_layout(guides = "collect") +
      patchwork::plot_annotation(
        title = tiss,
        theme = theme(plot.title = element_text(size = 12, face = "bold"))
      ) &
      theme(legend.position = "bottom")
  )
}
```

```{r, include = FALSE}
config = jsonlite::fromJSON("~/config.json")
repo_local_dir = config$precovid_repo_path
sex_differential_analysis = list.files(file.path(repo_local_dir, "data", "tmp", "freeze_DA"),
                                       full.names = TRUE)

read_sex_diff = lapply(sex_differential_analysis, function(f) {
  tissue = dplyr::case_when(
    grepl("muscle", f) ~ "muscle",
    grepl("blood|plasma", f) ~ "blood",
    grepl("adipose", f) ~ "adipose"
  )
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    dplyr::mutate(tissue = tissue)
}) %>%
  dplyr::bind_rows()

precawg_analysis = load_differential_analysis(single_matrix = TRUE) %>%
  dplyr::filter(contrast_type == "exercise_with_controls") %>%
  dplyr::select(tissue, assay, feature_id, logFC, CI.L, CI.R, adj_p_value, randomGroupCode, Timepoint)

# this is the set of plots where I visualize the log FC for male and female participants separately. Not including the direct comparison of male and female exercise group changes. 

sex_single_da_annotated = read_sex_diff %>%
  dplyr::filter(!(grepl("Male", contrast) & grepl("Female", contrast))) %>%
  dplyr::mutate(contrast_short = gsub("sex_group_timepoint|[()|]", "", contrast),
                contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
                sex_comparison = stringr::str_split_fixed(contrast_left, "\\.", 3)[,1],
                randomGroupCode = stringr::str_split_fixed(contrast_left, "\\.", 3)[,2],
                Timepoint = sub("^[^.]+\\.[^.]+\\.", "", contrast_left)) %>%
  dplyr::select(-contrast_left) %>%
  dplyr::mutate(Timepoint = factor(Timepoint,
                                   levels = c("pre_exercise",
                                              "during_20_min",
                                              "during_40_min",
                                              "post_10_min",
                                              "post_15_30_45_min",
                                              "post_3.5_4_hr",
                                              "post_24_hr"))) %>%
  dplyr::rename(platform = assay) %>%
  dplyr::mutate(assay = ifelse(grepl("metab", platform), "metab", platform)) %>%
  dplyr::select(tissue, assay, platform, feature_id, logFC, CI.L, CI.R, adj_p_value, sex_comparison, randomGroupCode, Timepoint) %>%
  dplyr::left_join(
    precawg_analysis %>% dplyr::select(tissue, feature_id, Timepoint, randomGroupCode, pre_cawg_adj_p = adj_p_value),
    by = c("tissue", "Timepoint", "randomGroupCode", "feature_id")
  )
```

# Male vs. Female logFC Scatter

Per-facet correlation using significant features only, computed per Group × Timepoint.
Note: correlations are of logFC point estimates (exercise vs. pre-exercise) restricted
to EE and RE; control-group comparisons are excluded given the low per-sex n in controls
makes those point estimates unreliable.

```{r scatter-prep}
scatter_wide = sex_single_da_annotated %>%
  dplyr::select(tissue, assay, platform, feature_id, randomGroupCode, Timepoint,
                sex_comparison, logFC, CI.L, CI.R, adj_p_value, pre_cawg_adj_p) %>%
  tidyr::pivot_wider(
    names_from  = sex_comparison,
    values_from = c(logFC, CI.L, CI.R, adj_p_value)
  ) %>%
  dplyr::filter(!is.na(logFC_Male), !is.na(logFC_Female)) %>%
  dplyr::mutate(
    Ome = dplyr::case_when(
      grepl("metab", assay)              ~ "Metabolomics",
      assay == "prot-ph"                 ~ "Phosphoproteomics",
      assay %in% c("prot-pr", "prot-ol") ~ "Proteomics",
      grepl("rna",  assay)               ~ "Transcriptomics",
      TRUE ~ assay
    ),
    Ome = factor(Ome, levels = c("Transcriptomics", "Proteomics",
                                 "Phosphoproteomics", "Metabolomics")),
    Group = dplyr::recode(randomGroupCode,
                          "ADUEndur" = "EE", "ADUResist" = "RE", "ADUControl" = "CON"),
    Group = factor(Group, levels = c("EE", "RE", "CON"))
  )

scatter_cor = scatter_wide %>%
  dplyr::filter(!is.na(pre_cawg_adj_p) & pre_cawg_adj_p < 0.05, Group != "CON") %>%
  dplyr::group_by(Ome, tissue, Group, Timepoint) %>%
  dplyr::summarise(
    r = round(cor(logFC_Female, logFC_Male,
                  method = "pearson", use = "complete.obs"), 3),
    n_sig = dplyr::n(),
    .groups = "drop"
  )
print(scatter_cor)
```

```{r scatter-plots, fig.width = 10, fig.height = 12}
# One plot per tissue × ome, faceted by Timepoint × Group (EE/RE only).
scatter_sig = scatter_wide %>%
  dplyr::filter(!is.na(pre_cawg_adj_p) & pre_cawg_adj_p < 0.05, Group != "CON")

tissue_ome_combos = scatter_sig %>%
  dplyr::distinct(tissue, Ome) %>%
  dplyr::arrange(tissue, Ome)

for (i in seq_len(nrow(tissue_ome_combos))) {
  tiss     = tissue_ome_combos$tissue[i]
  ome_type = tissue_ome_combos$Ome[i]
  p = .make_sex_scatter(tiss, ome_type, scatter_sig, scatter_cor)
  if (!is.null(p)) print(p)
}
```

---

```{r, include = FALSE}
sex_direct_da_annotated = read_sex_diff %>%
  dplyr::filter((grepl("Male", contrast) & grepl("Female", contrast))) %>%
  dplyr::mutate(contrast_short = gsub("sex_group_timepoint|[()|]", "", contrast),
                contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
                randomGroupCode = stringr::str_split_fixed(contrast_left, "\\.", 3)[,2],
                Timepoint = sub("^[^.]+\\.[^.]+\\.", "", contrast_left)) %>%
  dplyr::select(-contrast_left) %>%
  dplyr::mutate(Timepoint = factor(Timepoint,
                                   levels = c("pre_exercise",
                                              "during_20_min",
                                              "during_40_min",
                                              "post_10_min",
                                              "post_15_30_45_min",
                                              "post_3.5_4_hr",
                                              "post_24_hr"))) %>%
  dplyr::rename(platform = assay) %>%
  dplyr::mutate(assay = ifelse(grepl("metab", platform), "metab", platform)) %>%
  dplyr::select(tissue, assay, platform, feature_id, logFC, CI.L, CI.R, adj_p_value, randomGroupCode, Timepoint)
#precawg_analysis is the direct load_da call above. 
precawg_and_m_vs_f_diff = precawg_analysis %>%
  dplyr::right_join(sex_direct_da_annotated,
                    by = c("tissue", "assay", "feature_id", "randomGroupCode", "Timepoint"),
                    suffix = c("_precawg", "_sex_diff")) 
```


```{r, fig.width=10, fig.height=4}
plot_precawg_mvsf_matrix(precawg_and_m_vs_f_diff)
```

There is a distinct difference in the Male and Female blood RE, especially at 3.5 - So what's going on there? The correlation above for that timepoint looks like it's very high, so why are there so many significant M vs F features at that timepoint? 

```{r blood-re-p35-scatter, fig.width = 6, fig.height = 5}
highlight_ids = precawg_and_m_vs_f_diff %>%
  dplyr::filter(tissue == "blood",
                randomGroupCode == "ADUResist",
                Timepoint == "post_3.5_4_hr",
                adj_p_value_sex_diff < 0.05) %>%
  dplyr::pull(feature_id)

blood_re_p35 = scatter_wide %>%
  dplyr::filter(tissue == "blood",
                Group == "RE",
                Timepoint == "post_3.5_4_hr",
                pre_cawg_adj_p < 0.05)

blood_re_p35_cor = scatter_cor %>%
  dplyr::filter(tissue == "blood", Group == "RE", Timepoint == "post_3.5_4_hr")

p = .make_sex_scatter("blood", "Transcriptomics", blood_re_p35, blood_re_p35_cor,
                      highlight_ids = highlight_ids)
if (!is.null(p)) print(p)
```

```{r}
annotated_differences = precawg_and_m_vs_f_diff %>% 
  left_join(HUMAN_FEATURE_TO_GENE %>% filter(assay == "transcript-rna-seq"), by = c("feature_id", "assay")) 

sig_differences = annotated_differences %>%
  dplyr::filter(Timepoint == "post_3.5_4_hr",
                randomGroupCode == "ADUResist",
                assay == "transcript-rna-seq",
                tissue == "blood",
                adj_p_value_precawg < 0.05 & adj_p_value_sex_diff < 0.05)
write.csv(sig_differences, "outputs/sig_differences_blood_RE_P35_rna.csv", row.names = FALSE)

all_blood_transcripts = unique(annotated_differences %>% dplyr::filter(assay == "transcript-rna-seq", tissue == "blood") %>% pull(gene_symbol)) %>% as.character()

ora_results = run_ORA(
  input = sig_differences$gene_symbol %>% as.character(),
  background = all_blood_transcripts
)
write.csv(ora_results, "outputs/ora_blood_RE_P35_rna.csv", row.names = FALSE)
ora_results %>% dplyr::filter(collection != "CELLMARKER") %>% dplyr::select(set_short, adj_p_value) %>% head(20)
```



# Transcripts included in PreCAWG Analysis

Essentially trying to answer - are we ignoring sexual dimorphism and actually only showing off results that correspond to the female phenotype

```{r, results = "asis"}
manuscript_transcripts = c("NR4A1", "PPARGC1A", "MYC", "CCN1", "FOXO1", "FOXO3")
relevant_feature_ids = HUMAN_FEATURE_TO_GENE %>%
  dplyr::filter(assay == "transcript-rna-seq",
                gene_symbol %in% manuscript_transcripts) %>%
  dplyr::pull(feature_id) %>% as.character()

for (feat in relevant_feature_ids) {
  tryCatch({
    cat("\n\n###", feat, "\n\n")
    print(sex_differences_single_feature(feat))
    print(sex_differences_logfc_feature(feat))
  }, error = function(e) {
    message("Skipping ", feat, ": ", conditionMessage(e))
  })
}
```

For PGC1a specifically we highlight just the muscle, its not really mentioned as much in the context of other tissues in the landscape. 
```{r, results = "asis"}
print(sex_differences_single_feature("PPARGC1A", selected_tissues = "muscle", output_file = "outputs/PPARGC1A.png"))
```


# Metabs included in PreCAWG analysis

1. Carnitines, Glutamic acid, Citrate, Leucine + isoleucine, lactate

```{r, results = "asis"}
manuscript_metabs = c("Lactic acid", "Leucine", "Isoleucine", "Glutamic acid")

for (feat in manuscript_metabs) {
  tryCatch({
    cat("\n\n###", feat, "\n\n")
    print(sex_differences_single_feature(feat))
    print(sex_differences_logfc_feature(feat))
  }, error = function(e) {
    message("Skipping ", feat, ": ", conditionMessage(e))
  })
}
```


# X/Y Linked Features

Identify features mapped to chromosomes X or Y in the sex DA. 

```{r xy-annotation, message = FALSE, warning = FALSE}
# Get chromosome annotations for all RNA-seq features in the sex DA
rna_features = read_sex_diff %>%
  dplyr::filter(grepl("rna", assay)) %>%
  dplyr::distinct(feature_id)

# Join gene symbols via HUMAN_FEATURE_TO_GENE
rna_gene_map = HUMAN_FEATURE_TO_GENE %>%
  dplyr::filter(assay == "transcript-rna-seq") %>%
  dplyr::select(feature_id, gene_symbol) %>%
  dplyr::distinct()

rna_features = rna_features %>%
  dplyr::left_join(rna_gene_map, by = "feature_id")

ensembl = tryCatch(
  biomaRt::useEnsembl("ensembl", dataset = "hsapiens_gene_ensembl"),
  error = function(e) {
    message("biomaRt unavailable, skipping chromosome annotation: ", e$message)
    NULL
  }
)

chrom_annotation = NULL
if (!is.null(ensembl)) {
  chrom_annotation = biomaRt::getBM(
    attributes = c("hgnc_symbol", "chromosome_name"),
    filters    = "hgnc_symbol",
    values     = rna_features$gene_symbol[!is.na(rna_features$gene_symbol)],
    mart       = ensembl
  ) %>%
    dplyr::filter(chromosome_name %in% c(as.character(1:22), "X", "Y")) %>%
    dplyr::distinct(hgnc_symbol, chromosome_name)
}
```

```{r xy-summary, eval = !is.null(chrom_annotation)}
sex_diff_rna_chrom = precawg_and_m_vs_f_diff %>%
  dplyr::filter(grepl("rna", assay)) %>%
  dplyr::left_join(rna_gene_map, by = "feature_id") %>%
  dplyr::left_join(chrom_annotation,
                   by = c("gene_symbol" = "hgnc_symbol")) %>%
  dplyr::mutate(
    chrom_class = dplyr::case_when(
      chromosome_name == "X"  ~ "chrX",
      chromosome_name == "Y"  ~ "chrY",
      !is.na(chromosome_name) ~ "Autosomal",
      TRUE                    ~ "Unknown"
    )
  )

# Proportion sig by chromosome class
chrom_summary = sex_diff_rna_chrom %>%
  dplyr::filter(chrom_class != "Unknown") %>%
  dplyr::group_by(tissue, Timepoint, randomGroupCode, chrom_class) %>%
  dplyr::summarise(
    n_total    = dplyr::n(),
    n_sig      = sum(adj_p_value_sex_diff < 0.05, na.rm = TRUE),
    pct_sig    = round(100 * n_sig / n_total, 1),
    .groups = "drop"
  )
print(chrom_summary)

```


