---
title: "baseline_sex_differences"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(biomaRt)
library(stringr)
library(ggplot2)
# 
# config = jsonlite::fromJSON("~/config.json")
# repo_local_dir = config$precovid_repo_path
# devtools::load_all(config$data_package_path) 
```


So this document/report aims to deliver and summarize the sex-baseline comparisons.
For rationale and sensitivity analysis, please review \code{sex_differences_sens_analysis.Rmd}
for explanation as to how the models were specifically chosen. 

Starting parameters for every tissue
```{r}
desired_tissues = c("adipose")



```

Starting with transcript-rna-seq
```{r, eval = F}
desired_ome = 'transcript-rna-seq'
for(tissue in desired_tissues){
  parsed_qc_norm = load_qc(selected_omes = desired_ome,
                           selected_tissues = tissue,
                           load_acute_only = FALSE)
  #here we're using the qc-norm for transcriptomics even though we'd normally
  #want to do something to model the counts to first regress out some
  #technical covs
  metadata = parsed_qc_norm[[tissue]][[desired_ome]][['sample_metadata']] %>%
    dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
  rownames(metadata) = metadata$vialLabel
  
  qc_norm_matrix = parsed_qc_norm[[tissue]][[desired_ome]][["qc_norm"]] %>%
    dplyr::select(rownames(metadata))
  process_metadata = process_covariates(meta = metadata,
                                        selected_ome = desired_ome,
                                        tissue_input = tissue)
  process_metadata$metadata$Sex <- relevel(process_metadata$metadata$Sex, ref = "Female")
  full_formula = "~ Sex" 
  #so because we're using the qc-norm logcpm, we dont' use voom 
  fit = variancePartition::dream(qc_norm_matrix, full_formula, process_metadata$metadata)
  fit = variancePartition::eBayes(fit)
  .convert_sex_diff_output(fit, 
                           metadata = process_metadata$metadata,
                           formula = full_formula,
                           tissue = tissue,
                           ome = desired_ome)
  #so here there's actually no need for a random effect--there's no repeated samples bc its all just baseline. 
  #I'm only using the dream syntax & package (which defaults to just a fixed effect model) because the rest of the formatting is consistent
  
}
```

Prot-pr/h
```{r, eval = F}
prot_omes = c('prot-pr', 'prot-ph')
for(desired_ome in prot_omes){
  for(tissue in desired_tissues){
    if(tissue == "blood") next #prot-pr has no blood tissue
    ome_data = load_qc(selected_tissues = tissue,
                       selected_omes = desired_ome,
                       load_acute_only = FALSE)
    
    metadata = ome_data[[tissue]][[desired_ome]][['sample_metadata']] %>% 
      dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
    rownames(metadata) = metadata$vialLabel
    data_matrix = ome_data[[tissue]][[desired_ome]][['qc_norm']] %>%
      dplyr::select(as.character(metadata$vialLabel))
    process_metadata = process_covariates(meta = metadata,
                                          selected_ome = desired_ome,
                                          tissue_input = tissue)
    process_metadata$metadata$Sex <- relevel(process_metadata$metadata$Sex, ref = "Female")
    
    full_formula = "~ Sex" 
    fit = variancePartition::dream(data_matrix, full_formula, process_metadata$metadata)
    fit = variancePartition::eBayes(fit)
    .convert_sex_diff_output(fit, 
                             metadata = process_metadata$metadata,
                             formula = full_formula,
                             tissue = tissue,
                             ome = desired_ome)
    #so here there's actually no need for a random effect--there's no repeated samples bc its all just baseline. 
    #I'm only using the dream syntax & package (which defaults to just a fixed effect model) because the rest of the formatting is consistent
    
  }
}
```

Prot-ol
```{r, eval = F}
desired_ome = "prot-ol"
tissue = "blood" #o-link is only in blood 
ome_data = load_qc(selected_tissues = tissue,
                   selected_omes = desired_ome,
                   load_acute_only = FALSE)

metadata = ome_data[[tissue]][[desired_ome]][['sample_metadata']] %>% 
  dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
rownames(metadata) = metadata$vialLabel
data_matrix = ome_data[[tissue]][[desired_ome]][['qc_norm']] %>%
  dplyr::select(as.character(metadata$vialLabel))
process_metadata = process_covariates(meta = metadata,
                                      selected_ome = desired_ome,
                                      tissue_input = tissue)
process_metadata$metadata$Sex <- relevel(process_metadata$metadata$Sex, ref = "Female")

full_formula = "~ Sex" 
fit = variancePartition::dream(data_matrix, full_formula, process_metadata$metadata)
fit = variancePartition::eBayes(fit)
.convert_sex_diff_output(fit, 
                         metadata = process_metadata$metadata,
                         formula = full_formula,
                         tissue = tissue,
                         ome = desired_ome)
```
Metab
```{r, eval = F}
for(metab_platform in metab_only_list()){
  desired_ome = metab_platform
  for(tissue in desired_tissues){
    ome_data = load_qc(selected_tissues = tissue,
                       selected_omes = desired_ome,
                       load_acute_only = FALSE,
                       remove_unnamed_metab = TRUE,
                       remove_redundant_metab = TRUE)
    if(length(ome_data) == 0) next #not all platforms were run in all tissues
    metadata = ome_data[[tissue]][[desired_ome]][['sample_metadata']] %>% 
      dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
    rownames(metadata) = metadata$vialLabel
    data_matrix = ome_data[[tissue]][[desired_ome]][['qc_norm']] %>%
      dplyr::select(as.character(metadata$vialLabel))
    if(nrow(data_matrix) == 0) next #not all platforms were run in all tissues

    process_metadata = process_covariates(meta = metadata,
                                          selected_ome = desired_ome,
                                          tissue_input = tissue)
    process_metadata$metadata$Sex <- relevel(process_metadata$metadata$Sex, ref = "Female")
    
    full_formula = "~ Sex"
    
    fit = variancePartition::dream(data_matrix, full_formula, process_metadata$metadata)
    fit = variancePartition::eBayes(fit)
    .convert_sex_diff_output(fit, 
                             metadata = process_metadata$metadata,
                             formula = full_formula,
                             tissue = tissue,
                             ome = desired_ome)
    #so here there's actually no need for a random effect--there's no repeated samples bc its all just baseline. 
    #I'm only using the dream syntax & package (which defaults to just a fixed effect model) because the rest of the formatting is consistent
    
  }
}
```

Atac
```{r, eval = F}
desired_ome = 'epigen-atac-seq'
for(tissue in desired_tissues){
  if(tissue == "adipose") next
  counts_data_path = .find_path_name(desired_ome = desired_ome, tissue = tissue, data_type = "epigen-atac-seq_counts")
  raw_counts_input = MotrpacBicQC::dl_read_gcp(counts_data_path, sep = '\t', tmpdir = file.path(repo_local_dir, "data/tmp/")) %>%
    dplyr::mutate(rownames = stringr::str_c(chrom,":",start,"-",end,sep = "")) %>%
    tibble::column_to_rownames("rownames") %>%
    dplyr::filter(!(chrom %in% c("chrX","chrY"))) %>%
    dplyr::select(-chrom,-start,-end) %>%
    as.data.frame()
  parsed_qc_norm = load_qc(repo_local_dir,
                           selected_omes = desired_ome,
                           selected_tissues = tissue,
                           load_acute_only = FALSE,
                           epigen = TRUE)
  metadata = parsed_qc_norm[[tissue]][[desired_ome]][['sample_metadata']] %>%
    dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
  rownames(metadata) = metadata$vialLabel
  
  raw_counts_input = raw_counts_input %>%
    dplyr::filter(rownames(.) %in% rownames(parsed_qc_norm[[tissue]][[desired_ome]][['qc_norm']])) %>% #filter to select genes
    dplyr::select(as.character(metadata$vialLabel)) #reorganize and filter to only desired participants
  dim(raw_counts_input) 
  
  pre_dge <- edgeR::DGEList(counts = raw_counts_input) #standard dge processing
  pre_dge <- edgeR::calcNormFactors(pre_dge) #standard dge processing
  process_metadata = process_covariates(meta = metadata,
                                        selected_ome = desired_ome,
                                        tissue_input = tissue)
  process_metadata$metadata$Sex <- relevel(process_metadata$metadata$Sex, ref = "Female")
  full_formula = "~ Sex"
  expression_object = variancePartition::voomWithDreamWeights(pre_dge, full_formula, process_metadata$metadata) 
  fit = variancePartition::dream(expression_object, full_formula, process_metadata$metadata)
  fit = variancePartition::eBayes(fit)
  .convert_sex_diff_output(fit, 
                           metadata = process_metadata$metadata,
                           formula = full_formula,
                           tissue = tissue,
                           ome = desired_ome)
  #so here there's actually no need for a random effect--there's no repeated samples bc its all just baseline. 
  #I'm only using the dream syntax & package (which defaults to just a fixed effect model) because the rest of the formatting is consistent
  
}
```


Reading in all the data from the baseline results & running enrichment.
CAMERA for all non metab omes

```{r}
desired_specific_tissue = "adipose"
#i basically uploaded everything from output_dir into this bucket.
# results_path = "gs://pre-cawg/analysis/sex-differences-baseline"
#if you want to work with exactly the same dataset
output_dir = file.path(repo_local_dir, "data/tmp/sex_differences_baseline/")
sex_differences_analysis = list()
for(file in list.files(output_dir, full.names = TRUE)){
  if(!grepl("txt", file)) next
  tissue = .find_tissue(file)
  if(!tissue == desired_specific_tissue) next
  ome = .find_ome(file)
  if(is.null(ome)) next
  tissue_ome = interaction(tissue, ome) %>% as.character()
  # message(tissue_ome)
  single_file_read = read.csv(file, sep = "\t") %>%
    mutate(z.std = as.numeric(sign(t) * qnorm(p_value / 2, lower.tail = FALSE))) 
  #calculate z.std the same way as the variancePartition
  if(ome == "prot-ph")
    single_file_read = single_file_read %>% 
    select(-flanking_sequence)
  if(!grepl("metab", ome))
    single_file_read = single_file_read %>% 
    select(-gene_symbol)
  
  single_file_read = single_file_read %>%
    select(-any_of(c("entrez_gene", "ensembl_gene", "custom_annotation", "flanking_sequence", "refmet_name", "kegg_id")))
  
  
  sex_differences_analysis[[tissue]][[ome]] = single_file_read
}
# sex_differences_analysis$`adipose.prot-ph` = NULL
sex_differences_enrichment = run_cameraPR(DA_list = sex_differences_analysis)

write_with_path_name(sex_differences_enrichment,
                     local_path = output_dir,
                     ome = "transcript-rna-seq", #temp. will manually change to "all"
                     tissue = tissue,
                     data_category = "enrichment",
                     data_details = "sex_diff_baseline_simple",
                     version = "1.31")

```

Lets actually describe the DA results then describe the enrichment results.
```{r}
sex_diff_unlist = unlist(sex_differences_analysis, recursive = FALSE)
single_da_matrix = rbindlist(sex_diff_unlist, fill = TRUE) 
single_da_matrix = single_da_matrix %>%
  mutate(assay = case_when(
    grepl("metab", assay) ~ "metab",
    TRUE ~ assay
  ))

single_da_matrix %>%
  ggplot(aes(x = p_value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ assay, scales = "free_y") +
  theme_minimal()
```
```{r}
single_da_matrix %>%
  ggplot(aes(x = logFC, y = -log10(p_value))) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ assay, scales = "free") +
  theme_minimal() +
  labs(x = "log2 Fold Change", y = "-log10(p-value)")
```
```{r}
signif_tables = single_da_matrix %>%
  filter(adj_p_value < 0.05) %>%
  group_split(assay)

names(signif_tables) = single_da_matrix %>%
  filter(adj_p_value < 0.05) %>%
  group_keys(assay) %>%
  pull(assay)

percent_signif = single_da_matrix %>%
  group_by(assay) %>%
  summarise(
    total_features = n(),
    significant = sum(adj_p_value < 0.05),
    percent_significant = 100 * significant / total_features
  )
percent_signif
```

```{r, eval = F}
sex_differences_enrichment_input = sex_differences_enrichment %>%
  mutate(contrast = assay)

sex_differences_enrichment = run_cameraPR(DA_list = sex_differences_analysis) 


```

Plots
```{r}
sex_differences_enrichment_input = sex_differences_enrichment %>%
  mutate(contrast = assay)

#basically using `pathway_heatmap` from Tyler
args <- list(x = sex_differences_enrichment_input,
             n_top = 25, 
             set_column = "set_short",
             statistic_column = "z.std",
             contrast_column = "assay",
             padj_column = "adj_p_value",
             padj_legend_title = "test",
             padj_cutoff = 0.05,
             plot_sig_only = TRUE,
             padj_fill = "grey",
             colors = c("#3366ff", "darkred"),            
             scale_by = "row",
             cell_size = unit(14, "points"),
             height = 10,
             width = 15,
             units = "in",
             heatmap_args = list(cluster_rows = FALSE,
                                 cluster_columns = FALSE),
             padj_args = list(),
             save_args = list(),
             draw_args = list())
args[["filename"]] <- file.path(output_dir, "figures", paste0(desired_specific_tissue, "_baseline_enrichment_camera.pdf"))
do.call(what = enrichmap, args = args)
```
