---
title: "c-means_main_figure"
author: "Dan Katz"
date: "2025-11-13"
output: html_document
---


```{r setup, message=FALSE, warning=FALSE}
# --- Configuration Setup ---
# This script uses a JSON configuration file to manage local file paths and machine-specific settings.
#
# To use the MoTrPAC config file:
# 1. Open the template 'example_config.json' in the 'precovid-analyses/' root folder.
# 2. Save a copy to a persistent location (e.g., your home directory: '~/motrpac_config.json').
# 3. Customize the file paths for your machine and remove any comments from the JSON file.
# 4. This script requires the following fields in the config file:
#    a. "gsutil_user" and "gsutil_path": Specifies the path to the gsutil executable.
#    b. "local_path": The absolute path to the 'precovid-analyses' repository, without a trailing slash.
# 5. Provide the path to your config file below.

# config <- jsonlite::read_json(path = "//mnt/c/Users/dankatz/Documents/motrpac_config.json")
# config <- rjson::fromJSON(file = '//mnt/c/Users/dankatz/Documents/motrpac_config.json')
# config = jsonlite::fromJSON("~/config.json")
# source(file.path(config$precovid_repo_path,"/library/library-general.R"))

# --- Package Loading ---
# Load all required packages for the session.
session_packages <- c('pheatmap', 'limma', 'UpSetR', 'ComplexHeatmap', 'gprofiler2',
                      'plier', 'cluster', 'readxl','qvalue', 'reshape2', 'ggh4x','ggVennDiagram','ggpattern',
                      'MASS', 'splitstackshape', 'MotrpacBicQC', 'tidyverse', 'here',
                      'powerjoin','MotrpacHumanPreSuspensionAnalysis','variancePartition','patchwork','ggbeeswarm','circlize', 'forcats',
                      'grid','gridExtra')

for(i in session_packages) {
  tryCatch(
    {
      library(i, character.only = TRUE)
    }, error = function(e) {
      print(paste("Cannot load package:", i, ", please install."))
    }
  )
}

# --- Namespace Conflict Preferences ---
# conflict_prefer('select','dplyr')
# conflict_prefer('filter','dplyr')

sessionInfo()
```

# C-means Trajectory Visualizations

## Combined C-means Trajectory Line Plots
This plot generates trajectories for all cluster centroids across all tissues. While not used in the final paper, its code provides a basis for subsequent analyses.

```{r All c-means lines} 
# --- User Configuration ---
# Define which clusters to highlight (alpha = 1) for each tissue.
# Non-selected clusters will be dimmed (alpha = 0.3).
selected_clusters <- list(
  muscle = c("7","4"),    # Highlight clusters 7 and 4 in 'muscle'
  adipose = c("7", "8"),  # Highlight clusters 7 and 8 in 'adipose'
  blood = c("5","6","4")  # Highlight clusters 5, 6, and 4 in 'blood'
)

# --- Abbreviation Map ---
# Maps long contrast names from the data to abbreviated labels for the plot.
contrast_recode_map <- c(
  "Pre" = "pre_exercise",
  "P10M" = "post_10_min",
  "P15-45M" = "post_15_30_45_min",
  "P3.5/4H" = "post_3.5_4_hr",
  "P24H" = "post_24_hr"
)
# Define the chronological order for the abbreviated contrast levels.
ordered_contrast_levels <- names(contrast_recode_map)

# --- Data Preparation ---
tissues <- c("muscle", "adipose", "blood")
all_tissue_dfs <- list()

for (t in tissues) {
  # (Assuming FCM_CLUSTERS is available in the environment)
  centers <- FCM_CLUSTERS[[t]]$centers
  
  # 1. Standardize column names
  colnames(centers) <- sub("group_timepointADU([^ ]+) .*",
                           "\\1",
                           colnames(centers))
  
  # 2. Include pre-exercise timepoints (all zeros)
  pre_ex <- matrix(data = 0,
                   nrow = nrow(centers),
                   ncol = 2L)
  colnames(pre_ex) <- c("Endur.pre_exercise", "Resist.pre_exercise")
  centers <- cbind(pre_ex, centers)
  
  # 3. Reorder columns to group by modality (Endurance, Resistance)
  col_order <- c(grep("Endur", colnames(centers)),
                 grep("Resist", colnames(centers)))
  centers <- centers[, col_order]
  
  # 4. Pivot and clean data
  df_temp <- centers %>%
    as.data.frame() %>%
    tibble::rownames_to_column("cluster") %>%
    tidyr::pivot_longer(
      cols = -cluster,
      names_to = "contrast",
      values_to = "center"
    ) %>%
    dplyr::mutate(modality = ifelse(grepl("Endur", contrast),
                                    "EE",
                                    "RE"),
                  contrast = sub("[^.]+\\.", "", contrast),
                  tissue = t) # Add tissue column
  
  all_tissue_dfs[[t]] <- df_temp
}

# Combine data from all tissues
center_df <- dplyr::bind_rows(all_tissue_dfs) %>%
  # Apply recoding to the contrast column using the abbreviation map
  dplyr::mutate(contrast = forcats::fct_recode(contrast, !!!contrast_recode_map)) %>%
  # Re-factor all categorical columns based on the combined data
  dplyr::mutate(
    # Set explicit chronological order for contrast
    contrast = factor(contrast, levels = ordered_contrast_levels),
    # Re-factor other categorical columns
    across(.cols = c(cluster, modality, tissue),
           .fns = ~ factor(.x, levels = unique(.x)))
  ) %>%
  # Determine if a cluster is selected for highlighting (tissue-specific)
  dplyr::rowwise() %>%
  dplyr::mutate(
    is_selected = cluster %in% selected_clusters[[tissue]]
  ) %>%
  dplyr::ungroup()

# --- Color Logic: Determine Color by Tissue-Specific Max Absolute Value ---
# Group by tissue and cluster to calculate color independently for each trajectory.
center_df <- center_df %>%
  dplyr::group_by(tissue, cluster) %>%
  dplyr::mutate(
    # Find the index of the row with the max absolute center value within the group
    max_abs_idx = which.max(abs(center)),
    # Get the contrast name at that index
    max_abs_contrast = contrast[max_abs_idx],
    # Get the sign of the center value at that index
    max_abs_sign = ifelse(center[max_abs_idx] >= 0, "Positive", "Negative"),
    # Combine contrast and sign into a final color group
    max_abs_contrast_group = paste(max_abs_contrast, max_abs_sign, sep = "_")
  ) %>%
  dplyr::ungroup() %>%
  # Ensure the new color variable is a factor
  dplyr::mutate(
    max_abs_contrast_group = factor(max_abs_contrast_group, 
                                    levels = unique(max_abs_contrast_group))
  ) %>%
  # Arrange by 'is_selected' so highlighted lines (TRUE) are plotted on top
  dplyr::arrange(is_selected)

# --- Plotting ---
p <- ggplot2::ggplot(center_df,
                     aes(x = contrast,
                         y = center,
                         # Group by interaction to ensure independent lines across facets
                         group = interaction(tissue, cluster),
                         # Color by the max absolute contrast group
                         color = max_abs_contrast_group,
                         # Map alpha to the selection status for highlighting
                         alpha = is_selected)) + 
  
  # Lines are solid and thin
  ggplot2::geom_line(linewidth = 0.8) + 
  
  # Use shape 21 (filled circle) and map 'fill' to the color group
  ggplot2::geom_point(
    aes(fill = max_abs_contrast_group), # Map fill to the color group
    color = "black", # Black stroke for contrast
    shape = 21,      # Shape 21 uses both 'color' (stroke) and 'fill'
    size = 5,        # Increased point size
    show.legend = FALSE
  ) + 
  
  # Add cluster number text inside the points
  ggplot2::geom_text(
    aes(label = cluster), 
    # Filter out 'Pre' points for labeling
    data = dplyr::filter(center_df, contrast != "Pre"), 
    color = "black", # Text color for high contrast
    size = 3, 
    show.legend = FALSE
  ) +
  
  # Facet by Modality (rows) and Tissue (columns) with free x-scales
  ggplot2::facet_grid(modality ~ tissue, scales = "free_x", space = "free_x") +
  
  ggplot2::scale_x_discrete(
    expand = ggplot2::expansion(add = 0.15) 
  ) +
  
  # Manual alpha scale for highlighting (TRUE=1, FALSE=0.2)
  ggplot2::scale_alpha_manual(values = c("FALSE" = 0.2, "TRUE" = 1), guide = "none") +
  
  # Set color for lines (suppress legend)
  ggplot2::scale_color_brewer(palette = "Set2", guide = "none") +
  # Set fill for points (suppress legend)
  ggplot2::scale_fill_brewer(palette = "Set2", guide = "none") +
  
  ggplot2::labs(x = NULL,
                y = "Cluster Centroid",
                color = "Max Absolute Center Group") + 
  
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90,
                                                     hjust = 1,
                                                     vjust = 0.5),
                 panel.grid.major.x = ggplot2::element_blank(),
                 legend.position = "none")

# --- Export to PDF ---
# pdf("cluster_centroid_lineplot.pdf", width = 6.5, height = 6.75)
print(p)
# dev.off()

# Display the plot in the R environment
# print(p)
```

## Sparkline Plots
This section generates a series of small, individual "sparkline" plots for each cluster group.
```{r Sparklines Figure 4C}

# ==============================================================================
# 1. SETUP & CONFIGURATION
# ==============================================================================

library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(purrr)
library(RColorBrewer)
library(patchwork)

# Contrast level mapping
contrast_recode_map <- c(
  "Pre"     = "pre_exercise",
  "P10M"    = "post_10_min",
  "P15-45M" = "post_15_30_45_min",
  "P3.5/4H" = "post_3.5_4_hr",
  "P24H"    = "post_24_hr"
)
ordered_contrast_levels <- names(contrast_recode_map)


# ==============================================================================
# 2. DATA PREPARATION & CENTROID PROCESSING
# ==============================================================================

tissues <- c("muscle", "adipose", "blood")
all_tissue_dfs <- list()

for (t in tissues) {
  centers <- FCM_CLUSTERS[[t]]$centers
  colnames(centers) <- sub("group_timepointADU([^ ]+) .*", "\\1", colnames(centers))
  
  pre_ex <- matrix(data = 0, nrow = nrow(centers), ncol = 2L)
  colnames(pre_ex) <- c("Endur.pre_exercise", "Resist.pre_exercise")
  centers <- cbind(pre_ex, centers)
  
  col_order <- c(grep("Endur", colnames(centers)), grep("Resist", colnames(centers)))
  centers <- centers[, col_order]
  
  df_temp <- centers %>%
    as.data.frame() %>%
    tibble::rownames_to_column("cluster") %>%
    tidyr::pivot_longer(cols = -cluster, names_to = "contrast", values_to = "center") %>%
    dplyr::mutate(
      modality = ifelse(grepl("Endur", contrast), "EE", "RE"),
      contrast = sub("[^.]+\\.", "", contrast),
      tissue   = t
    )
  
  all_tissue_dfs[[t]] <- df_temp
}

# Merge and calculate max magnitude timepoints
center_df <- dplyr::bind_rows(all_tissue_dfs) %>%
  dplyr::mutate(contrast = forcats::fct_recode(contrast, !!!contrast_recode_map)) %>%
  dplyr::mutate(
    contrast = factor(contrast, levels = ordered_contrast_levels),
    across(.cols = c(cluster, modality, tissue), .fns = ~ factor(.x, levels = unique(.x)))
  ) %>%
  dplyr::rowwise() %>%
  dplyr::ungroup()

center_df <- center_df %>%
  dplyr::group_by(tissue, cluster) %>%
  dplyr::mutate(
    max_abs_idx      = which.max(abs(center)),
    max_abs_contrast = contrast[max_abs_idx],
    max_abs_sign     = ifelse(center[max_abs_idx] >= 0, "Positive", "Negative")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    max_abs_contrast = as.character(max_abs_contrast),
    max_abs_contrast = dplyr::case_when(
      tissue == "blood" & max_abs_contrast == "P10M" ~ "P15-45M", #trajectories with max deviation at P10 or P30 are grouped in blood
      TRUE ~ max_abs_contrast
    ),
    max_abs_contrast       = factor(max_abs_contrast, levels = unique(max_abs_contrast)),
    max_abs_contrast_group = paste(max_abs_contrast, max_abs_sign, sep = "_"),
    max_abs_contrast_group = factor(max_abs_contrast_group, levels = unique(max_abs_contrast_group))
  )


# ==============================================================================
# 3. COLOR PALETTE GENERATION
# ==============================================================================

user_palette_map <- c(
  "P3.5/4H_Negative" = "Reds",
  "P3.5/4H_Positive" = "Greens",
  "P15-45M_Positive" = "Blues",
  "P24H_Positive"    = "Purples",
  "P15-45M_Negative" = "Oranges",
  "P24H_Negative"    = "YlOrBr"
)

all_groups_in_data <- levels(center_df$max_abs_contrast_group)
unmapped_groups <- setdiff(all_groups_in_data, names(user_palette_map))

if (length(unmapped_groups) > 0) {
  warning(paste("Found unmapped groups, assigning 'Greys' palette:", paste(unmapped_groups, collapse = ", ")))
  default_map <- rep("Greys", length(unmapped_groups))
  names(default_map) <- unmapped_groups
} else {
  default_map <- c()
}

group_palette_map <- c(user_palette_map, default_map)

color_lookup_df <- center_df %>%
  distinct(tissue, cluster, max_abs_contrast_group) %>%
  group_by(tissue, max_abs_contrast_group) %>%
  tidyr::nest(items_to_color = c(cluster)) %>%
  mutate(palette_name = group_palette_map[as.character(max_abs_contrast_group)]) %>%
  rowwise() %>%
  mutate(
    n_clusters = nrow(items_to_color),
    cluster_shades = list(
      if (n_clusters > 9) {
        colorRampPalette(rev(RColorBrewer::brewer.pal(9, palette_name)))(n_clusters)
      } else {
        rev(RColorBrewer::brewer.pal(max(3, n_clusters), palette_name))[1:n_clusters]
      }
    )
  ) %>%
  ungroup() %>%
  mutate(
    items_with_color = map2(
      items_to_color, 
      cluster_shades, 
      ~ .x %>% arrange(cluster) %>% bind_cols(cluster_color = .y)
    )
  ) %>%
  select(tissue, max_abs_contrast_group, items_with_color) %>%
  tidyr::unnest(cols = items_with_color) %>%
  select(tissue, cluster, max_abs_contrast_group, cluster_color)

center_df <- center_df %>%
  select(-any_of("cluster_color")) %>%
  left_join(color_lookup_df, by = c("tissue", "cluster", "max_abs_contrast_group"))


# ==============================================================================
# 4. PLOTTING - SPARKLINE PANELS
# ==============================================================================

# Calculate global Y-limits across entire dataset to enforce identical y-scales
global_y_min <- min(center_df$center, na.rm = TRUE)
global_y_max <- max(center_df$center + 0.3, na.rm = TRUE) # Includes label height offset
global_y_limits <- c(global_y_min, global_y_max)

tissue_order   <- c("adipose", "blood", "muscle")
contrast_order <- c("P15-45M", "P3.5/4H", "P24H")

plot_groups <- center_df %>%
  distinct(tissue, max_abs_contrast) %>%
  mutate(
    tissue           = factor(tissue, levels = tissue_order),
    max_abs_contrast = factor(max_abs_contrast, levels = contrast_order)
  ) %>%
  filter(!is.na(tissue), !is.na(max_abs_contrast)) %>% 
  arrange(tissue, max_abs_contrast) %>%
  mutate(plot_index = 1:n())

# Generate list of individual sparkline plots
plot_list <- pmap(
  .l = list(t = plot_groups$tissue, c = plot_groups$max_abs_contrast, i = plot_groups$plot_index),
  .f = function(t, c, i) {
    
    plot_data <- center_df %>% filter(tissue == t, max_abs_contrast == c)
    all_clusters_in_panel <- unique(as.character(plot_data$cluster))
    
    # 4a. Smart Label Spacing Algorithm (Greedy + Fallback)
    ranking_data <- plot_data %>%
      group_by(cluster, contrast) %>%
      summarise(peak_score = max(abs(center), na.rm = TRUE), .groups = "drop") %>%
      mutate(cluster = as.character(cluster), contrast = as.character(contrast)) %>%
      arrange(desc(peak_score))
    
    final_assignments <- tibble(cluster = character(), contrast = character())
    remaining_ranking  <- ranking_data
    
    # Assign available unique timepoints
    while (nrow(remaining_ranking) > 0) {
      winner <- remaining_ranking %>% slice(1)
      
      final_assignments <- bind_rows(
        final_assignments, 
        tibble(cluster = winner$cluster, contrast = winner$contrast)
      )
      
      remaining_ranking <- remaining_ranking %>%
        filter(cluster != winner$cluster, contrast != winner$contrast)
    }
    
    # Fallback for overflow clusters: assign to peak timepoint
    unassigned <- setdiff(all_clusters_in_panel, final_assignments$cluster)
    if (length(unassigned) > 0) {
      fallback <- ranking_data %>%
        filter(cluster %in% unassigned) %>%
        group_by(cluster) %>%
        slice_max(peak_score, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        select(cluster, contrast)
      
      final_assignments <- bind_rows(final_assignments, fallback)
    }
    
    # --- HARDCODED LABEL OVERRIDES ---
    # Move Blood Cluster 7 label to match Cluster 1 (or midpoint "P15-45M")
    # This helps the labels not overlap. the hard coding affects labels, not data
    if (t == "blood" && "7" %in% final_assignments$cluster) {
      c1_contrast <- final_assignments %>% filter(cluster == "1") %>% pull(contrast)
      target_contrast <- if (length(c1_contrast) > 0) c1_contrast[1] else "P15-45M"
      
      final_assignments <- final_assignments %>%
        mutate(contrast = if_else(cluster == "7", target_contrast, contrast))
    }
    
    # Move Adipose Cluster 13 label to the last timepoint ("P24H")
    if (t == "adipose" && "13" %in% final_assignments$cluster) {
      final_assignments <- final_assignments %>%
        mutate(contrast = if_else(cluster == "13", "P24H", contrast))
    }
    
    label_data <- plot_data %>%
      mutate(cluster = as.character(cluster), contrast = as.character(contrast)) %>%
      inner_join(final_assignments, by = c("cluster", "contrast"))

    cluster_color_map <- plot_data %>% distinct(cluster, cluster_color) %>% tibble::deframe()
    
    # 4b. Single Sparkline Construction
    p_single <- ggplot(plot_data, aes(x = contrast, y = center, group = cluster)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.25) +
      
      # Trajectory lines
      geom_line(color = "black", linewidth = 0.85, alpha = 0.7) +
      geom_line(aes(color = cluster), linewidth = 0.75, alpha = 0.9) +
      
      # Circle labels & text
      geom_point(
        data = label_data, 
        aes(fill = cluster, y = center + 0.3), 
        shape = 21, color = "black", size = 1.2, stroke = 0.2
      ) +
      geom_text(
        data = label_data, 
        aes(label = cluster, y = center + 0.3), 
        color = "black", size = 0.8, fontface = "bold"
      ) +
      
      # Color & Uniform Global Y-Scale
      scale_color_manual(values = cluster_color_map, guide = "none") +
      scale_fill_manual(values = cluster_color_map, guide = "none") +
      scale_y_continuous(
        limits = global_y_limits,
        expand = expansion(mult = c(0.06, 0.14))
      ) +
      
      coord_cartesian(clip = "off") +
      labs(x = NULL, y = NULL) + 
      theme_bw(base_size = 8) +
      theme(
        panel.border  = element_rect(color = "black", fill = NA, linewidth = 0.5),
        axis.ticks    = element_blank(), 
        axis.text     = element_blank(), 
        axis.title    = element_blank(),
        panel.grid    = element_blank(), 
        legend.position = "none",
        panel.spacing = unit(0, "lines"),
        plot.margin   = margin(1, 1, 1, 1, "pt") 
      )
    
    # Add modality facet strip to first plot only
    if (i == 1) {
      p_single <- p_single + 
        facet_grid(modality ~ ., switch = "y") +
        theme(
          strip.text.y = element_text(size = 5, angle = 0), 
          strip.background = element_blank()
        )
    } else {
      p_single <- p_single + 
        facet_grid(modality ~ .) +
        theme(strip.text.y = element_blank())
    }
    
    return(p_single)
  }
)


# ==============================================================================
# 5. HEADER STRIPS (TISSUE & CONTRAST LABELS)
# ==============================================================================

# Contrast headers (Middle Row) with "P10-45M" label for blood
contrast_color_lookup <- tibble(
  max_abs_contrast   = names(contrast_recode_map), 
  long_contrast_name = unname(contrast_recode_map)
)

plot_groups_for_labels <- left_join(plot_groups, contrast_color_lookup, by = "max_abs_contrast") %>%
  mutate(
    display_contrast = ifelse(
      tissue == "blood" & max_abs_contrast == "P15-45M", 
      "P10-45M", 
      as.character(max_abs_contrast)
    )
  )

p_contrast_labels <- ggplot(plot_groups_for_labels) +
  geom_rect(
    aes(xmin = plot_index - 0.5, xmax = plot_index + 0.5, ymin = 0, ymax = 1, fill = long_contrast_name), 
    color = NA
  ) +
  geom_text(
    aes(x = plot_index, y = 0.5, label = display_contrast), 
    color = "black", fontface = "bold", size = 1.3
  ) +
  scale_fill_manual(values = HUMAN_ACUTE_TIMEPOINT_COLORS, guide = "none") +
  scale_x_continuous(limits = c(0.5, nrow(plot_groups) + 0.5), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void()

# Tissue headers (Top Row)
tissue_label_data <- plot_groups %>%
  group_by(tissue) %>%
  summarise(
    x_center = mean(plot_index), 
    xmin     = min(plot_index) - 0.5, 
    xmax     = max(plot_index) + 0.5, 
    .groups  = 'drop'
  )

p_tissue_labels <- ggplot(tissue_label_data, aes(label = tissue, fill = tissue)) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1), color = NA) +
  geom_text(aes(x = x_center, y = 0.5), color = "black", fontface = "bold", size = 1.5) +
  scale_fill_manual(values = HUMAN_TISSUE_COLORS, guide = "none") +
  scale_x_continuous(limits = c(0.5, nrow(plot_groups) + 0.5), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void()


# ==============================================================================
# 6. COMBINE & EXPORT
# ==============================================================================

sparklines_plot <- patchwork::wrap_plots(plot_list, nrow = 1)

final_combined_plot <- p_tissue_labels / p_contrast_labels / sparklines_plot +
  plot_layout(heights = c(0.1, 0.08, 1))

plot_width_in  <- 0.5
plot_height_in <- 0.9
num_plots      <- length(plot_list) 
total_width_in <- num_plots * plot_width_in

pdf("cluster_centroid_sparklines_with_labels_new_axes.pdf", width = total_width_in, height = plot_height_in)
print(final_combined_plot)
dev.off()
```
# Top ORA Enrichments of C-means Clusters
This section identifies the top Over-Representation Analysis (ORA) gene sets from the C-means clusters (hard-clustered with probability > 0.3) and finds which sets are common across all three tissues.

```{r ora_enrichment}
# Filter for significant ORA results
FCM_ORA_top_hits <- FCM_ORA %>% 
  filter(adj_p_value < 0.05)

# Extract unique significant sets for each tissue
adipose_ids <- FCM_ORA_top_hits %>%
  filter(tissue == "adipose") %>%
  pull(set) %>%
  unique()

blood_ids <- FCM_ORA_top_hits %>%
  filter(tissue == "blood") %>%
  pull(set) %>%
  unique()

muscle_ids <- FCM_ORA_top_hits %>%
  filter(tissue == "muscle") %>%
  pull(set) %>%
  unique()

# Find the intersection of the three vectors to get common sets
common_sets <- intersect(intersect(adipose_ids, blood_ids), muscle_ids)

# Print the list of common sets
cat("The following sets appear in all three tissues (adipose, blood, and muscle):\n")
if (length(common_sets) > 0) {
  cat(common_sets, sep = "\n")
} else {
  cat("No common sets were found.\n")
}
```

# Bar Graphs for Hard-Clustered FCM Results
This section performs a "hard clustering" assignment on the fuzzy c-means results. It assigns each feature to the cluster with the highest membership score, filters assignments below a confidence threshold, and creates a faceted, stacked bar plot showing the omics composition of each cluster by tissue.

```{r hard_clustering_prep}
# --- Hard Clustering and Data Aggregation ---

# Set the threshold for hard clustering. Features with a maximum membership
# probability below this threshold will not be assigned to any cluster (NA).
# Note: A value of 0.3 was used for the FCM_ORA results. A value of 0 would
# assign ALL features to their top cluster.
cluster_theshold <- 0.3

# Load differential analysis results
da_obj <- load_differential_analysis(single_matrix = TRUE, epigen = FALSE, combine_with_featgene = TRUE)

signif_da <- da_obj %>% 
  filter(adj_p_value < 0.05, contrast_type == "exercise_with_controls") %>% 
  select(feature_id, tissue, gene_symbol, assay) %>% 
  distinct()

# --- Process and Aggregate Data from All Tissues ---
# Get tissue names to iterate over
tissue_names <- names(FCM_CLUSTERS)

# Use `lapply` to process each tissue and `bind_rows` to combine the results.
all_tissues_df <- bind_rows(lapply(tissue_names, function(tissue) {
  
  # Extract the feature-by-cluster membership matrix
  membership_matrix <- FCM_CLUSTERS[[tissue]][["membership"]]
  
  # Process the membership matrix
  processed_df <- membership_matrix %>% 
    as.data.frame() %>% 
    rownames_to_column("rowname") %>% 
    # Split the feature identifier (e.g., "TRNSCRPT ENSEMBL_ID") into ome and feature_id
    separate(col = rowname, sep = " ", into = c("Ome", "feature_id"), extra = "merge") %>% 
    mutate(
      # --- Hard Clustering Step ---
      # Assign each feature to the cluster with the highest membership probability.
      max_cluster = {
        # Isolate numeric columns (cluster memberships)
        numeric_cols <- select(., where(is.numeric))
        # `max.col` efficiently finds the column index of the maximum value for each row.
        names(numeric_cols)[max.col(numeric_cols, ties.method = "first")]
      }
    ) %>% 
    rowwise() %>%
    mutate(
      # Find the maximum membership score for each feature
      max_value = max(c_across(where(is.numeric)), na.rm = TRUE)
    ) %>%
    ungroup() %>% 
    # Filter for high-confidence assignments based on the threshold
    mutate(high_cluster = if_else(max_value > cluster_theshold, max_cluster, NA)) %>% 
    # Add a column for the tissue of origin
    mutate(tissue = tissue)
  
  return(processed_df)
}))
```

### Combined Bar Graph and Dot Plot
This visualization combines a bar chart showing the omics composition of each cluster with a dot plot showing the top ORA pathway enrichments for those clusters.

```{r combined_bar_dot_plot figure 4D}
# --- Load Libraries ---
library(dplyr)
library(forcats)
library(ggplot2)
library(ggnewscale) # For using multiple 'fill' scales
library(patchwork)  # For combining plots

# --- 1. Global Plotting Parameters ---
my_base_size <- 6 
my_geom_text_size <- my_base_size / 3.5 
my_legend_point_max <- my_base_size / 1.5 # Max size of legend points
my_legend_key_size_pt <- 8 # Size of legend color squares (in points)

# --- 2. Data Preparation ---
# (Assumes center_df, all_tissues_df, and FCM_ORA are in the environment)

# Create a lookup table for cluster sign (Positive/Negative)
cluster_sign_lookup <- center_df %>%
  distinct(tissue, cluster, max_abs_sign) %>%
  mutate(cluster_num = as.numeric(as.character(cluster)))

# Create a lookup table for the timepoint of maximum effect
max_cluster_timepoint <- center_df %>% 
  select(tissue, cluster, max_abs_contrast) %>% 
  mutate(cluster = as.numeric(cluster)) %>% 
  mutate(max_abs_contrast = if_else(max_abs_contrast == "P10M", "P15-45M", max_abs_contrast)) %>% 
  distinct()

# Define ordering for tissues, contrasts, and signs
tissue_order <- c("adipose", "blood", "muscle")
contrast_order <- c("P15-45M", "P3.5/4H", "P24H")
sign_order <- c("Positive", "Negative")

# Create all possible group levels in the correct order for sorting
contrast_group_levels <- tidyr::crossing(
  max_abs_contrast = factor(contrast_order, levels = contrast_order), 
  max_abs_sign = factor(sign_order, levels = sign_order)
) %>%
  mutate(max_abs_contrast_group = paste(max_abs_contrast, max_abs_sign, sep = "_")) %>%
  pull(max_abs_contrast_group)

# --- 3. Create Bar Plot Data (plot_data_bar) ---
plot_data_bar <- all_tissues_df %>%
  group_by(tissue, high_cluster, Ome) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  mutate(high_cluster = as.numeric(as.character(high_cluster))) %>% 
  left_join(MotrpacBicQC::assay_codes, by = c("Ome" = "assay_code"), relationship = "many-to-many") %>% 
  select(tissue, high_cluster, Ome, omics_text, Count) %>% 
  mutate(omics_text = if_else(Ome == "metab", "Metabolomics", omics_text)) %>% 
  distinct() %>% 
  filter(!is.na(high_cluster)) %>% 
  left_join(max_cluster_timepoint, by = c("tissue", "high_cluster" = "cluster")) %>%
  # Join the Positive/Negative sign info
  left_join(cluster_sign_lookup, 
            by = c("tissue" = "tissue", "high_cluster" = "cluster_num")) %>%
  # Create the full sorting group
  mutate(max_abs_contrast_group = paste(max_abs_contrast, max_abs_sign, sep = "_")) %>%
  # Re-factor tissue and the new contrast group for sorting
  mutate(
    tissue = factor(tissue, levels = tissue_order),
    max_abs_contrast_group = factor(max_abs_contrast_group, levels = contrast_group_levels) 
  ) %>%
  # Arrange data based on the new factors
  arrange(tissue, max_abs_contrast_group, high_cluster) %>%
  # Create the final cluster label for the x-axis
  mutate(cluster_label = paste(high_cluster, tissue, sep = " - ")) %>%
  mutate(cluster_label = forcats::fct_inorder(cluster_label))

# --- 4. Prepare ORA Data and Select Sets for Plotting ---
FCM_ORA_top_hits <- FCM_ORA %>% 
  filter(adj_p_value < 0.05) %>% 
  mutate(set = as.character(set))

# Create a p-value lookup for all sets to use as a tie-breaker
set_pvalue_lookup <- FCM_ORA_top_hits %>%
  group_by(set) %>%
  summarise(min_p_value = min(adj_p_value)) %>%
  ungroup()

# --- Hybrid Set Selection: Manual + Greedy Algorithm ---
# Manually define a list of key pathways to include in the plot.

manually_curated_pathways = readxl::read_excel(path = file.path("~/GitHub/MoTrPAC/motrpac-human-presuspension-acute/", "figures", "landscape", "curated_pathways.xlsx"),
                                          sheet = "3")

manual_sets_to_plot <- manually_curated_pathways %>%  pull(set)

print(paste("Manually selected", length(manual_sets_to_plot), "sets:"))
print(manual_sets_to_plot)

# Determine which tissue-cluster items need to be covered by pathways
all_items_to_cover <- FCM_ORA_top_hits %>%
  mutate(tissue_cluster_id = paste(tissue, cluster, sep = "_")) %>%
  distinct(tissue_cluster_id) %>%
  pull(tissue_cluster_id)

# Create a mapping of sets to the items they cover
set_coverage_all <- FCM_ORA_top_hits %>%
  mutate(tissue_cluster_id = paste(tissue, cluster, sep = "_")) %>%
  distinct(set, tissue_cluster_id)

items_covered_by_manual <- set_coverage_all %>%
  filter(set %in% manual_sets_to_plot) %>%
  distinct(tissue_cluster_id) %>%
  pull(tissue_cluster_id)

uncovered_items <- setdiff(all_items_to_cover, items_covered_by_manual)
print(paste("Manual sets covered", length(items_covered_by_manual), "items."))
print(paste("There are", length(uncovered_items), "items left to cover."))

# --- Optional: Greedy Algorithm to Cover Remaining Items ---
# This block can be enabled to automatically select additional pathways
# to ensure every significant cluster has at least one pathway shown.
selected_sets <- manual_sets_to_plot

# # The pool of sets the greedy algorithm can pick from (excluding manual selections).
# remaining_coverage_pool <- set_coverage_all %>%
#   filter(!set %in% manual_sets_to_plot)
# 
# if (length(uncovered_items) > 0) {
#   print("Running greedy algorithm to find remaining sets...")
# 
#   while (length(uncovered_items) > 0) {
# 
#     # Find the best set from the remaining pool based on coverage and p-value.
#     best_set <- remaining_coverage_pool %>%
#       filter(tissue_cluster_id %in% uncovered_items) %>%
#       count(set, name = "coverage_count") %>%
#       left_join(set_pvalue_lookup, by = "set") %>%
#       # Sort by coverage (desc) then p-value (asc)
#       arrange(desc(coverage_count), min_p_value) %>%
#       slice(1)
# 
#     # Safety break
#     if (nrow(best_set) == 0) {
#       print("Warning: No remaining sets can cover the remaining items.")
#       break
#     }
# 
#     chosen_set_name <- best_set$set
# 
#     # Add the new set to our selected list
#     selected_sets <- c(selected_sets, chosen_set_name)
# 
#     # Find items this set just covered
#     items_just_covered <- remaining_coverage_pool %>%
#       filter(set == chosen_set_name) %>%
#       pull(tissue_cluster_id)
# 
#     # Update the list of uncovered items
#     uncovered_items <- setdiff(uncovered_items, items_just_covered)
# 
#     # Remove the chosen set from the pool
#     remaining_coverage_pool <- remaining_coverage_pool %>%
#       filter(set != chosen_set_name)
# 
#     print(paste0("...chose '", chosen_set_name, "'. ", length(uncovered_items), " items left."))
#   }
# } else {
#   print("Manual sets provided full coverage. No greedy algorithm needed.")
# }
# 
# print("---")
# print(paste("Finished. Total sets selected:", length(selected_sets)))
# print(selected_sets)

# --- 5. Create ORA Data for Clustering Pathway Order ---
# This data is used only to determine the optimal order of pathways on the y-axis.
# We aggregate by max ratio to get one value per (cluster_label, set).
plot_data_ora_raw_for_clustering <- FCM_ORA_top_hits %>%
  filter(set %in% selected_sets) %>%
  mutate(included_ratio = set_size_in_cluster/cluster_size) %>% 
  mutate(cluster_label = paste(cluster, tissue, sep = " - ")) %>%
  # Aggregate by taking the max ratio if a set appears in multiple assays for the same cluster
  group_by(cluster_label, set) %>%
  summarise(included_ratio = max(included_ratio), .groups = 'drop')

plot_data_ora_complete_for_clustering <- plot_data_ora_raw_for_clustering %>%
  mutate(
    cluster_label = factor(cluster_label, levels = levels(plot_data_bar$cluster_label)),
    set = factor(set, levels = selected_sets)
  ) %>%
  complete(cluster_label, set) %>%
  filter(!is.na(cluster_label))

# Pivot data to a wide matrix for clustering
wide_data <- plot_data_ora_complete_for_clustering %>%
  mutate(included_ratio = replace_na(included_ratio, 0)) %>%
  pivot_wider(
    id_cols = set,
    names_from = cluster_label,
    values_from = included_ratio,
    values_fill = 0 
  )

# Convert to a numeric matrix
wide_matrix <- wide_data %>%
  column_to_rownames("set") %>%
  as.matrix()

# Filter matrix for clustering (e.g., exclude 'adipose' columns)
cols_to_use <- !grepl("adipose", colnames(wide_matrix), ignore.case = TRUE)
matrix_for_clustering <- wide_matrix[, cols_to_use]

print("---")
print("Clustering will be based on these columns:")
print(colnames(matrix_for_clustering))

# Perform hierarchical clustering on rows (pathways)
row_dist <- dist(matrix_for_clustering, method = "euclidean")
row_hclust <- hclust(row_dist, method = "complete")

# Get the new row order from clustering and reverse it for plotting
new_row_indices <- row_hclust$order
new_row_order <- rownames(wide_matrix)[new_row_indices]
reversed_row_order <- rev(new_row_order)

# --- 6. Create Final ORA Data for Plotting ---
# Create the raw data with assay info, propagating all necessary columns
plot_data_ora_raw_for_plot <- FCM_ORA_top_hits %>%
  filter(set %in% selected_sets) %>%
  mutate(included_ratio = set_size_in_cluster/cluster_size) %>% 
  mutate(cluster_label = paste(cluster, tissue, sep = " - ")) %>%
  left_join(MotrpacBicQC::assay_codes, by = c("assay" = "assay_code"), relationship = "many-to-many") %>%
  mutate(omics_text = if_else(assay == "metab", "Metabolomics", omics_text)) %>%
  filter(!is.na(omics_text)) %>%
  select(cluster_label, set, included_ratio, omics_text) %>%
  distinct()

# Create the complete data frame for plotting, ensuring alignment
plot_data_ora_complete <- plot_data_ora_raw_for_plot %>%
  mutate(
    cluster_label = factor(cluster_label, levels = levels(plot_data_bar$cluster_label)),
    set = factor(set, levels = reversed_row_order)
  ) %>%
  filter(!is.na(cluster_label), !is.na(set)) %>%
  # Use complete() to add NA rows for missing combinations, ensuring alignment with the bar plot
  complete(cluster_label, set)

print("---")
print("Check the new (reversed) factor levels for 'set':")
print(levels(plot_data_ora_complete$set))

# --- 7. Create Manual Name Map for Plotting Labels ---
# This map provides shorter, more readable names for the pathways.
# Note: To build this map, you can run `dput(levels(plot_data_ora_complete$set))`
# in the console and copy the output.
set_rename_map = manually_curated_pathways$figure_name
names(set_rename_map) = manually_curated_pathways$set

# --- 8. Create Plot Enhancements (Colors & Backgrounds) ---
set_names <- levels(plot_data_ora_complete$set)
n_sets <- length(set_names)

# Define a high-contrast color palette for pathway labels
base_dark2_palette <- c(
  "#CC0000", "#006400", "#E65C00", "#0000B2", "#B8860B",
  "#008080", "#8B008B", "#556B2F", "#800000", "#4B0082"
)
# Repeat the palette if there are more sets than colors
set_colors_ordered <- rep_len(base_dark2_palette, n_sets)
set_color_map <- set_colors_ordered
names(set_color_map) <- set_names

# Data for background tissue rectangles
background_data <- plot_data_bar %>%
  mutate(x_index = as.numeric(cluster_label)) %>%
  group_by(tissue) %>%
  summarise(
    xmin = min(x_index) - 0.5,
    xmax = max(x_index) + 0.5
  ) %>%
  ungroup()

# Data for the top circles indicating cluster number and color
color_lookup <- center_df %>%
  distinct(tissue, cluster, cluster_color) %>%
  mutate(cluster_num = as.numeric(as.character(cluster))) 

circle_data <- plot_data_bar %>%
  distinct(cluster_label, high_cluster, tissue) %>%
  left_join(color_lookup, 
            by = c("high_cluster" = "cluster_num", "tissue" = "tissue"))

# --- 9. Generate Individual Plots ---

# Plot 0: Top Circles (p_top_circles)
p_top_circles <- ggplot(circle_data, aes(x = cluster_label, y = 1)) +
  geom_point(aes(fill = cluster_color), 
             shape = 21, 
             size = 3, 
             color = "black") +
  geom_text(aes(label = high_cluster), 
            size = my_geom_text_size, 
            color = "black") +
  scale_fill_identity() + 
  theme_void() + 
  theme(
    plot.margin = unit(c(2, 0, 0, 0), "pt"), 
    legend.position = "none"
  )

# Plot 1: Bar Chart (Top)
p1_vert <- ggplot(plot_data_bar, aes(x = cluster_label)) + 
  geom_rect(
    data = background_data,
    aes(xmin = xmin, xmax = xmax, fill = tissue),
    ymin = -Inf, ymax = Inf,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = HUMAN_TISSUE_COLORS) +
  guides(fill = "none") + 
  new_scale_fill() +
  geom_col(aes(y = -Count, fill = omics_text), position = "stack") +
  scale_y_continuous(labels = abs, position = "left") +
  scale_fill_manual(values = c(HUMAN_OME_COLORS, "metab" = "#6D4B08")) + 
  labs(
    y = "Number of Features", 
    x = NULL, 
    fill = "Omics Layer" 
  ) + 
  theme_minimal(base_size = my_base_size) +
  theme(
    legend.position = "left",
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(),
    # Use 'color = NA' to make gridlines invisible without warnings
    panel.grid.major.x = element_line(),
    panel.grid.major.y = element_line(color = NA),
    panel.grid.minor.y = element_line(color = NA),
    plot.margin = margin(t = 0.1, b = 0.1, unit = "pt") 
  ) +
  guides(fill = guide_legend(
    reverse = TRUE,
    keywidth = unit(my_legend_key_size_pt, "pt"),
    keyheight = unit(my_legend_key_size_pt, "pt")
  ))

# Plot 2: Dot Plot (Bottom)
p2_vert <- ggplot(plot_data_ora_complete, aes(x = cluster_label, y = set)) +
  # Background tissue rectangles
  geom_rect(
    data = background_data,
    aes(xmin = xmin, xmax = xmax, fill = tissue),
    ymin = -Inf, ymax = Inf,
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = HUMAN_TISSUE_COLORS) +
  guides(fill = "none") +
  new_scale_fill() +
  # Horizontal lines for each pathway
  geom_hline(aes(yintercept = set, color = set), linetype = "solid", alpha = 0.6) +
  # Points representing enrichment
  geom_point(
    aes(size = included_ratio, fill = omics_text),
    shape = 21,
    color = "black"
  ) +
  scale_x_discrete(position = "bottom") + 
  scale_y_discrete(
    position = "left",
    labels = set_rename_map, # Apply custom short names
    expand = expansion(mult = c(0.04, 0.07))
  ) +
  scale_color_manual(values = set_color_map) + # For hline
  scale_fill_manual(
    name = "Omics Layer",
    values = c(HUMAN_OME_COLORS, "metab" = "#6D4B08"),
    na.translate = FALSE
  ) +
  labs(
    x = NULL,
    y = "Pathway Set",
    size = "Included Ratio"
  ) +
  scale_size_continuous(range = c(1, my_legend_point_max)) + 
  theme_minimal(base_size = my_base_size) +
  theme(
    axis.text.x = element_blank(), 
    axis.ticks.x = element_blank(), 
    axis.text.y = element_text(color = set_colors_ordered),
    axis.ticks.y = element_line(),
    panel.grid.major.x = element_line(color = "grey90", linetype = "solid"),
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.margin = margin(t = -5, r = 0, b = 0, l = 0, unit = "pt"),
    legend.box.margin = margin(t = -5, r = 0, b = 0, l = 0, unit = "pt"),
    plot.margin = margin(t = 0.1, unit = "pt")
  ) +
  guides(
    color = "none", # Hide hline legend
    fill = "none"   # Hide "Omics Layer" legend
  )

# --- 10. Combine and Export Plots ---
# Combine all three plots vertically
final_plot <- p_top_circles / p1_vert / p2_vert + 
  plot_layout(
    heights = c(0.2, 1, 1.8) # Adjust relative heights
  )

# Save to PDF
pdf("combined_cluster_plot_resized_colorized.pdf", width = 7, height = 2.5)
print(final_plot)
dev.off()
```

# Feature Heatmaps of Selected Pathways

## Custom Heatmap Function with Row Annotations
This section defines a modified heatmap function that includes a custom row annotation to display C-means cluster membership, using the same colors as the combined plot above.

```{r special_heatmap_function}
#' @title Create a feature heatmap with custom row annotations
#'
#' @description A wrapper around the `feature_heatmap` function that adds a
#'   `right_annotation` to display cluster membership for each feature.
#'
#' @param feature_ids character or \code{NULL}; vector of feature IDs.
#' @param da_obj A local DA object to save time.
#' @param set_id character of \code{NULL}; if not \code{NULL}, used to select features.
#' @param platforms character or \code{NULL}; for metabolomics filtering.
#' @param selected_tissue character; the tissue(s) for the heatmap.
#' @param selected_ome character; the ome for the heatmap.
#' @param filename character; optional file name to save the heatmap.
#' @param row_annotation_matrix data.frame; data for the row annotation.
#' @param row_annotation_colors list; named list of colors for the row annotation.
#' @param ... Additional arguments passed to the internal `feature_heatmap` function.
#'
#' @returns A ComplexHeatmap object.
#'
#' @author Tyler Sagendorf, Damon Leach, Dan Katz
#'
#' @import ComplexHeatmap
#' @importFrom dplyr %>% filter mutate bind_rows
#' @importFrom grDevices dev.off
#' @importFrom tibble column_to_rownames
#' @importFrom tidyr pivot_wider

## Example Usage:
# feature_heatmap_with_RA(
#   set_id = "11725",
#   contrast_type = "exercise_with_controls",
#   selected_tissue = "muscle",
#   selected_ome = "prot-ph",
#   filename = "sandbox/test_heatmap_with_RA.pdf",
#   row_annotation_matrix = my_row_anno_data,
#   row_annotation_colors = my_row_anno_colors
# )

library(MotrpacHumanPreSuspensionAnalysis)
library(ComplexHeatmap)
library(gridtext)
library(dplyr)
library(grDevices)
library(tibble)
library(tidyr)
library(TMSig) # enrichmap

feature_heatmap_with_RA <- function(feature_ids = NULL,
                                    da_obj = NULL,
                                    set_id = NULL,
                                    platforms = NULL,
                                    selected_tissue = c("adipose", "blood", "muscle"),
                                    selected_ome = c("transcript-rna-seq", "prot-pr",
                                                     "prot-ph", "metab", "prot-ol"),
                                    contrast_type = "exercise_with_controls",
                                    column_title = "",
                                    multi_tissue_clust_rows = FALSE,
                                    filename = NULL,
                                    post_min = NULL,
                                    post_hr = NULL,
                                    full_modality_names = FALSE,
                                    row_annotation_matrix = NULL,
                                    row_annotation_colors = NULL) 
{ 
  if(is.null(da_obj)){
    DA_list <- load_differential_analysis(combine_with_featgene = TRUE)
  } else {
    DA_list <- da_obj
  }
  
  # Parameter validation
  if(!is.null(post_min)){
    if(length(post_min) > 1){ stop("If not NULL, post_min can only have length 1") }
    if(!is.numeric(post_min)){ stop("If not NULL, post_min must be numeric") }
    if(!post_min %in% c(15,30,45)){ stop("If not NULL, post_min can only take on the numeric values of 15, 30, or 45") }
  }
  if(!is.null(post_hr)){
    if(length(post_hr) > 1){ stop("If not NULL, post_hr can only have length 1") }
    if(!is.numeric(post_hr)){ stop("If not NULL, post_hr must be numeric") }
    if(!post_hr %in% c(3.5,4)){ stop("If not NULL, post_hr can only take on the numeric values of 3.5 or 4") }
  }
  if(length(full_modality_names) > 1){ stop("full_modality_names can only have length 1") }
  if(!is.logical(full_modality_names)){ stop("full_modality_names must be a logical argument") }
  
  # If set_id is provided, select the features from the set
  if (!is.null(set_id)) {
    if (!is.character(set_id) || length(set_id) != 1L) {
      stop("`set_id` must be a character string specifying a set ID from the `SET_TO_ID` object.")
    }
    pathway <- SET_TO_ID %>%
      filter(set_id == !!set_id) %>%
      pull(set)
    pathways <- unlist(structure(MOLECULAR_SIGNATURES, names = NULL), recursive = FALSE)
    feature_ids <- pathways[[pathway]]
  }
  
  if (!is.vector(feature_ids, mode = "character")) {
    stop("If `set_id` is not provided, `feature_ids` must be a character vector of feature IDs.")
  }
  
  selected_tissue <- match.arg(selected_tissue, choices = c("adipose", "blood", "muscle"), several.ok = TRUE)
  selected_ome <- match.arg(selected_ome, choices = c("transcript-rna-seq", "prot-pr", "prot-ph", "metab", "prot-ol"))
  contrast_type <- match.arg(contrast_type, choices = c("exercise_with_controls", "exercise_no_controls", "Endur_vs_Resist", "baseline", "control_only"))
  
  # --- Prepare DA results ---
  DA_res_list <- purrr::map(selected_tissue, function(st){
    filt_dat = DA_list[[st]][[selected_ome]] %>%
      filter(contrast_type == !!contrast_type) %>%
      mutate(across(.cols = any_of(c("feature_id", "gene_symbol", "platform", "flanking_sequence")), .fns = as.character)) %>%
      droplevels.data.frame()
    filt_dat
  })
  names(DA_res_list) <- selected_tissue
  DA_res = dplyr::bind_rows(DA_res_list)
  
  if (is.null(set_id)) {
    DA_res <- filter(DA_res, feature_id %in% feature_ids)
  }
  
  if (!nrow(DA_res)) {
    stop(sprintf("`feature_ids` do not match feature IDs in the %s %s differential analysis results.", selected_tissue, selected_ome))
  }
  
  # Format feature IDs based on ome
  if (selected_ome %in% c("prot-pr", "transcript-rna-seq", "prot-ol")) {
    DA_res <- DA_res %>% mutate(feature_id = ifelse(!is.na(gene_symbol), gene_symbol, feature_id))
  } else if (selected_ome == "prot-ph") {
    DA_res <- DA_res %>% mutate(feature_id = paste(gene_symbol, sub(".*_", "", sub("(.*);$", "\\1", gsub("[sty]", ";", feature_id)))))
  } else if (selected_ome == "metab" && !is.null(platforms)) {
    platforms <- match.arg(arg = platforms, choices = paste0("metab-", c("t-amines", "t-conv", "t-imm-crt", "t-imm-glc", "t-imm-ins", "t-oxylipneg", "t-tca", "u-hilicpos", "u-ionpneg", "u-lrpneg", "u-lrppos", "u-rpneg", "u-rppos")), several.ok = TRUE)
    DA_res <- DA_res %>% filter(platform %in% platforms)
    if (!nrow(DA_res)) { stop("Metabolites in `feature_ids` not found in the provided platforms.") }
  }
  
  if (!is.null(set_id)) {
    if (selected_ome %in% c("prot-pr", "transcript-rna-seq")) {
      DA_res <- filter(DA_res, feature_id %in% feature_ids)
    } else if (selected_ome == "prot-ph") {
      DA_res <- filter(DA_res, flanking_sequence %in% feature_ids)
    } else if (selected_ome == "metab") {
      DA_res <- filter(DA_res, feature_id %in% feature_ids)
    }
  }
  
  if (nrow(DA_res) == 0L) {
    stop("The combination of `set_id`, `selected_tissue`, and `selected_ome` is not valid.")
  }
  
  # if (selected_ome == "metab") {
  #   DA_res <- DA_res %>%
  #     # Include platform in the row names of the heatmap
  #     mutate(feature_id = paste(sub("^metab-", "", platform), feature_id))
  # }
  
  x <- DA_res %>%
    dplyr::mutate(contrast2 = paste(tissue, contrast),
                  contrast2 = factor(contrast2, levels = unique(contrast2))) %>%
    arrange(contrast2, abs(z.std), feature_id) %>%
    filter(.by = contrast2, !duplicated(feature_id)) %>%
    select(feature_id, contrast, contrast2, z.std, adj_p_value, tissue) %>%
    droplevels.data.frame()
  
  # --- Prepare Annotations ---
  contrast_df <- .add_contrast_labels() %>% 
    filter(contrast %in% levels(x$contrast)) %>% 
    droplevels.data.frame()
  contrast_colors_vector <- .contrast_colors()
  
  if(!is.null(post_min)){
    post_min_str = paste0("post ", post_min, " min")
    new_levels = levels(contrast_df$contrast_labels)
    new_levels[new_levels == "post 15/30/45 min"] = post_min_str
    contrast_df$contrast_labels <- factor(as.character(contrast_df$contrast_labels) %>% 
                                            `[<-`(. == "post 15/30/45 min", post_min_str), levels = new_levels)
    names(contrast_colors_vector)[names(contrast_colors_vector) == "post 15/30/45 min"] = post_min_str
  }
  
  if(!is.null(post_hr)){
    post_hr_str = paste0("post ", post_hr, " hr")
    new_levels = levels(contrast_df$contrast_labels)
    new_levels[new_levels == "post 3.5/4 hr"] = post_hr_str
    contrast_df$contrast_labels <- factor(as.character(contrast_df$contrast_labels) %>% 
                                            `[<-`(. == "post 3.5/4 hr", post_hr_str), levels = new_levels)
    names(contrast_colors_vector)[names(contrast_colors_vector) == "post 3.5/4 hr"] = post_hr_str
  }
  
  # --- Create Heatmap ---
  column_df <- distinct(x, tissue, contrast) %>%
    mutate(tissue = factor(tissue, levels = selected_tissue)) %>%
    arrange(contrast, tissue) %>%
    left_join(contrast_df, by = "contrast") %>%
    rename(modality = anno_group) %>%
    mutate(modality = factor(modality, levels = c("EE", "RE")))
  
  if(full_modality_names == TRUE){
    column_df$modality <- factor(ifelse(column_df$modality == "EE", "Endurance Exercise", "Resistance Exercise"), levels = c("Endurance Exercise", "Resistance Exercise"))
  }
  
  anno_df <- dplyr::select(column_df, Tissue = tissue, 
                           Modality = modality, 
                           Timepoint = contrast_labels) %>%
    dplyr::arrange(Tissue, Modality, Timepoint)
  
  anno_col <- list(
    "Tissue" = HUMAN_TISSUE_COLORS[selected_tissue],
    "Modality" = if(full_modality_names) setNames(c("#d95f02", "#1b9e77"), 
                                                  c("Endurance Exercise", "Resistance Exercise")) else setNames(c("#d95f02", "#1b9e77"), c("EE", "RE")),
    "Timepoint" = contrast_colors_vector[levels(anno_df$Timepoint)]
  )
  
  column_df <- column_df %>% mutate(contrast2 = factor(paste(tissue, contrast), levels = unique(paste(tissue, contrast))))
  if (length(selected_tissue) == 1L) { anno_df$Tissue <- NULL; anno_col["Tissue"] <- NULL }
  
  show_column_names <- FALSE
  if (contrast_type == "baseline") { anno_df$Timepoint <- NULL; anno_col["Timepoint"] <- NULL; show_column_names <- TRUE }
  if (!contrast_type %in% c("exercise_with_controls", "exercise_no_controls")) { anno_df$Modality <- NULL; anno_col["Modality"] <- NULL }
  
  if (length(anno_col)) {
    column_split <- if (!is.null(anno_df[["Tissue"]]) && !is.null(anno_df[["Modality"]])) {
      anno_df %>% 
        mutate(column_split = paste(Tissue, Modality), row_order = 1:n()) %>% 
        arrange(Tissue, Modality) %>% 
        mutate(column_split = factor(column_split, levels = unique(column_split))) %>% 
        arrange(row_order) %>% 
        pull(column_split)
    } else if (!is.null(anno_df[["Tissue"]])) {
      anno_df[["Tissue"]]
    } else if (!is.null(anno_df[["Modality"]])) {
      anno_df[["Modality"]]
    }
    
    top_annotation <- HeatmapAnnotation(df = anno_df, 
                                        col = anno_col, 
                                        which = "column", 
                                        border = TRUE, 
                                        gap = unit(2, "pt"), 
                                        annotation_name_gp = gpar(fontsize = 0.9 * 14), 
                                        annotation_legend_param = list(border = TRUE, 
                                                                       title_gp = gpar(fontsize = 0.9 * unit(14, "pt"), 
                                                                                       fontface = "bold"), 
                                                                       labels_gp = gpar(fontsize = 0.9 * unit(14, "pt"))))
  } else {
    top_annotation <- column_split <- NULL
  }
  
  n_features <- length(unique(x[["feature_id"]]))
  height <- max(as.numeric(convertUnit(n_features * unit(14, "pt"), "in")), 4.5) + 1.5 + show_column_names
  row_label_width <- convertUnit(ComplexHeatmap::max_text_width(text = x[["feature_id"]], gp = gpar(fontsize = 0.9 * 14)), "in")
  width <- as.numeric(convertUnit(nlevels(x[["contrast"]]) * unit(14, "pt"), "in") + row_label_width) + 5
  
  draw_args <- list(annotation_legend_list = NULL, merge_legends = TRUE)
  if (n_features <= 10L) { draw_args[["padding"]] <- unit(c(80, 0, 0, 0), "pt") }
  
  extended_range <- TMSig::extendRangeNum(x[["z.std"]], nearest = 0.1)
  breaks <- if (all(extended_range <= 0)) c(extended_range[1], 0) else if (all(extended_range >= 0)) c(0, extended_range[2]) else c(extended_range[1], 0, extended_range[2])
  
  # Handle multi-tissue clustering and row annotation filtering
  if (multi_tissue_clust_rows == TRUE){
    test_x <- x %>% group_by(feature_id) %>% 
      summarise(count = n()) %>% 
      filter(count == max(count))
    
    df_to_filter_x <- x %>% 
      left_join(row_annotation_matrix, by = c("feature_id" = "gene_symbol")) %>%
      filter(feature_id %in% test_x$feature_id)
    
    x <- df_to_filter_x %>% 
      select(-adipose, -muscle, -blood)
    
    row_annotation_matrix_filtered <- df_to_filter_x %>% 
      select(adipose, blood, muscle, feature_id) %>% 
      distinct()
    
    row_annotation_matrix_final <- row_annotation_matrix %>% select(gene_symbol) %>%
      inner_join(row_annotation_matrix_filtered, 
                 by = c("gene_symbol" = "feature_id")) %>%  
      distinct() %>%
      select(-gene_symbol)
    
    clust_row_info <- TRUE 
  } else {
    row_annotation_matrix_final <- row_annotation_matrix %>% select(-gene_symbol)
    clust_row_info <- ifelse(length(selected_tissue) > 1, FALSE, TRUE)
  }
  
  # Create row annotations if provided
  row_anno <- if (!is.null(row_annotation_matrix)) rowAnnotation(df = row_annotation_matrix_final, col = row_annotation_colors) else NULL
  
  # Draw the heatmap
  heatmap_result <- TMSig::enrichmap(
    x = x, n_top = Inf, set_column = "feature_id", statistic_column = "z.std", contrast_column = "contrast2",
    padj_column = "adj_p_value", plot_sig_only = FALSE, filename = filename, height = height, width = width,
    heatmap_color_fun = .feature_color_function,
    heatmap_args = list(
      layer_fun = .feature_layer_fun, cluster_rows = clust_row_info, column_split = column_split,
      column_labels = contrast_df$specific, show_column_names = show_column_names, column_names_side = "top",
      column_title = gt_render(column_title, padding = unit(c(0, 0, 0, 0.8), "in")),
      column_title_gp = gpar(fontsize = 12), top_annotation = top_annotation,
      right_annotation = row_anno,  # Add row annotations here
      heatmap_legend_param = list(title = "Z-Score", at = breaks, labels = breaks)
    ),
    draw_args = draw_args
  )
  
  return(heatmap_result)
}

# --- Helper Functions (re-defined for this scope) ---
.add_contrast_labels <- function() {
  # (Function content is identical to the previous definition)
  contrast_labels <- c(rep(c("during 20 min", "during 40 min", "post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr", "post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr"), times = 2L), c("post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr"), c("Endur - Resist", "Endur - Control", "Resist - Control"), c("during 20 min", "during 40 min", "post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr"))
  contrast_labels <- factor(contrast_labels, levels = c("Endur - Resist", "Endur - Control", "Resist - Control", "during 20 min", "during 40 min", "post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr"))
  anno_group <- c(rep(rep(c("EE", "RE"), c(6L, 4L)), times = 2L), rep(NA, 13L))
  return(cbind(CONTRAST_CONVERTER, contrast_labels, anno_group))
}
.contrast_colors <- function() {
  # (Function content is identical to the previous definition)
  structure(c("#fde725", "#bad071", "#d1bbd7", "#ae76a3", "#882e72", "#61194f"), names = c("during 20 min", "during 40 min", "post 10 min", "post 15/30/45 min", "post 3.5/4 hr", "post 24 hr"))
}
.feature_layer_fun <- function(j, i, x, y, w, h, f) {
  # (Function content is identical to the previous definition)
  grid.rect(x = x, y = y, width = w, height = h, gp = gpar(col = heatmap_args[["rect_gp"]][["col"]], fill = f))
  grid.text("*", x = x, y = y, gp = gpar(col = ifelse(pindex(padj_mat, i, j) < padj_cutoff, "black", NA)))
}
.feature_color_function <- function(statistics, colors = c("#3366ff", "darkred")) {
  # (Function content is identical to the previous definition)
  r <- range(statistics, na.rm = TRUE)
  extended_range <- TMSig::extendRangeNum(r, nearest = 0.1)
  max_r <- max(abs(extended_range))
  if (all(r >= 0)) { breaks <- c(0, +1) * max_r; colors <- c("white", colors[2]) } else if (all(r <= 0)) { breaks <- c(-1, 0) * max_r; colors <- c(colors[1], "white") } else { breaks <- c(-1, 0, 1) * max_r; colors <- c(colors[1], "white", colors[2]) }
  return(list(breaks = breaks, colors = colors))
}
```

## Generating Curated Heatmaps
This section uses the custom heatmap function to generate plots for specific pathways, with row annotations indicating cluster membership.
```{r}
#' @title Find pathway features from a numeric SET_ID value
#'
#' @description Basically just pulls info from SET_TO_ID for a given pathway
#'
#' @param pathway The pathway's SET_ID value from the
#'  \code{\link{MotrpacHumanPreSuspensionAnalysis::SET_TO_ID}}
#'
#' @return A character vector of values from
#'  \code{\link{MotrpacHumanPreSuspensionAnalysis::MOLECULAR_SIGNATURES}}
#' @export
#'
#' @examples {.choose_pathway_features_numeric(1)}
.choose_pathway_features_numeric = function(pathway){
  selected_pathway = SET_TO_ID %>% dplyr::filter(set_id == sprintf("%05d", pathway)) #we change pathway to a character using sprintf
  if(length(pathway) == 0)
    stop("no pathways fit this name, check again")
  actual_features = MotrpacHumanPreSuspensionAnalysis::MOLECULAR_SIGNATURES[[unique(selected_pathway$database)]][[unique(selected_pathway$set)]]
  return(actual_features)
}

# Note: I made this function before I realized `SET_TO_ID` was a thing...should probably re-write it at some point
# -Chris Sep 23, 2024

#' @title Find pathway features from a character set name
#' @description Basically just pulls info from SET_TO_ID for a given pathway. If
#' there are multiple pathways that have similar names (which actually happens
#' a lot), it just stops and tells you to be more specific.
#'
#' @param pathway  The pathway's SET_ID value from the
#'  \code{\link{MotrpacHumanPreSuspensionAnalysis::SET_TO_ID}}
#'
#' @return A character vector of values from
#'  \code{\link{MotrpacHumanPreSuspensionAnalysis::MOLECULAR_SIGNATURES}}
#' @export
#'
#' @examples .choose_pathway_features_text("BIOCARTA_41BB_PATHWAY")
.choose_pathway_features_text = function(pathway){
  all_pathways = base::unlist(MotrpacHumanPreSuspensionAnalysis::MOLECULAR_SIGNATURES, recursive = F) #unlist only to 1 layer essentially, not recursively
  desired_pathway = all_pathways[grep(pathway, names(all_pathways))]
  if(length(desired_pathway) > 1)
    stop("multiple pathways fit this name, check again")
  if(length(desired_pathway) == 0)
    stop("no pathways fit this name, check again")
  return(unlist(desired_pathway, use.names = F))
}
```




```{r prepare_heatmap_parameters}
# --- 1. Define Shared Parameters ---
CURRENT_PATHWAY <- "WP_VEGFA_VEGFR2_SIGNALING"
CURRENT_OME <- "transcript-rna-seq"
TISSUES_TO_PLOT <- c("adipose", "blood", "muscle")
CONTRAST_TO_USE <- "exercise_with_controls"

# --- 2. Prepare Custom Color List for Annotations ---
# This creates the required color mapping for the heatmap row annotations.
# The final format is a named list, e.g., list(adipose = c("1" = "#hex1"), ...).
center_df_formatted <- center_df %>%
  mutate(cluster = as.character(cluster))

# Create the named list of named color vectors
custom_row_colors <- split(center_df_formatted, center_df_formatted$tissue) %>%
  purrr::map(~ tibble::deframe(dplyr::select(., cluster, cluster_color))) %>%
  # Filter for only the tissues being plotted
  .[names(.) %in% TISSUES_TO_PLOT]
```

```{r heatmap_generator_wrapper}

#' @title Generate a Filtered Multi-Tissue Heatmap
#' @description A wrapper function to prepare data for a specific pathway and
#'   cluster, then generate a heatmap with custom row annotations.
#'
#' @param filter_tissue character. The tissue to filter by (e.g., "muscle").
#' @param filter_cluster character or numeric. The cluster(s) to filter by.
#' @param ... Additional filter conditions passed to \code{dplyr::filter()}.
#' @param pathway_name character. The pathway name or set id for feature selection.
#' @param ome_name character. The ome name for feature selection.
#' @param output_filename character. The path for the output PDF.
#' @param color_list list. The pre-computed list of custom annotation colors.
#' @param tissues_to_plot character vector. Tissues for the heatmap columns.
#' @param contrast_to_use character. The contrast type for the heatmap.
#' @return A ComplexHeatmap object or NULL if no features are found.

generate_cluster_heatmap <- function(
    filter_tissue,
    filter_cluster,
    ..., # For additional filter conditions
    pathway_name,
    ome_name,
    output_filename,
    color_list,
    tissues_to_plot = TISSUES_TO_PLOT,
    contrast_to_use = CONTRAST_TO_USE
) {
  
  # --- 1. Get Key Clusters for the Pathway/Ome ---
  key_clusters <- FCM_ORA_top_hits %>%
    filter((set == pathway_name | set_id == pathway_name) & assay == ome_name) %>%
    mutate(cluster = as.character(as.numeric(cluster))) %>%
    select(cluster, tissue) %>%
    distinct() %>%
    mutate(sig_cluster = "Yes")
  
  if(str_detect(pathway_name,pattern = "^\\d+$")==F){
    feature_list <- .choose_pathway_features_text(pathway = pathway_name)
  } else {
    feature_list <- .choose_pathway_features_numeric(pathway = as.numeric(pathway_name))
  }
  print(feature_list)
  # --- 2. Get Feature List for the Pathway ---
  if(ome_name !="metab"){
    feature_list <- feature_list %>%
      as_tibble() %>%
      dplyr::rename(gene_symbol = value) %>%
      left_join(HUMAN_FEATURE_TO_GENE) %>%
      filter(assay == ome_name) %>%
      mutate(feature_id = as.character(feature_id)) %>%
      select(feature_id) %>%
      distinct()
  } else {
    feature_list <- feature_list %>%
      as_tibble() %>%
      dplyr::rename(refmet_name = value) %>%
      left_join(HUMAN_FEATURE_TO_GENE) %>%
      filter(assay == ome_name) %>%
      mutate(feature_id = as.character(feature_id)) %>%
      select(feature_id) %>%
      distinct()
  }
  
  # --- 3. Build Row Annotation Data ---
  filter_column_name <- paste0("high_cluster_", filter_tissue)
  
  row_ha <- all_tissues_df %>%
    left_join(feature_list) %>%
    select(feature_id, tissue, high_cluster) %>%
    distinct() %>%
    left_join(key_clusters, by = c("tissue", "high_cluster" = "cluster")) %>%
    pivot_wider(id_cols = feature_id,
                values_from = c(high_cluster, sig_cluster),
                names_from = tissue) %>%
    # Filter dynamically using rlang (!!) and pass additional filters (...)
    filter(
      !!sym(filter_column_name) %in% as.character(filter_cluster),
      ...
    ) %>%
    select(feature_id,
           adipose = high_cluster_adipose,
           blood = high_cluster_blood,
           muscle = high_cluster_muscle)
  
  # --- 4. Finalize Annotation Data ---
  arranged_row_ha <- inner_join(feature_list, row_ha) %>%
    as.data.frame() %>%
    left_join(HUMAN_FEATURE_TO_GENE) %>%
    select(adipose, blood, muscle, gene_symbol, feature_id) %>%
    distinct()
  
  print(nrow(arranged_row_ha))
  print(arranged_row_ha$gene_symbol)
  
  # --- 5. Generate Heatmap ---
  if (nrow(arranged_row_ha) == 0) {
    warning(sprintf("No features found for %s cluster %s in pathway %s (with additional filters). Skipping plot.",
                    filter_tissue, paste(filter_cluster, collapse=", "), pathway_name))
    return(NULL)
  }
  
  hm <- feature_heatmap_with_RA(
    feature_ids = arranged_row_ha$feature_id,
    contrast_type = contrast_to_use,
    selected_tissue = tissues_to_plot,
    selected_ome = ome_name,
    filename = file.path(output_filename),
    multi_tissue_clust_rows = TRUE,
    full_modality_names = FALSE,
    row_annotation_matrix = arranged_row_ha,
    row_annotation_colors = color_list
  )
  cat("Generated heatmap:", output_filename, "\n")
  return(hm)
}
```

```{r generate_curated_heatmaps Figures 3F, ED5D, and ED5E}
# # Plot 1: VEGF, Adipose Cluster 10
# adipose10_VEGF <- generate_cluster_heatmap(
#   filter_tissue = "adipose",
#   filter_cluster = "10",
#   pathway_name = CURRENT_PATHWAY,
#   ome_name = CURRENT_OME,
#   output_filename = "multitissue_heatmap_adipose-10_VEGF.png",
#   color_list = custom_row_colors
# )
# 
# # Plot 2: VEGF, Blood Cluster 4 (with features also in muscle)
# blood_4_VEGF <- generate_cluster_heatmap(
#   filter_tissue = "blood",
#   filter_cluster = "4",
#   !is.na(high_cluster_muscle),
#   pathway_name = CURRENT_PATHWAY,
#   ome_name = CURRENT_OME,
#   output_filename = "multitissue_heatmap_blood-4_VEGF.png",
#   color_list = custom_row_colors
# )

# Plot 1: Early muscle/adipose overlap. CCN1 does not plot here becuase it is not measured in blood
# Corresponds to figure 3H
muscle_adipose_early_VEGF <- generate_cluster_heatmap(
  filter_tissue = "muscle",
  filter_cluster = c("10","6"),
  (high_cluster_adipose %in% c("8")) ,
  pathway_name = CURRENT_PATHWAY,
  ome_name = CURRENT_OME,
  output_filename = "multitissue_heatmap_muscle-adipose-early_overlap_VEGF.png",
  color_list = custom_row_colors
)

pdf("multitissue_heatmap_muscle-adipose-early_overlap_VEGF.pdf", width = 10, height = 10)
draw(muscle_adipose_early_VEGF, padding = unit(c(5, 1, 1, 1), "cm"))
dev.off()




# Plot 2: Middle muscle/blood overlap
# Corresponds to 3I
muscle_blood_middle_VEGF <- generate_cluster_heatmap(
  filter_tissue = "muscle",
  filter_cluster = c("12"),
  (high_cluster_blood %in% c("7","9")) ,
  pathway_name = CURRENT_PATHWAY,
  ome_name = CURRENT_OME,
  output_filename = "multitissue_heatmap_muscle-blood-middle_overlap_VEGF.png",
  color_list = custom_row_colors
)

pdf("multitissue_heatmap_muscle-blood-middle_overlap_VEGF.pdf", width = 10, height = 10)
draw(muscle_blood_middle_VEGF, padding = unit(c(5, 1, 1, 1), "cm"))
dev.off()



# Plot 3: VEGF, Muscle Clusters 6  
# Corresponds to 3J
muscle_6_VEGF <- generate_cluster_heatmap(
  filter_tissue = "muscle",
  filter_cluster = c("7"),
  #!is.na(high_cluster_adipose),
  pathway_name = CURRENT_PATHWAY,
  ome_name = CURRENT_OME,
  output_filename = "multitissue_heatmap_muscle-7_VEGF.png",
  color_list = custom_row_colors
)

pdf("multitissue_heatmap_muscle-7_VEGF.pdf", width = 10, height = 11)
draw(muscle_6_VEGF, padding = unit(c(5, 1, 1, 1), "cm"))
dev.off()



# # Plot 4: Autophagy, Muscle Clusters 12  
# # Corresponds to S3D
# muscle_12_autophagy <- generate_cluster_heatmap(
#   filter_tissue = "muscle",
#   filter_cluster = c("12"),
#   #!is.na(high_cluster_adipose),
#   pathway_name = "GOBP_PROCESS_UTILIZING_AUTOPHAGIC_MECHANISM",
#   ome_name = CURRENT_OME,
#   output_filename = "multitissue_heatmap_muscle-12_autophagy.png",
#   color_list = custom_row_colors
# )
# 
# pdf("multitissue_heatmap_muscle-12_autophagy.pdf", width = 10, height = 20)
# draw(muscle_12_autophagy, padding = unit(c(5, 1, 1, 1), "cm"))
# dev.off()
# 
# 
# # Plot 5: Autophagy, blood Clusters 4 
# # Corresponds to S3B
# blood_4_autophagy <- generate_cluster_heatmap(
#   filter_tissue = "blood",
#   filter_cluster = c("4"),
#   (high_cluster_muscle != "12"),
#   pathway_name = "GOBP_PROCESS_UTILIZING_AUTOPHAGIC_MECHANISM",
#   ome_name = CURRENT_OME,
#   output_filename = "multitissue_heatmap_blood4_autophagy.png",
#   color_list = custom_row_colors
# )
# 
# pdf("multitissue_heatmap_blood4_autophagy.pdf", width = 10, height = 15)
# draw(blood_4_autophagy, padding = unit(c(5, 1, 1, 1), "cm"))
# dev.off()
# 
# # Plot 6: Autophagy, blood Clusters 5
# # Corresponds to S3C
# blood_5_autophagy <- generate_cluster_heatmap(
#   filter_tissue = "blood",
#   filter_cluster = c("5"),
#    (high_cluster_muscle != "12"),
#   pathway_name = "GOBP_PROCESS_UTILIZING_AUTOPHAGIC_MECHANISM",
#   ome_name = CURRENT_OME,
#   output_filename = "multitissue_heatmap_blood5_autophagy.png",
#   color_list = custom_row_colors
# )
# 
# pdf("multitissue_heatmap_blood5_autophagy.pdf", width = 10, height = 15)
# draw(blood_5_autophagy, padding = unit(c(5, 1, 1, 1), "cm"))
# dev.off()

```

#Upset plot of VEGF enriched clusters

```{r ACTIVE USE}
#' @title Generate UpSet Plot with Merged Sets and Stacked Original Cluster Colors
#' @description Filters data for a specific pathway, merges sets for the main interaction 
#'   plot, but visualizes the composition of those sets using stacked bars of the 
#'   original cluster colors.
#'
#' @param pathway_name character. The name of the pathway.
#' @param ome_name character. The assay type.
#' @param merge_groups named list. Defines groups to merge.
#' @param fcm_hits_df dataframe. The FCM_ORA_top_hits data frame.
#' @param all_data_df dataframe. The all_tissues_df data frame.
#' @param cluster_info_df dataframe. Dataframe containing cluster colors. 
#'   Must have columns: "tissue", "cluster", "cluster_color". 
#'   Defaults to `center_df_formatted`.
#' @return A ComplexHeatmap UpSet plot object.

generate_pathway_upset <- function(
    pathway_name, 
    ome_name, 
    merge_groups = NULL,
    fcm_hits_df = FCM_ORA_top_hits, 
    all_data_df = all_tissues_df,
    cluster_info_df = center_df_formatted,
    top_anno_height = NULL,
    right_anno_width = NULL
) {
  
  # --- 1. Build Color Map from center_df_formatted ---
  # Create a named vector: "tissue-cluster" -> "hex_code"
  # We ensure cluster is treated as a character to match the ID generation later
  flat_color_map <- cluster_info_df %>%
    mutate(unique_id = paste0(tissue, "-", as.character(cluster))) %>%
    select(unique_id, cluster_color) %>%
    tibble::deframe()
  
  # --- 2. Identify Enriched Tissue-Cluster Combinations ---
  enriched_targets <- fcm_hits_df %>%
    filter(set == pathway_name | set_id == pathway_name, assay == ome_name) %>%
    mutate(cluster = as.character(as.numeric(cluster))) %>% 
    mutate(group_id = paste0(tissue, "-", cluster)) %>% 
    select(tissue, cluster, group_id) %>%
    distinct()
  
  if (nrow(enriched_targets) == 0) {
    warning(paste("No enriched clusters found for", pathway_name))
    return(NULL)
  }
  
  # --- 3. Get Feature List for the Pathway ---
  pathway_features_df <- .choose_pathway_features_text(pathway = pathway_name) %>%
    as_tibble() %>%
    dplyr::rename(gene_symbol = value) %>%
    left_join(HUMAN_FEATURE_TO_GENE, by = "gene_symbol") %>% 
    filter(assay == ome_name) %>%
    select(feature_id) %>%
    distinct()
  
  # --- 4. Filter Main Data (Granular Level) ---
  plot_data <- all_data_df %>%
    filter(Ome == ome_name) %>% 
    inner_join(pathway_features_df, by = "feature_id") %>% 
    mutate(cluster = as.character(high_cluster)) %>%
    mutate(group_id = paste0(tissue, "-", cluster)) %>% 
    filter(group_id %in% enriched_targets$group_id)
  
  if (nrow(plot_data) == 0) return(NULL)
  
  # --- 5. Build the "Merged" List for the UpSet Matrix ---
  input_list_granular <- split(plot_data$feature_id, plot_data$group_id)
  input_list_merged <- input_list_granular 
  
  if (!is.null(merge_groups)) {
    for (new_group_name in names(merge_groups)) {
      targets_to_merge <- merge_groups[[new_group_name]]
      valid_targets <- intersect(targets_to_merge, names(input_list_merged))
      
      if (length(valid_targets) > 0) {
        # Create merged set (Union)
        combined_features <- unique(unlist(input_list_merged[valid_targets]))
        input_list_merged[[new_group_name]] <- combined_features
        
        # Remove originals from the main list
        targets_to_remove <- setdiff(valid_targets, new_group_name)
        if (length(targets_to_remove) > 0) {
          input_list_merged[targets_to_remove] <- NULL
        }
      }
    }
  }
  
  # --- 6. Build the Stacked Bar Matrix for Right Annotation ---
  final_set_names <- names(input_list_merged)
  original_clusters_present <- names(input_list_granular)
  
  # Initialize matrix with 0s
  count_matrix <- matrix(0, 
                         nrow = length(final_set_names), 
                         ncol = length(original_clusters_present),
                         dimnames = list(final_set_names, original_clusters_present))
  
  # Fill the matrix
  for (set_name in final_set_names) {
    if (!is.null(merge_groups) && set_name %in% names(merge_groups)) {
      constituents <- intersect(merge_groups[[set_name]], original_clusters_present)
    } else {
      constituents <- intersect(set_name, original_clusters_present)
    }
    
    for (orig in constituents) {
      count_matrix[set_name, orig] <- length(input_list_granular[[orig]])
    }
  }
  
  # --- 7. Prepare Colors for the Stacked Bars ---
  bar_colors <- flat_color_map[colnames(count_matrix)]
  bar_colors[is.na(bar_colors)] <- "grey50" # Fallback
  
  # --- 8. Generate Plot ---
  comb_mat <- make_comb_mat(input_list_merged)
  
  # upset_plot <- UpSet(
  #   comb_mat,
  #   pt_size = unit(3, "mm"), 
  #   lwd = 1,
  #   
  #   # REMOVED: col = "black" (This fixes the error; UpSet defaults to black/grey)
  #   
  #   comb_order = order(comb_size(comb_mat), decreasing = TRUE),
  #   set_order = order(set_size(comb_mat), decreasing = TRUE),
  #   
  #   top_annotation = upset_top_annotation(comb_mat, add_numbers = TRUE, height = unit(top_anno_height,"cm")),
  #   
  #   # Custom Right Annotation: Stacked Bars
  #   right_annotation = rowAnnotation(
  #     "Set Size" = anno_barplot(
  #       count_matrix[set_name(comb_mat), ], 
  #       gp = gpar(fill = bar_colors, col = NA), 
  #       border = FALSE,
  #       axis_param = list(side = "bottom")
  #     ),
  #     width = unit(right_anno_width*0.6, "cm")
  #   ),
  #   
  #   column_title = paste0("Feature Overlap: ", pathway_name)
  # )
  
  # 1. Define Physical Dimensions (Inches)
  target_w <- 2.16
  target_h <- 1.68
  
  # 2. Define Proportions based on your request
  top_h_dim   <- target_h * 0.40  # 40% of height
  right_w_dim <- target_w * 0.20  # 30% of width
  
  # 3. Define Font Sizes (optimized for small figures)
  # At this small scale, 12pt is massive. We scale down to ~6-8pt.
  fs_title <- 3.5      # Title font size
  fs_label <- 3.5      # Axis labels/numbers
  fs_names <- 3.5      # Set names (rows)
  pt_size_val <- 1.5 # Dot size in mm (reduced from 3mm to fit scale)
  
  # 4. Create the Plot Object
  upset_plot <- UpSet(
    comb_mat,
    
    # Matrix Dot Styling
    pt_size = unit(pt_size_val, "mm"), 
    lwd = 1,
    
    # Sorting
    comb_order = order(comb_size(comb_mat), decreasing = TRUE),
    set_order = order(set_size(comb_mat), decreasing = TRUE),
    
    # --- Top Annotation (Intersection Size) ---
    top_annotation = upset_top_annotation(
      comb_mat, 
      add_numbers = TRUE,
      numbers_offset = unit(1, "mm"),
      height = unit(top_h_dim, "inch"),
      numbers_rot = 0,
      # Control font size of the intersection numbers
      numbers_gp = gpar(fontsize = fs_label*0.8), 
      # Control font size of the y-axis
      annotation_name_gp = gpar(fontsize = fs_label),
      axis_param = list(gp = gpar(fontsize = fs_label)) 
    ),
    
    # --- Right Annotation (Set Size / Stacked Bars) ---
    right_annotation = rowAnnotation(
      "Set Size" = anno_barplot(
        count_matrix[set_name(comb_mat), ], 
        gp = gpar(fill = bar_colors, col = NA), 
        border = FALSE,
        axis_param = list(
          side = "bottom", 
          gp = gpar(fontsize = fs_label) # Axis numbers size
        )
      ),
      # Set width exactly to 30% of total canvas
      width = unit(right_w_dim, "inch"), 
      annotation_name_gp = gpar(fontsize = fs_label)
    ),
    
    # --- Text Styling ---
    # Row names (the set names on the left of the matrix)
    row_names_gp = gpar(fontsize = fs_names),
    
    # Column Title
    column_title = paste0("Feature Overlap: ", pathway_name),
    column_title_gp = gpar(fontsize = fs_title, fontface = "bold")
  )
  
  return(upset_plot)
}
```

```{r figure ED5C}
# 1. Define Merges
my_merges <- list(
  "blood-7/9 (Combined)" = c("blood-7","blood-9"),
  "muscle-6/10 (Combined)" = c("muscle-10", "muscle-6")
)

# 2. Run
hm <- generate_pathway_upset(
  pathway_name = CURRENT_PATHWAY,
  ome_name = CURRENT_OME,
  merge_groups = my_merges,
  top_anno_height = 1.3*3, right_anno_width = 2.1*3
  # cluster_info_df = center_df_formatted (Implicit default)
)

# 3. Draw
if (!is.null(hm)) {
  draw(hm)
}

target_w <- 2.3
target_h <- 1.68


#pdf("VEGF_c-means_upset.pdf", width = 2.1*3, height = 1.3*2.5*1.5)
pdf(file = "VEGF_c-means_upset.pdf", width = target_w, height = target_h)
# use draw() to render ComplexHeatmap objects
draw(hm)
dev.off()
```

