---
title: "n_documentation_features"
output: html_document
date: "2026-03-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(MotrpacHumanPreSuspensionData)
library(MotrpacHumanPreSuspensionAnalysis)

library(ComplexHeatmap)
library(circlize)
library(grid)
library(ggplot2)

source("n_documentation_helpers.R")
```


## Review Comment

Furthermore, inconsistent sampling strategies across tissues and time points (e.g., proteomics analysis for adipose tissue was performed at only a single intermediate time point) create uneven statistical power for cross-tissue and cross-time comparisons, potentially introducing bias. The authors must transparently articulate the extent to which key conclusions are influenced by these small-sample subgroups throughout the manuscript.

## Draft Reviewer Response

We thank the reviewer for this comment. To address this concern, we compiled a comprehensive table documenting the sample sizes underlying each specifically described molecular feature in the manuscript. For assays without feature-level missingness (RNA-seq and Olink), sample sizes are uniform within a given tissue-assay-group-timepoint combination and are reported at that level. For proteomics and phosphoproteomics — where data-dependent acquisition introduces per-feature missingness — we report the observed sample size for each individual feature across every group-timepoint combination, reflecting the actual number of participants contributing to each differential abundance estimate. For metabolomics, sample sizes are reported at the platform level, as missingness in this assay is driven by metabolite detection thresholds that vary by platform rather than by individual feature. This table is provided as a supplementary resource and is referenced in the relevant results sections so that readers can assess the statistical basis for each reported finding in context.

## Overview

The goal of this document is to systematically describe the sample sizes (n) underlying each comparison performed across the landscape analysis. For any statistical result reported in the manuscript — differential abundance tests, pathway enrichment analyses, cross-tissue correlations, or multi-omics factor analyses — readers require a clear understanding of the number of participants, features, tissues, and time points that contributed to that comparison. This document therefore serves as a reference to contextualize the statistical power and generalizability of each finding, making explicit any variation in n that arises from missing data, tissue availability, outlier exclusion, or assay-specific filtering. By consolidating this information in one place, we aim to provide transparency into the statistical background for all reported results and facilitate interpretation of effect sizes and significance thresholds in light of the underlying sample structure.


This is a table of the specifically described features in the landscape paper. 
```{r}
motrpac_named_features = read.csv("motrpac_named_features.csv")
```

```{r, include = F}
all_dataset = load_qc(epigen = FALSE)

split_tissues = unlist(all_dataset, recursive = FALSE)
split_tissue_assay = unlist(split_tissues, recursive = FALSE)
qc_norm_only = split_tissue_assay[grep("qc_norm", names(split_tissue_assay))]

#makes a big list of tissues, assays
samples_per_plat = data.frame(
  list_name = names(qc_norm_only),
  vialLabels = sapply(qc_norm_only, function(x) paste(colnames(x), collapse = ";"))
) %>%
  mutate(
    tissue = sapply(strsplit(list_name, "\\."), `[`, 1),
    assay = sapply(strsplit(list_name, "\\."), `[`, 2)
  ) %>%
  dplyr::select(-list_name)

#then uses the pheno file to annotate group and timepoint info and split the vector
qc_norm_counts = lapply(seq_len(nrow(samples_per_plat)), function(row) {
  vials_vec = strsplit(samples_per_plat[row, "vialLabels"], ";")[[1]] %>%
    base::trimws()
  
  pheno$data %>%
    filter(vialLabel %in% vials_vec) %>%
    group_by(randomGroupCode, Timepoint) %>%
    summarise(
      n = n(),
      vialLabels = paste(vialLabel, collapse = ","),
      .groups = "drop"
    ) %>%
    mutate(
      tissue = samples_per_plat$tissue[row],
      assay = samples_per_plat$assay[row]
    )
}) %>%
  bind_rows() %>%
  dplyr::select(tissue, assay, randomGroupCode, Timepoint, n)
```

However, since the proteomics and phosphoproteomics have missingness patterns, the number of samples per feature are more limited. This table 1 loaded above is the most optimistic case, where there's no missingness present. 

```{r, include = F}
qc_data = load_qc(selected_omes = c("prot-pr", "prot-ph"), 
                  epigen = F)
all_group_stats = data.frame()
for(tissue in names(qc_data)){
  for(ome in names(qc_data[[tissue]])){
    curr_data = qc_data[[tissue]][[ome]][["qc_norm"]]
    if(nrow(curr_data) == 0) next
    curr_meta = qc_data[[tissue]][[ome]][["sample_metadata"]] %>%
      dplyr::filter(visitcode == "ADU_BAS")
    curr_data = curr_data[,colnames(curr_data) %in% curr_meta$vialLabel]
    curr_meta = curr_meta[match(colnames(curr_data), curr_meta$vialLabel), ] #reorder so they're in the same order as colnames of raw_counts
    
    vial_to_participant = setNames(curr_meta$pid, curr_meta$vialLabel)
    vial_to_group = setNames(curr_meta$randomGroupCode, curr_meta$vialLabel)
    vial_to_timepoint = setNames(curr_meta$Timepoint, curr_meta$vialLabel)
    
    sample_participant = vial_to_participant[colnames(curr_data)]
    sample_groups = vial_to_group[colnames(curr_data)]
    sample_timepoints = vial_to_timepoint[colnames(curr_data)]
    
    #needed to make the matrix long - probably a better way to do it, but should work.
    curr_data_long = as.data.frame(t(curr_data))
    curr_data_long$Sample <- rownames(curr_data_long)
    curr_data_long = tidyr::pivot_longer(curr_data_long, -Sample, names_to = "feature_id", values_to = "Value")
    curr_data_long$randomGroupCode = sample_groups[curr_data_long$Sample]
    curr_data_long$Timepoint = sample_timepoints[curr_data_long$Sample]
    curr_data_long$Participant = sample_participant[curr_data_long$Sample]
    
    group_stats = curr_data_long %>%
      group_by(randomGroupCode, feature_id, Timepoint) %>%
      summarize(
        Count = sum(!is.na(Value))
      )
    group_stats$tissue = tissue; group_stats$ome = ome
    
    all_group_stats = rbind(all_group_stats, group_stats)
  }
}
```


RNAseq + Olink are pretty easy because there is no missingness and just the one platform
```{r}
named_features_rna_ol = motrpac_named_features %>%
  dplyr::filter(assay %in% c("transcript-rna-seq", "prot-ol")) %>%
  dplyr::left_join(qc_norm_counts, by = c("tissue", "assay"))
```

Metab is slightly harder because I have to make sure the exact n of the platform is pulled up.
```{r}
da_platforms = load_differential_analysis(selected_omes = "metab", single_matrix = TRUE) %>%
  dplyr::select(feature_id, tissue, assay, platform) %>%
  distinct()

named_features_metab = motrpac_named_features %>%
  dplyr::filter(assay == "metab") %>%
  dplyr::left_join(da_platforms, by = c("feature_id", "tissue", "assay")) %>%
  dplyr::mutate(assay = platform) %>%
  dplyr::left_join(qc_norm_counts, by = c("tissue", "assay"))
```

```{r}
prot_omes = c("prot-pr", "prot-ph")

named_features_prot = motrpac_named_features %>%
  dplyr::filter(assay %in% prot_omes) %>%
  dplyr::left_join(all_group_stats, by = c("feature_id", "tissue", "assay" = "ome")) %>%
  rename(n = Count)
```


```{r}
named_features_with_n = dplyr::bind_rows(named_features_rna_ol, named_features_metab, named_features_prot) %>%
  mutate(randomGroupCode = gsub("ADU", "", randomGroupCode)) %>%
  dplyr::arrange(randomGroupCode, Timepoint) %>%
  tidyr::pivot_wider(
    names_from = c(randomGroupCode, Timepoint),
    values_from = n,
    names_sep = "_"
  ) %>%
  dplyr::select(-platform) %>%
  mutate(tissue = factor(tissue, levels = c("muscle", "blood", "adipose"))) %>%
  arrange(tissue, assay)
head(named_features_with_n)

# write.csv(named_features_with_n,
#           file = "outputs/named_features_w_n.csv",
#           row.names = FALSE)
```

```{r n-value-distribution}
n_vals_long = dplyr::bind_rows(named_features_rna_ol, named_features_metab, named_features_prot) 

summary(named_features_prot$n)
print(summary(n_vals_long$n))

ggplot(n_vals_long, aes(x = n)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", linewidth = 0.2) +
  facet_wrap(~assay, scales = "free_y") +
  labs(
    title = "Distribution of sample counts (n) across named features",
    x = "n (samples per feature × group × timepoint)",
    y = "Count"
  ) +
  theme_bw(base_size = 11) +
  theme(
    strip.text = element_text(size = 9),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11),
    plot.title = element_text(size = 12)
  )
```

```{r}
# Build a boolean significance matrix mirroring named_features_with_n,
# TRUE where adj_p_value < 0.05 for that feature × group × timepoint.
#
# transcript-rna-seq: named_features uses gene symbols as feature_id, but DA
# results use actual transcript feature IDs — bridge via HUMAN_FEATURE_TO_GENE.
transcript_gene_map = HUMAN_FEATURE_TO_GENE %>%
  dplyr::filter(
    gene_symbol %in% motrpac_named_features$feature_id[
      motrpac_named_features$assay == "transcript-rna-seq"
    ]
  ) %>%
  dplyr::select(feature_id, gene_symbol)

da_all_for_sig = load_differential_analysis(single_matrix = TRUE) %>%
  dplyr::filter(contrast_type == "exercise_with_controls") %>%
  dplyr::mutate(
    randomGroupCode = gsub("ADU", "", randomGroupCode),
    significant = adj_p_value < 0.05
  )

da_transcript = da_all_for_sig %>%
  dplyr::filter(assay == "transcript-rna-seq") %>%
  dplyr::inner_join(transcript_gene_map, by = "feature_id") %>%
  dplyr::mutate(feature_id = gene_symbol) %>%
  dplyr::select(-gene_symbol)

da_other = da_all_for_sig %>%
  dplyr::filter(
    assay != "transcript-rna-seq",
    feature_id %in% motrpac_named_features$feature_id
  )

sig_wide = dplyr::bind_rows(da_transcript, da_other) %>%
  dplyr::group_by(feature_id, tissue, randomGroupCode, Timepoint) %>%
  dplyr::summarise(significant = any(significant, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = c(randomGroupCode, Timepoint),
    values_from = significant,
    names_sep = "_"
  )

# For each Endur/Resist column that is TRUE, mark the matching Control column TRUE too.
# Control samples are included in the exercise_with_controls model, so if EE or RE is
# significant at a timepoint the control observations contributed to that estimate.
er_cols = grep("^(Endur|Resist)_", colnames(sig_wide), value = TRUE)
for (col in er_cols) {
  tp = sub("^(Endur|Resist)_", "", col)
  con_col = paste0("Control_", tp)
  er_true = !is.na(sig_wide[[col]]) & sig_wide[[col]]
  if (con_col %in% colnames(sig_wide)) {
    sig_wide[[con_col]] = sig_wide[[con_col]] | er_true
  } else {
    sig_wide[[con_col]] = er_true
  }
}

# If any post-baseline timepoint for a group is significant, also mark that
# group's pre-exercise column TRUE (the baseline sample contributed to every
# within-group contrast for that feature).
for (grp in c("Control", "Endur", "Resist")) {
  pre_col = paste0(grp, "_pre_exercise")
  grp_cols = setdiff(grep(paste0("^", grp, "_"), colnames(sig_wide), value = TRUE), pre_col)
  if (length(grp_cols) == 0) next
  any_sig = apply(
    sig_wide[, grp_cols, drop = FALSE], 1,
    function(row) any(!is.na(row) & row)
  )
  if (pre_col %in% colnames(sig_wide)) {
    sig_wide[[pre_col]] = sig_wide[[pre_col]] | any_sig
  } else {
    sig_wide[[pre_col]] = any_sig
  }
}
```


```{r, fig.width=14, fig.height=28}
group_max_n = c(Control = 37, Endur = 65, Resist = 73)

n_cols = grep("^(Control|Endur|Resist)_", colnames(named_features_with_n), value = TRUE)
col_groups = sub("_.*", "", n_cols)

group_cols = c(Control = "#7570b3", Endur = "#d95f02", Resist = "#1b9e77")
col_fun_pct = colorRamp2(c(0, 1), c("white", "#2171b5"))
assay_cols = HUMAN_OME_COLORS[names(HUMAN_OME_COLORS) %in% unique(named_features_with_n$assay)]
tissue_cols = HUMAN_TISSUE_COLORS

col_anno = HeatmapAnnotation(
  Group = col_groups,
  col = list(Group = group_cols),
  show_annotation_name = FALSE,
  annotation_legend_param = list(Group = list(title = "Group"))
)

pct_lg = ComplexHeatmap::Legend(
  title = "% of max N\n(CON=37, EE=65, RE=73)",
  col_fun = col_fun_pct,
  at = c(0, 1),
  labels = c("0%", "100%"),
  legend_height = grid::unit(4, "cm")
)
assay_lg = ComplexHeatmap::Legend(
  title = "Assay",
  at = names(assay_cols),
  legend_gp = grid::gpar(fill = assay_cols)
)
tissue_lg = ComplexHeatmap::Legend(
  title = "Tissue",
  at = names(tissue_cols),
  legend_gp = grid::gpar(fill = tissue_cols)
)
na_lg = ComplexHeatmap::Legend(
  title = "No measurement done",
  at = "NA",
  legend_gp = grid::gpar(fill = "grey65", col = NA)
)

.make_ht = function(df, use_cols = n_cols, sig_df = NULL) {
  use_groups = sub("_.*", "", use_cols)
  mat = as.matrix(df[, use_cols])
  rownames(mat) = df$feature_id
  pct_mat = mat
  for (j in seq_along(use_cols)) pct_mat[, j] = mat[, j] / group_max_n[use_groups[j]]

  # Boolean matrix: TRUE where that group × timepoint is DA-significant (adj_p < 0.05)
  sig_mat = matrix(FALSE, nrow = nrow(df), ncol = length(use_cols),
                   dimnames = list(df$feature_id, use_cols))
  if (!is.null(sig_df)) {
    sig_cols = intersect(use_cols, colnames(sig_df))
    joined = df %>%
      dplyr::select(feature_id, tissue) %>%
      dplyr::left_join(
        sig_df %>% dplyr::select(feature_id, tissue, dplyr::all_of(sig_cols)),
        by = c("feature_id", "tissue")
      )
    for (col in sig_cols) {
      vals = joined[[col]]
      sig_mat[, col] = !is.na(vals) & vals
    }
  }

  row_anno = rowAnnotation(
    Assay  = df$assay,
    Tissue = df$tissue,
    col    = list(Assay = assay_cols, Tissue = tissue_cols),
    show_annotation_name = TRUE,
    annotation_name_side = "top",
    show_legend = c(Assay = FALSE, Tissue = FALSE)
  )

  col_anno_sub = HeatmapAnnotation(
    Group = use_groups,
    col = list(Group = group_cols),
    show_annotation_name = FALSE,
    annotation_legend_param = list(Group = list(title = "Group"))
  )

  Heatmap(
    pct_mat,
    name = "pct",
    col = col_fun_pct,
    na_col = "grey65",
    cell_fun = function(j, i, x, y, width, height, fill) {
      val = mat[i, j]
      if (!is.na(val)) {
        grid::grid.text(as.character(val), x, y,
                        gp = grid::gpar(fontsize = 11, col = "black"))
        if (isTRUE(sig_mat[i, j])) {
          grid::grid.rect(x, y, width, height,
                          gp = grid::gpar(col = "black", fill = NA, lwd = 2))
        }
      }
    },
    top_annotation = col_anno_sub,
    left_annotation = row_anno,
    row_split = df$tissue,
    row_gap = grid::unit(2, "mm"),
    column_split = use_groups,
    column_gap = grid::unit(2, "mm"),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_row_names = TRUE,
    show_column_names = TRUE,
    column_names_rot = 45,
    row_names_gp = grid::gpar(fontsize = 10),
    column_names_gp = grid::gpar(fontsize = 8),
    row_title_rot = 0,
    row_title_gp = grid::gpar(fontsize = 9),
    show_heatmap_legend = FALSE
  )
}

muscle_cols = n_cols[!grepl("during|post_10_min", n_cols)]
muscle_df = named_features_with_n %>% dplyr::filter(tissue == "muscle")
blood_adi_df = named_features_with_n %>% dplyr::filter(tissue %in% c("blood", "adipose"))

ht_muscle = .make_ht(muscle_df, use_cols = muscle_cols, sig_df = sig_wide)
ht_blood_adi = .make_ht(blood_adi_df, sig_df = sig_wide)

png("outputs/n_named_features_muscle.png", width = 14, height = 14, units = "in", res = 150)
ComplexHeatmap::draw(
  ht_muscle,
  heatmap_legend_list = list(pct_lg, assay_lg, tissue_lg, na_lg),
  merge_legend = TRUE,
  heatmap_legend_side = "left"
)
dev.off()

#we split this into 2 because the figure would be too large otherwise and the text would be unreadable. 

png("outputs/n_named_features_blood_adipose.png", width = 14, height = 14, units = "in", res = 150)
ComplexHeatmap::draw(
  ht_blood_adi,
  heatmap_legend_list = list(pct_lg, assay_lg, tissue_lg, na_lg),
  merge_legend = TRUE,
  heatmap_legend_side = "left"
)
dev.off()
```


## Sample Count Distributions

Distribution of observed sample counts for SCION-imputed features (muscle phosphoproteomics).
```{r, eval = FALSE, fig.width = 8, fig.height = 5}
imputed_data = MotrpacHumanPreSuspensionData::load_qc(selected_tissues = "muscle", selected_omes = "prot-ph")
imputed_features = rownames(imputed_data$muscle$`prot-ph`$qc_imputed)
imputed_group_stats = all_group_stats %>% filter(
  feature_id %in% imputed_features,
  tissue == "muscle"
)
plot_group_stats_histogram(
  imputed_group_stats,
  title = "Distribution of observed sample counts for SCION-imputed features",
  subtitle = "Muscle phosphoproteomics (prot-ph); each observation is one feature × group × timepoint"
)
ggsave("outputs/scion_imputation_n_histogram.png", width = 8, height = 5, units = "in", dpi = 600)
```

Slight modification to this figure. We want stacked barcharts for sig features and non-sig. features. 

For a feature to be considered "significant" we include any comparison involved in the exercise delta-delta comparison. This means, for example, for a comparison of PGC1a at resistance at 3.5 hr, the RE group at 3.5 hour, the RE pre-exercise group, and 

We include both `exericse_with_controls` and `Endur_vs_Resist`

```{r, include = FALSE}
significant_prot_pr_ph_features = load_differential_analysis(
  selected_omes = c("prot-pr", "prot-ph"),
  selected_tissues = "all",
  single_matrix = TRUE
) %>%
  dplyr::filter(contrast_type %in% c("exercise_with_controls", "Endur_vs_Resist"),
                adj_p_value < 0.05) %>%
  dplyr::select(feature_id, tissue, assay, contrast) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    terms = stringr::str_extract_all(
      gsub("group_timepoint", "", contrast),
      "ADU[A-Za-z]+\\.[\\w\\.]+"
    )
  ) %>%
  tidyr::unnest(terms) %>%
  dplyr::group_by(feature_id, tissue, assay, contrast) %>%
  dplyr::mutate(
    randomGroupCode = stringr::str_split_fixed(terms, "\\.", 2)[, 1],
    Timepoint = stringr::str_split_fixed(terms, "\\.", 2)[, 2]
  ) %>%
  dplyr::ungroup() %>%
  dplyr::rename(ome = assay) %>%
  dplyr::select(feature_id, tissue, ome, randomGroupCode, Timepoint) %>%
  dplyr::distinct()

sig_prot_pr_ph_stats = all_group_stats %>%
  dplyr::semi_join(significant_prot_pr_ph_features,
                   by = c("feature_id", "tissue", "ome", "randomGroupCode", "Timepoint"))

non_sig_prot_pr_ph_stats = all_group_stats %>%
  dplyr::anti_join(significant_prot_pr_ph_features,
                   by = c("feature_id", "tissue", "ome", "randomGroupCode", "Timepoint"))
```

```{r, eval = TRUE, fig.width = 12, fig.height = 8}
plot_sig_nonsig_histogram(
  sig_prot_pr_ph_stats,
  non_sig_prot_pr_ph_stats,
  title = "Observed sample counts for significant vs. non-significant (phospho)proteomic features",
  subtitle = "Each panel is one tissue × assay; each observation is one feature × group × timepoint",
  vline = 3
)
ggsave("outputs/sig_nonsig_prot_n_histogram.png", width = 10, height = 8, units = "in", dpi = 600)
```

Same figure restricted to the specifically named features in the manuscript, across all omes.
```{r, eval = TRUE, fig.width = 14, fig.height = 10}
group_recode = c(Control = "ADU_CON", Endur = "ADU_EE", Resist = "ADU_RE")
named_features_long = named_features_with_n %>%
  tidyr::pivot_longer(
    cols = matches("^(Control|Endur|Resist)_"),
    names_to = c("Group", "Timepoint"),
    names_pattern = "^(Control|Endur|Resist)_(.*)",
    values_to = "Count"
  ) %>%
  mutate(randomGroupCode = group_recode[Group]) %>%
  filter(!is.na(Count))
plot_ome_stats_histogram(
  named_features_long,
  title = "Distribution of observed sample counts for features highlighted in the manuscript",
  subtitle = "Each color is one assay; each observation is one feature × group × timepoint"
)
ggsave("outputs/named_all_omes_n_histogram.png", width = 10, height = 5, units = "in", dpi = 600)
```
