---
title: "control_only_analysis"
output: html_document
date: "2026-03-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(dplyr)
library(ggplot2)
library(grid)

source(here::here("revisions/landscape/control_analysis/plot_control_da_matrix.R"))
```


## Review Comment: 

The inclusion of a non-exercise resting control group is a strength, allowing for control of circadian rhythms, fasting, and handling stress. However, the manuscript fails to fully quantify and present these "non-exercise effects." I suggest a supplementary analysis that systematically reports the molecular features changing over time in the control group. This will allow for a clearer isolation of responses induced purely by exercise.

## How to address this comment

To some extent, I think this reviewer only put this comment because we somewhat emphasized the control aspects of the analysis in Fig S2A-B.

We already have a pretty good amount of control-only analysis. 

```{r}
control_only_analysis = load_differential_analysis(single_matrix = TRUE) %>%
  filter(contrast_type == "control_only")
```

```{r, fig.width = 14, fig.height = 3}
plot_control_da_matrix(control_only_analysis)
```

## Top control DA features

```{r, include = FALSE}
top_control_features = control_only_analysis %>%
  dplyr::group_by(feature_id) %>%
  dplyr::slice_min(adj_p_value, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::filter(adj_p_value < 0.05) %>%
  dplyr::arrange(adj_p_value) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::pull(feature_id)
```

```{r, results = "asis"}
for (feat in top_control_features) {
  tryCatch({
    cat("\n\n###", feat, "\n\n")
    print(plot_single_feature(feat))
  }, error = function(e) {
    message("Skipping ", feat, ": ", conditionMessage(e))
  })
}
```


## Enrichments

```{r}
control_camera = CAMERA_RESULTS %>%
  filter(contrast_type == "control_only",
         collection != "CELLMARKER")

top_control_pathways = control_camera %>%
  dplyr::group_by(tissue, set_id) %>%
  dplyr::slice_min(adj_p_value, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(tissue, adj_p_value) %>%
  dplyr::group_by(tissue) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::ungroup() %>%
  dplyr::pull(set_id) %>%
  unique() %>%
  as.character()

enrich_pdf = tempfile(fileext = ".pdf")
plot_enrich_heatmap(
  x = CAMERA_RESULTS,
  set_ids = top_control_pathways,
  contrast_type = "control_only",
  filename = enrich_pdf
)
```

<iframe src="`r enrich_pdf`" width="100%" height="600px"></iframe>

And now to figure out why the controls are going down so much in oxphos related pathways?

```{r, fig.width = 10, fig.height = 20}
oxphos_setid = SET_TO_ID %>% filter(set_short == "OXPHOS") %>% pull(set_id)
plot_feature_heatmap(set_id = oxphos_setid,
                     selected_ome = "transcript-rna-seq",
                     contrast_type = "control_only",
                     max_size = 100)
```

Quick check for vo2 max info
```{r}
controls = load_qc(selected_tissues = "muscle",
                   selected_omes = "transcript-rna-seq")
control_meta = controls$muscle$`transcript-rna-seq`$sample_metadata %>%
  filter(randomGroupCode == "ADUControl")

vo2_info = MotrpacHumanPreSuspensionData::cln_curated_screening[["data"]]

all_control_pids = control_meta %>% pull(pid) %>% unique()

pids_pre = control_meta %>%
  filter(Timepoint == "pre_exercise") %>%
  pull(pid)

pids_post24 = control_meta %>%
  filter(Timepoint == "post_24_hr") %>%
  pull(pid)

pids_not_post24 = setdiff(all_control_pids, pids_post24)

vo2_plot_data = dplyr::bind_rows(
  vo2_info %>% filter(pid %in% pids_pre) %>% mutate(timepoint_group = "pre_exercise"),
  vo2_info %>% filter(pid %in% pids_post24) %>% mutate(timepoint_group = "post_24_hr"),
  vo2_info %>% filter(pid %in% pids_not_post24) %>% mutate(timepoint_group = "not post_24_hr")
) %>%
  mutate(timepoint_group = factor(timepoint_group,
                                  levels = c("pre_exercise", "post_24_hr", "not post_24_hr")))
```

```{r}
ggplot(vo2_plot_data, aes(x = timepoint_group, y = VO2max_L)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_beeswarm(size = 1.5, alpha = 0.7) +
  theme_bw() +
  xlab("Timepoint group") +
  ylab("VO2max (L)")
```

# Other theories 

This could be a cell-type related answer too. I think it's possible that the increase in the immune response means less muscle cells => lower contribution from muscle cells => lower representation in oxphos related transcripts...? 
