---
title: "sex_baseline_sensitivity_analysis"
output: html_document
date: "2025-04-10"
---


```{r}
library(dplyr)
library(tidyr)
library(MotrpacHumanPreSuspension)
library(MotrpacBicQC)
library(ggplot2)
library(stringr)
library(variancePartition)
library(edgeR)

knitr::opts_chunk$set(
  fig.width=8, 
  fig.height=4,
  out.width = '100%',
  fig.dpi=300,
  cache = FALSE
)

source("sex_diff_helpers.R")
```

# Motivation of this document

The initial models for sex-differences at baseline were built directly off of the
models from the larger exercise-acute bout trajectories. Those were informed by sensitivity
analysis from `/DA/DA_Sensitivity_Analysis/cov_sensitivity_analysis.Rmd`

However, as we focus on the baseline sex differences, which are essentially just
comparisons of the pre-exercise timepoint for M vs F participants independent of 
acute exercise, the number of samples that we're analyzing is quite a bit smaller 
than the overall acute bout (esp for prot-pr/ph). So we repeat a similar analysis
and report back in this document to see what type of analysis the models can support. 

The code parameters are a little outdated in some areas, and the code is a little clunky, but the conclusions are still here. 

```{r}
#so I'm basically just expanding out the DA functions for given omes so I can test and save intermediates and visualize num parameters, fit, etc. 
desired_ome = "transcript-rna-seq"; tissue = 'adipose'
parsed_qc_norm = load_qc(selected_omes = desired_ome, 
                         selected_tissues = tissue, 
                         load_acute_only = T)
qc_norm_mat = parsed_qc_norm[[tissue]][[desired_ome]][["qc_norm"]]
#-> so here we want to make sure the genes we chose and the participants we chose are the same as the ones that pass through expression logcpm cutoffs
metadata = parsed_qc_norm[[tissue]][[desired_ome]][['sample_metadata']] %>% 
  filter(Timepoint == "pre_exercise") %>%
  mutate(vialLabel = as.character(vialLabel))

qc_norm_mat = qc_norm_mat %>% select(any_of(metadata$vialLabel))
rownames(metadata) = metadata$vialLabel
process_metadata = process_covariates(meta = metadata, selected_ome = desired_ome, tissue_input = tissue)
meta_matrix = process_metadata$metadata

metadata_feat = cbind(t(qc_norm_mat), meta_matrix)
bic_table = data.frame()

for(feat in colnames(metadata_feat)[1:nrow(qc_norm_mat)]){
  #remember for adipose we don't do Batch regardless, due to the batching issues. 
  full_formula = paste0(feat, "~ Sex + randomGroupCode +  calculatedAge + codedsiteid + pct_umi_dup + RIN") %>% as.formula()
  without_site = paste0(feat, "~ Sex + randomGroupCode + calculatedAge + Batch + pct_umi_dup + RIN") %>% as.formula()
  without_batch_or_site = paste0(feat, "~ Sex + randomGroupCode + calculatedAge + pct_umi_dup + RIN") %>% as.formula()
  without_batch_group_site = paste0(feat, "~ Sex +  calculatedAge + pct_umi_dup + RIN") %>% as.formula()
  only_sex = paste0(feat, "~ Sex") %>% as.formula()
  
  
  full_fit = lm(formula = full_formula, data = metadata_feat)
  without_site_fit = lm(formula = without_site, data = metadata_feat)
  without_batch_or_site_fit = lm(formula = without_batch_or_site, data = metadata_feat)
  without_batch_group_site_fit = lm(formula = without_batch_group_site, data = metadata_feat)
  only_sex_fit = lm(formula = only_sex, data = metadata_feat)
  
  
  bic_table = rbind(bic_table,
                    c(BIC(full_fit), "original model", feat),
                    c(BIC(without_site_fit), "without_site", feat),
                    c(BIC(without_batch_or_site_fit), "without_batch_or_site", feat),
                    c(BIC(without_batch_group_site_fit), "without_batch_group_site", feat),
                    c(BIC(only_sex_fit), "only_sex", feat)
  )
}
colnames(bic_table) = c("BIC", "Model", "Feature")

bic_table = bic_table %>%
  mutate(BIC = as.numeric(BIC)) %>%
  group_by(Feature) %>%
  mutate(mean_BIC = mean(BIC),
         delta_BIC_from_mean = BIC - mean_BIC) %>%
  ungroup() 

model_summary = bic_table %>%
  group_by(Model) %>%
  summarise(mean_delta_BIC_from_mean = mean(delta_BIC_from_mean, na.rm = TRUE)) %>%
  mutate(ome = desired_ome,
         tissue = tissue) %>%
  arrange(mean_delta_BIC_from_mean)
print(model_summary)

```

We do the same analysis for proteomics
```{r, warning = F, message=F}
prot_omes = c('prot-pr', 'prot-ph')
for(desired_ome in prot_omes){
  for(tissue in c("adipose")){ #desired_tissues){
    if(tissue == "blood") next #prot-pr has no blood tissue
    ome_data = load_qc(selected_tissues = tissue,
                       selected_omes = desired_ome,
                       load_acute_only = FALSE)
    
    metadata = ome_data[[tissue]][[desired_ome]][['sample_metadata']] %>% 
      dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
    rownames(metadata) = metadata$vialLabel
    data_matrix = ome_data[[tissue]][[desired_ome]][['qc_imputed']] %>%
      dplyr::select(as.character(metadata$vialLabel))
    rownames(data_matrix) = paste0("feat", gsub("[-/;:',^+() ]", "_",  rownames(data_matrix)))
    process_metadata = process_covariates(meta = metadata,
                                          selected_ome = desired_ome,
                                          tissue_input = tissue)
    
    meta_matrix = process_metadata$metadata
    metadata_feat = cbind(t(data_matrix), meta_matrix) 
    bic_table = data.frame()
    for(feat in colnames(metadata_feat)[1:nrow(qc_norm_mat)]){
      if(is.na(feat)) next
      
      full_formula = paste0(feat, "~ Sex + randomGroupCode + calculatedAge + codedsiteid + tmt_plex") %>% as.formula()
      without_site = paste0(feat, "~ Sex + randomGroupCode + calculatedAge + tmt_plex") %>% as.formula()
      without_batch_or_site = paste0(feat, "~ Sex + randomGroupCode + calculatedAge") %>% as.formula()
      without_batch_group_site = paste0(feat, "~ Sex + calculatedAge") %>% as.formula()
      only_sex = paste0(feat, "~ Sex") %>% as.formula()
      
      full_fit = lm(formula = full_formula, data = metadata_feat)
      without_site_fit = lm(formula = without_site, data = metadata_feat)
      without_batch_or_site_fit = lm(formula = without_batch_or_site, data = metadata_feat)
      without_batch_group_site_fit = lm(formula = without_batch_group_site, data = metadata_feat)
      only_sex_fit = lm(formula = only_sex, data = metadata_feat)
      
      bic_table = rbind(bic_table,
                        c(BIC(full_fit), "original model", feat),
                        c(BIC(without_site_fit), "without_site", feat),
                        c(BIC(without_batch_or_site_fit), "without_batch_or_site", feat),
                        c(BIC(without_batch_group_site_fit), "without_batch_group_site", feat),
                        c(BIC(only_sex_fit), "only_sex", feat)
                        
      )
    }
    colnames(bic_table) = c("BIC", "Model", "Feature")
    
    bic_table = bic_table %>%
      mutate(BIC = as.numeric(BIC)) %>%
      group_by(Feature) %>%
      mutate(mean_BIC = mean(BIC),
             delta_BIC_from_mean = BIC - mean_BIC) %>%
      ungroup() 
    
    model_summary = bic_table %>%
      group_by(Model) %>%
      summarise(mean_delta_BIC_from_mean = mean(delta_BIC_from_mean, na.rm = TRUE)) %>%
      mutate(ome = desired_ome,
             tissue = tissue) %>%
      arrange(mean_delta_BIC_from_mean)
    print(model_summary)
  }
}
```



And for metabolomics
```{r, warning = F, message=F}
bic_table = data.frame()
for(metab_platform in metab_only_list()){
  desired_ome = metab_platform
  for(tissue in c("adipose")){
    ome_data = load_qc(selected_tissues = tissue,
                       selected_omes = desired_ome,
                       load_acute_only = FALSE)
    if(length(ome_data) == 0) next 
    
    metadata = ome_data[[tissue]][[desired_ome]][['sample_metadata']] %>% 
      dplyr::filter(Timepoint == "pre_exercise") #because we're just looking for baseline measurements
    rownames(metadata) = metadata$vialLabel
    data_matrix = ome_data[[tissue]][[desired_ome]][['qc_norm']] %>%
      dplyr::select(as.character(metadata$vialLabel))
    process_metadata = process_covariates(meta = metadata,
                                          selected_ome = desired_ome,
                                          tissue_input = tissue)
    
    meta_matrix = process_metadata$metadata
    rownames(data_matrix) = paste0("feat", gsub("[-/;:',^+() ]", "_",  rownames(data_matrix)))
    metadata_feat = cbind(t(data_matrix), meta_matrix) 
    
    for(feat in colnames(metadata_feat)[1:nrow(qc_norm_mat)]){
      if(is.na(feat)) next
      full_formula = paste0(feat, "~ Sex + randomGroupCode + calculatedAge + codedsiteid") %>% as.formula()
      without_site = paste0(feat, "~ Sex + randomGroupCode + calculatedAge") %>% as.formula()
      without_batch_group_site = paste0(feat, "~ Sex + calculatedAge") %>% as.formula()
      only_sex = paste0(feat, "~ Sex") %>% as.formula()
      
      full_fit = lm(formula = full_formula, data = metadata_feat)
      without_site_fit = lm(formula = without_site, data = metadata_feat)
      without_batch_group_site_fit = lm(formula = without_batch_group_site, data = metadata_feat)
      only_sex_fit = lm(formula = only_sex, data = metadata_feat)
      
      bic_table = rbind(bic_table,
                        c(BIC(full_fit), "original model", feat),
                        c(BIC(without_site_fit), "without_site", feat),
                        c(BIC(without_batch_group_site_fit), "without_batch_group_site", feat),
                        c(BIC(only_sex_fit), "only_sex", feat)
                        
      )
    }
  }
}
colnames(bic_table) = c("BIC", "Model", "Feature")

bic_table = bic_table %>%
  mutate(BIC = as.numeric(BIC)) %>%
  group_by(Feature) %>%
  mutate(mean_BIC = mean(BIC),
         delta_BIC_from_mean = BIC - mean_BIC) %>%
  ungroup() 

model_summary = bic_table %>%
  group_by(Model) %>%
  summarise(mean_delta_BIC_from_mean = mean(delta_BIC_from_mean, na.rm = TRUE)) %>%
  mutate(ome = "metab",
         tissue = tissue) %>%
  arrange(mean_delta_BIC_from_mean)
print(model_summary)

```

So what is the verdict from these results?

Generally, the suggestion from BIC is that the pre-exercise timepoint doesn't have 
a very large number of samples to support the parameters in that are otherwise
included in the acute bout of exercise.

The BIC model diagnostics indicates that just including essentially only "Sex" and
age seems to be the most appropriate for the
