---
title: "MoTrPAC Genotype Quality Metrics"
output: html_document
date: "2022-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(UpSetR)
```

```{r}
ggplot.theme <- theme_minimal(base_size=18) +
  theme(
    panel.border=element_blank(),
    panel.grid=element_blank(),
    axis.line.x.bottom=element_line("black", 0.25),
    axis.line.y.left=element_line("black", 0.25)
  )
```

# Load Data

```{r}
DATA_DIR <- "/projects/motrpac/PRECOVID/DATA/GENOTYPE/FILTERED/"

depth.samples <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.mean_depth_individual.idepth"))
depth.sites <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.mean_depth_site.ldepth.mean"))

heterozygosity <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.heterozygosity.het"))

allele.freq <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.allele_frequencies.frq"))

hwe <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.hwe_statistics.hwe"))

site.qual <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.site_quality.lqual"))

missing.samples <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.missing.imiss"))
missing.sites <- fread(paste0(DATA_DIR, "merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.missing.lmiss"))
```

# Mean Depth per Individual

The mean depth (number of reads overlapping the variant) for each sample. This
ranges from 25.9774 to 41.0883 across the data. Depth is only calculated for samples
that have the alternate variant.

```{r}
range(depth.samples$MEAN_DEPTH)

depth.samples %>%
  as.data.frame() %>%
  ggplot(aes(x=MEAN_DEPTH)) +
  geom_histogram(binwidth=1, fill="grey60", alpha=0.5) +
  xlab("Mean Depth per Individual") + ylab("Frequency") +
  ggplot.theme
```

# Mean Depth per Site

Depth is only calculated for samples that have the alternate variant. That is,
samples that did not report the variant (and automatically considered reference
homozygotes) do not contribute to the measurement of depth.

The mean depth per site ranged from 1.875 to 4086.660 with a median of 31.896 
reads per alternate variant.

```{r}
summary(depth.sites$MEAN_DEPTH)

depth.sites %>%
  as.data.frame() %>%
  ggplot(aes(x=MEAN_DEPTH)) +
  geom_histogram(fill="grey60", alpha=0.5, bins=30) +
  xlab("Mean Depth per Site") + ylab("Frequency") +
  scale_x_log10() +
  ggplot.theme
```

# Hardy-Weinberg Equilibrium

Most variants are expected to be in Hardy-Weinberg Equilibrium (HWE). That is,
the number of observed homozygotes and heterozygotes should match those seen
under HWE.


```{r}
 hwe %>%
  as.data.frame() %>%
  dplyr::mutate(P.Interval=cut(P, breaks=seq(0, 1, length.out=101))) %>%
  dplyr::count(P.Interval) %>%
  dplyr::mutate(P.Interval=(as.numeric(P.Interval) - 0.5) / 100) %>%
  ggplot() +
  geom_bar(aes(x=P.Interval, y=n), stat="identity", fill="grey60", alpha=0.5) +
  xlab("HWE p-value") + ylab("Frequency") +
  ggplot.theme

ggsave("results/motrpac_hwe_p_values.svg", width=12, height=8)
```

# Heterozygosity

Heterozygosity is measured using the inbreeding coefficient $F$. It is defined
as 1 minus the ratio of observed heterozygotes to expected heterozygotes under
HWE. Therefore, low values of $F$ represent high heterozygosity and low
inbreeding.

This value varies from -0.09212 to 0.16777.

```{r}
range(heterozygosity$F)

heterozygosity %>%
  as.data.frame() %>%
  ggplot(aes(x=F)) +
  geom_histogram(fill="grey60", alpha=0.5, bins=30) +
  xlab("Inbreeding Coefficient") + ylab("Frequency") +
  ggplot.theme
```

# Allele Frequency Spectrum

The SNPs have been filtered to MAF > 0.025. The allele frequency spectrum should
show a larger number of rare variants compared to common variants.

```{r}
allele.freq %>%
  as.data.frame() %>%
  ggplot(aes(x=MAF)) +
  geom_histogram(fill="grey60", alpha=0.5, bins=50) +
  xlab("Minor Allele Frequency") + ylab("Frequency") +
  ggplot.theme

ggsave("results/motrpac_allele_frequency_spectrum.svg", width=12, height=8)
```

# Site Quality

The site quality is reported for each site. The quality is a PHRED score, which
is $q = -10\log_{10}(p)$. Thus, a value of $q=10$ is an error rate of 90%, a
value of $q=100$ is an error rate of 99%, and so on.

A generally useful cutoff is $q > 20$, based on the inflection point of error
associated with the PHRED score. This is discussed
[here](https://gatk.broadinstitute.org/hc/en-us/articles/360035531872-Phred-scaled-quality-scores).

In this case, all variants pass the Q20 cutoff. The variant caller likely has
a more stringent criteria.

```{r}
site.qual %>%
  as.data.frame() %>%
  ggplot(aes(x=QUAL)) +
  geom_histogram(fill="grey60", alpha=0.5, bins=100) +
  xlab("Site PHRED Score") + ylab("Frequency") +
  scale_x_log10() +
  ggplot.theme

site.qual %>%
  as.data.frame() %>%
  dplyr::mutate(Pass.Q20 = QUAL > 20) %>%
  dplyr::count(Pass.Q20) %>%
  dplyr::mutate(Total = sum(n)) %>%
  dplyr::mutate(Proportion = n / Total)
```

# Variant Filtering

```{r}
MERGED_DIR <- "/projects/motrpac/PRECOVID/DATA/GENOTYPE/MERGED/"
FILTERED_DIR <- "/projects/motrpac/PRECOVID/DATA/GENOTYPE/FILTERED/"
PLINK_DIR <- "/projects/motrpac/PRECOVID/DATA/GENOTYPE/PLINK/"

plot.data <- data.frame(
  Filters=c(
    "Detected Variants",
    "Missingness < 2%",
    "MAF > 2.5%",
    "HWE p-value ≥ 1E-6",
    "SNPs",
    "Autosomes"
  ),
  Variants=as.numeric(c(
    system(paste0("bcftools index -n ", MERGED_DIR, "/merged.annotated.vcf.gz"), intern=T),
    system(paste0("bcftools index -n ", FILTERED_DIR, "/merged.annotated.missingness_0_02.vcf.gz"), intern=T),
    system(paste0("bcftools index -n ", FILTERED_DIR, "/merged.annotated.missingness_0_02.maf_0_025.vcf.gz"), intern=T),
    system(paste0("bcftools index -n ", FILTERED_DIR, "/merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.vcf.gz"), intern=T),
    system(paste0("bcftools index -n ", FILTERED_DIR, "/merged.annotated.missingness_0_02.maf_0_025.hwe_1e-6.snps.vcf.gz"), intern=T),
    system(paste0("grep 'variants loaded from' ", PLINK_DIR, "/motrpac.precovid.n_238.maf_0_025.snps.log", " | cut -d ' ' -f 1"), intern=T)
  ))
)
```

```{r, fig.width=8, fig.height=6}
plot.data %>%
  dplyr::mutate(Filters=factor(Filters, levels=rev(Filters))) %>%
  ggplot() +
  geom_bar(aes(x=Filters, y=Variants), stat="identity", fill="grey60", alpha=0.5) +
  geom_text(aes(x=Filters, y=Variants, label=prettyNum(Variants, big.mark=",")), hjust=1.2, size=5) +
  coord_flip() +
  ggplot.theme +
  theme(axis.title.y=element_blank())

ggsave("results/motrpac_variant_filtering.svg", width=12, height=8)
```

