---
title: "SKM-atac-qc"
output: html_document
author: "Anne Marie Weitzel"
---

This analysis file uses a couple of functions that I created for other purposes in packages available on GitHub. Packages can be installed using the following commands: `pak::pak("weitzela/PeakGeneNet")`, `pak::pak("weitzela/bltools")`
```{r setup, include=FALSE}
pacman::p_load(tidyverse, ComplexHeatmap)
```

```{r}
source("fxns.R")

group_colors = c(MotrpacHumanPreSuspensionAnalysis::HUMAN_EXERCISE_GROUP_COLORS,
                 set_names(MotrpacHumanPreSuspensionAnalysis::HUMAN_EXERCISE_GROUP_COLORS, c("RE", "EE", "CON")))

saveFigures = function(file_prefix, plot_fxn, width, height) {
  bltools::pdfEmbed(file.path("tmp", "figures", paste0(file_prefix, ".pdf")), plot_fxn, width, height)
  grDevices::png(file.path("tmp", "figures", paste0(file_prefix, ".png")), width = width, height = height, units = "in", res = 300, family = "ArialMT"); print(plot_fxn); dev.off()
}

gg_add = list(
  scale_color_manual(values = group_colors, labels = prettyLabels, name = "Modality"),
  theme(
    legend.key.size = unit(2, "mm")
  )
)
```

```{r}
# prepare metadata
metadata = MotrpacHumanPreSuspensionData::load_pheno(load_acute_only = TRUE) |> 
  MotrpacHumanPreSuspensionData::attach_dictionary() |> 
  filter(visitcode == "ADULT Baseline Biospecimen Assessment Sequence (BAS)")

metadata_gcp = c("samples", "removed-samples") |> 
  purrr::set_names() |> 
  map(~ MotrpacBicQC::dl_read_gcp(paste0("gs://motrpac-data-hub/analysis/human-precovid-sed-adu/c2.0/epigenomics/metadata/human-precovid-sed-adu_t06-muscle_epigen-atac-seq_metadata_", .x, "_v2.0.txt"), tmpdir = "tmp") |> mutate(vialLabel = as.character(vialLabel)))

metadata = metadata |> 
  inner_join(metadata_gcp$samples, by = "vialLabel") |> 
  filter(!(vialLabel %in% metadata_gcp$`removed-samples`$vialLabel)) |> # drop samples swaps
  filter(n() > 1, .by = c("pid", "randomGroupCode")) |>  #  drop libraries where only one timepoint exists for the PID. they won't be used in any analysis so are removed))
  arrange(randomGroupCode, Sample_batch) |> 
  mutate(across(c(codedsiteid, Seq_flowcell_lane, Seq_flowcell_run, randomGroupCode), factor),
         Seq_batch_short = paste0(Seq_flowcell_lane, "__", Seq_flowcell_run) |> factor(),
         FRiP = peak_enrich.frac_reads_in_peaks.macs2.frip,
         across(c(Nuclei_extr_date, Tagmentation_date, PCR_date), ~ mdy(.x)),
         nucleiIsolation2Tagmentation = Tagmentation_date - Nuclei_extr_date, 
         Tagmentation2PCR = PCR_date - Tagmentation_date,
         nuclei2atac_lag = nucleiIsolation2Tagmentation > 0,
         library = vialLabel) |> 
  relocate(vialLabel, pid, randomGroupCode, Sex, Sample_batch, Seq_batch_short, nucleiIsolation2Tagmentation, Tagmentation2PCR, nuclei2atac_lag) |> 
  group_by(randomGroupCode, pid) |> 
  arrange(randomGroupCode, min(Sample_batch)) |> 
  ungroup() |> 
  dplyr::rename("nuclei_isolation_batch" = "Sample_batch") |> 
  droplevels()

tmp_binarized = metadata |> 
  (\(df) {
    binarized_df = df |> 
      (\(tmp_df) {
          map(c("randomGroupCode", "Timepoint", "codedsiteid", "Seq_batch_short"), function(tmp_var) binarizeVar(tmp_df[[tmp_var]], tmp_var)) |> 
            bind_cols() 
        })() 
    numeric_df = df |> 
      select(Sex, nuclei2atac_lag) |> 
      mutate(across(c(Sex, nuclei2atac_lag), as.numeric)) 
    bind_cols(numeric_df, binarized_df) |> 
      rename_with(function(.c) paste0("binarized___", .c))
  })()
metadata = bind_cols(metadata, tmp_binarized)

# prepare count matrix, voom normalized
features_to_keep = MotrpacBicQC::dl_read_gcp("gs://motrpac-data-hub/analysis/human-precovid-sed-adu/c2.0/epigenomics/qc-norm/human-precovid-sed-adu_t06-muscle_epigen-atac-seq_qc-norm_log-cpm_v2.0.txt", tmpdir = "tmp") |> 
  mutate(feature_id = str_replace_all(feature_id, "-", ":")) |> pull(feature_id)

voom = MotrpacBicQC::dl_read_gcp("gs://motrpac-data-hub/quant-id/human-precovid/c1.0/epigenomics/t06-muscle/epigen-atac-seq/motrpac_human-precovid_t06-muscle_epigen-atac-seq_counts_v1.0.txt.gz", tmpdir = "tmp") |> 
  unite(feature_id, chrom:end, sep = ":") |> 
  filter(feature_id %in% features_to_keep) |> 
  select(feature_id, all_of(metadata$vialLabel)) |> 
  column_to_rownames(var = "feature_id") |> 
  as.matrix() |> 
  (\(.mat) limma::voom(.mat[,metadata$vialLabel])$E)()

identical(colnames(voom), as.character(metadata$vialLabel))

rm(metadata_gcp, tmp_binarized, features_to_keep); invisible(gc())
```

```{r fig.width=20}
# did this to see what qc metric columns to select. i hand picked terms to represent a highly correlated cluster 

# metadata |> select(starts_with(c("align", "peak", "replication")) & where(is.numeric)) |>
#   select(-c(contains("syn_x_intercept"))) |>
#   select(!where(~ n_distinct(.x) == 1)) |>
#   as.matrix() |>
#   Hmisc::rcorr() |>
#   (\(mat) bltools::chHeatmap(mat[[1]], column_km = 7, row_km = 7))() #cluster_columns = F, cluster_rows = F
```

```{r}
selected_metadata_columns = list(
  "Biological" = data.frame(col_name = c("randomGroupCode", "Timepoint", "Sex", "calculatedAge", "BMI")),
  # "GroupTimepoint" = data.frame(col_name = "group_timepoint"),
  "Experimental" = data.frame(col_name = c("codedsiteid", "Seq_batch_short", "nuclei_isolation_batch", "nuclei2atac_lag")),
  "QC Metrics" = data.frame(col_name = intersect(colnames(metadata), c("FRiP", "align.samstat.total_reads", "align.dup.pct_duplicate_reads", "align.nodup_samstat.total_reads", "align_enrich.tss_enrich.tss_enrich", "align.frag_len_stat.frac_reads_in_nfr", "peak_stat.peak_region_size.idr_opt.mean", "align.samstat.pct_properly_paired_reads", "peak_enrich.frac_reads_in_annot.fri_blacklist", "replication.num_peaks.num_peaks", "align.dup.paired_optical_duplicate_reads", "align.frac_mito.frac_mito_reads", "peak_stat.peak_region_size.mean", "peak_stat.peak_region_size.overlap_opt.mean", "peak_stat.peak_region_size.max_size")))
) |> 
  bind_rows(.id = "category")
selected_metadata_columns = data.frame(full_col_name = grep("binarized___", colnames(metadata), value = TRUE)) |> 
  separate_wider_delim(full_col_name, "___", names = c("col_type", "col_name", "short_name"), too_few = "align_start", cols_remove = FALSE) |> 
  mutate(short_name = ifelse(col_name %in% c("codedsiteid", "Seq_batch_short"), paste0(col_name, "_", short_name), short_name)) |> 
  (\(df) left_join(selected_metadata_columns, df, by = "col_name"))() |> 
  mutate(col_type = ifelse(is.na(col_type), "numeric", col_type),
         short_name = ifelse(is.na(short_name), col_name, short_name),
         full_col_name = ifelse(is.na(full_col_name), col_name, full_col_name))
```

# General Figures
## Experimental Setup
```{r}
heatmap_legend_params = list(title_gp = gpar(fontsize = 7, fontface = "bold"), labels_gp = gpar(fontsize = 6), grid_width = unit(0.3, "cm"), grid_height = unit(0.3, "cm"))

hm_ls = list()
hm_ls$prep_by_pid = metadata |> 
  (\(tmp_df) {
    tmp_df = tmp_df |> 
      select(pid, randomGroupCode, Sex, Timepoint, codedsiteid, nuclei_isolation_batch) |> 
      pivot_wider(id_cols = pid:Sex, names_from = "Timepoint", values_from = "nuclei_isolation_batch", names_sort = TRUE)
    mat = tmp_df |> 
      column_to_rownames(var = "pid") |> 
      select(pre_exercise:last_col()) |> 
      as.matrix()
    hm = mat |> 
      Heatmap(
        na_col = "white", row_split = tmp_df$randomGroupCode, cluster_rows = FALSE, cluster_columns = FALSE,
        col = rev(rainbow(10)), column_labels = prettyLabels(colnames(mat)),
        column_title = "Nuclei\nIsolation\nBatch", column_title_gp = gpar(fontsize = 7),
        row_names_gp = grid::gpar(fontsize = 6),
        column_names_gp = grid::gpar(fontsize = 6), 
        row_title_gp = grid::gpar(fontsize = 7), 
        heatmap_legend_param = c(list(title = "Nuclei Isolation\nBatch (ordered)", 
                                      at = c(17, 27, 36, 53),
                                      # at = c(1, 2.5, 4.5, 6)*10, 
                                      direction = "horizontal", legend_width = unit(1.75, "cm"), grid_height = unit(0.25, "cm")), heatmap_legend_params |> discard_at(c("grid_width", "grid_height"))),
        height = nrow(tmp_df) * grid::unit(2.5, "mm"),
        width = 4 * grid::unit(2.5, "mm"), 
        left_annotation = rowAnnotation(df = tmp_df |> select(Group = randomGroupCode, Sex), col = list(Sex = MotrpacHumanPreSuspensionAnalysis::HUMAN_SEX_COLORS, Group = MotrpacHumanPreSuspensionAnalysis::HUMAN_EXERCISE_GROUP_COLORS), simple_anno_size = unit(3, "mm"), annotation_name_gp = gpar(fontsize = 6), annotation_legend_param = c(list(), heatmap_legend_params))
      )
    return(hm)
  })()

hm_ls$time_by_pid = metadata |> 
  select(pid, randomGroupCode, Sex, Timepoint, codedsiteid, nucleiIsolation2Tagmentation) |>
  mutate(nucleiIsolation2Tagmentation = as.numeric(nucleiIsolation2Tagmentation)) |>
  pivot_wider(id_cols = pid:Sex, names_from = "Timepoint", values_from = "nucleiIsolation2Tagmentation", names_sort = TRUE) |> 
  (\(tmp_df) {
    mat = tmp_df |>
      column_to_rownames(var = "pid") |>
      select(pre_exercise:last_col()) |>
      as.matrix()
    hm = mat |> 
      Heatmap(
        na_col = "white", row_split = tmp_df$randomGroupCode, cluster_rows = FALSE, cluster_columns = FALSE,
        col = circlize::colorRamp2(c(0, 1), c("#CBFBF1", "#36BBA7")), column_labels = prettyLabels(colnames(mat)),
        column_title = "ATAC From\nFrozen Nuclei", column_title_gp = gpar(fontsize = 7),
        row_names_gp = grid::gpar(fontsize = 6),
        column_names_gp = grid::gpar(fontsize = 6),
        row_title_gp = grid::gpar(fontsize = 7),
        heatmap_legend_param = c(list(title = "Days Between\nNuclei Isolation\n& Tagmentation", at = c(1, 0), labels = c("≥1", "0"), color_bar = "discrete"), heatmap_legend_params),
        height = nrow(tmp_df) * grid::unit(2.5, "mm"),
        width = 4 * grid::unit(3.5, "mm"), 
        layer_fun = function(j, i, x, y, width, height, fill) {
          v_P = ComplexHeatmap::pindex(mat, i, j)
          c_1 = (!is.na(v_P)) & (v_P > 0)
          grid.text(sprintf("%d", v_P[c_1]), x[c_1], y[c_1], gp = gpar(fontsize = 5))
        }
      )
    return(hm)
  })()

hm_ls$flowcells_by_pid = metadata |> 
  (\(tmp_df) {
    tmp_df = tmp_df |> 
      select(pid, randomGroupCode, Sex, Timepoint, codedsiteid, Seq_batch_short) |> 
      pivot_wider(id_cols = pid:Sex, names_from = "Timepoint", values_from = "Seq_batch_short", names_sort = TRUE)
    mat = tmp_df |> 
      column_to_rownames(var = "pid") |> 
      select(pre_exercise:last_col()) |> 
      as.matrix()
    hm = mat |> 
      Heatmap(
        na_col = "white", row_split = tmp_df$randomGroupCode, column_labels = prettyLabels(colnames(mat)),
        col = tmp_df |> select(pre_exercise:last_col()) |> unlist() |> na.omit() |> unique() |> (\(x) set_names(aesthetically_pleasing_colors$colorBlindFriendly2[1:length(x)], x))(),
        column_title = "Sequence\nBatch", column_title_gp = gpar(fontsize = 7),
        row_names_gp = grid::gpar(fontsize = 6),
        column_names_gp = grid::gpar(fontsize = 6), 
        row_title_gp = grid::gpar(fontsize = 7), 
        heatmap_legend_param = c(list(title = "Sequence\nBatch"), heatmap_legend_params),
        height = nrow(tmp_df) * grid::unit(2.5, "mm"),
        width = 4 * grid::unit(2.5, "mm")
      )
    return(hm)
  })()

hm_full = draw(reduce(hm_ls, `+`), heatmap_legend_side = "bottom", annotation_legend_side = "bottom", legend_grid_width = unit(1, "cm"), legend_grouping = "original")
saveFigures("metadata-by-pid_hm", plot(hm_full), 2.8, 6)
```

## FRiP
```{r}
p_frip = ggplot(metadata, aes(x = Timepoint, y = FRiP, color = randomGroupCode)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_wrap(~ randomGroupCode) + 
  xlab(NULL) + 
  scale_x_discrete(labels = ~ prettyLabels(.x, T)) +
  geom_boxplot(outliers = FALSE, show.legend = FALSE) +
  geom_jitter(aes(color = randomGroupCode, shape = Timepoint), size = 1.5) + 
  gg_add + 
  scale_shape_manual(values = c(15, 17, 18, 16)) + 
  theme(legend.position = "none")
p_frip
saveFigures("boxplot_frip-by-group", plot(p_frip), 3, 2)
```

# PCA  
Data adjustment uses `PeakGeneNet::adjustCovariateMatrix` that can be installed using, `pak::pak("weitzela/PeakGeneNet")`.
```{r}
original_vars = c("BMI", "Sex", "calculatedAge", "codedsiteid", "Timepoint", "randomGroupCode")

protect_vars = list("exercise" = c("randomGroupCode", "Timepoint"), "sex" = "Sex")

model_inclusion = list(
  "no_adj" = "",
  "~M0" = c("FRiP", original_vars)
)

filter_fxns = c(
  all = function(df) df,
  map(levels(metadata$randomGroupCode) |> set_names(prettyLabels), 
      function(.x) function(df) df |> filter(randomGroupCode == .x))
)
```

```{r}
pca_ls = imap(filter_fxns, function(.filter_fxn, .filter_label) {
  map(protect_vars, function(.protect_vars) {
    map(model_inclusion, function(.additional_vars) {
      try(pcaWrapper(voom, metadata, .filter_fxn, .additional_vars, .protect_vars), silent = TRUE)
    })
  })
})

pca_ls = pca_ls |> 
  map_depth(2, function(.ls) discard(.ls, ~ class(.x) == "try-error"))
```

## Figures
```{r, fig.width=8, fig.height=8, dpi=300}
# heatmap
hm_ls_split = imap(pca_ls[1], function(.sample_ls, .sample_inclusion) {
  imap(.sample_ls, function(.protect_ls, .protect_var) {
    imap(c(list("all" = c("ADUControl", "ADUEndur", "ADUResist")), list("ADUControl", "ADUEndur", "ADUResist") |> purrr::set_names()) |> keep_at(ifelse(.sample_inclusion == "all", TRUE, set_names(levels(metadata$randomGroupCode), prettyLabels)[.sample_inclusion])), function(.grp, .grp_filter) {
      pca_res_wide = .protect_ls |> 
        imap(function(.pca, .model) {
          .pca$x |> 
            filter(randomGroupCode %in% .grp) |> 
            select(library, set_names(names(.pca$variance), .pca$variance)[1:5]) |> 
            rename_with(function(.c) paste0(.c, "___PC ", .model), -library)
        }) |> 
        reduce(full_join, by = "library")
      

      binarized_tmp = metadata |> 
        select(library, all_of(selected_metadata_columns$full_col_name)) |> 
        inner_join(pca_res_wide, by = "library") |> 
        select(-library)
      
      split_tmp = selected_metadata_columns |> 
        right_join(data.frame(full_col_name = colnames(binarized_tmp)), by = "full_col_name") |>
        arrange(match(full_col_name, colnames(binarized_tmp))) |> 
        mutate(short_name = ifelse(is.na(category), str_remove_all(full_col_name, "___.*$"), short_name),
               category = ifelse(is.na(category), str_remove_all(full_col_name, "^.*___") |> str_replace(" ", "\n"), category) |> str_replace_all(c("\\+" = "\n\\+", "PC\n" = "", "no_adj" = "Voom\nCounts"))) |> 
        mutate(category = fct_inorder(category))
      
      row_vars = intersect(colnames(binarized_tmp), selected_metadata_columns$full_col_name)
      col_vars = setdiff(colnames(binarized_tmp), selected_metadata_columns$full_col_name)
      
      tmp_df = split_tmp |> 
        filter(full_col_name %in% col_vars) |> 
        mutate(correlation_inclusion = ifelse(.grp_filter == "all", "All Samples", "Within Group"),
               samples = ifelse(.grp_filter == "all", "All", .grp),
               adjustment = category |> str_replace_all("\n\\+", "\\+") |> str_replace_all("\n", " ") |> str_remove_all("PC\n")) |> 
        select(`Correlation Inclusion` = correlation_inclusion, samples, adjustment, full_col_name) |> 
        column_to_rownames(var = "full_col_name")
      if (.sample_inclusion != "all") tmp_df = tmp_df |> select(-`Correlation Inclusion`)
      
      hm = binarized_tmp |> 
        as.matrix() |> 
        Hmisc::rcorr() |> 
        discard_at("type") |> 
        map(~ .x[row_vars, col_vars]) |>
        (\(mat) bltools::chHeatmap(mat[[1]], name = "r", na_col = "gray90",
                                   col_and_row_name_size = 6, cell_mm = 2.5, heatmap_legend_param = list(title_gp = gpar(fontsize = 7, fontface = "bold"), labels_gp = gpar(fontsize = 6), grid_width = unit(0.3, "cm")),
                                   column_title_gp = gpar(fontsize = 8), row_title_gp = gpar(fontsize = 8),
                                   column_title_side = "bottom",
                                   column_split = pull(split_tmp, category, full_col_name)[colnames(mat[[1]])],
                                   row_split = pull(split_tmp, category, full_col_name)[rownames(mat[[1]])],
                                   column_labels = pull(split_tmp, short_name, full_col_name)[colnames(mat[[1]])],
                                   row_labels = pull(split_tmp, short_name, full_col_name)[rownames(mat[[1]])],
                                   cluster_row_slices = FALSE, cluster_column_slices = FALSE,
                                   cluster_rows = FALSE, cluster_columns = FALSE,
                                   cell_fun = bltools::hmSigDots(mat$P, threshold = 0.05, fun = "cell", dot_pt_size = 2, label_mat = round(mat$r, 1), fontsize = 3),
                                   row_names_max_width = unit(2, "cm"),
                                   top_annotation = columnAnnotation(
                                     df = tmp_df,
                                     col = list(adjustment = c("#E78AC3", "#A6D854", "#FFD92F", "#6699CC")[1:n_distinct(tmp_df$adjustment)] |> set_names(fct_inorder(unique(tmp_df$adjustment))), `Correlation Inclusion` = c("#999933", "#882255") |> set_names(c("All Samples", "Within Group")), samples = c("All" = "#E5C494", MotrpacHumanPreSuspensionAnalysis::HUMAN_EXERCISE_GROUP_COLORS)),
                                     annotation_name_gp = gpar(fontsize = 6),
                                     annotation_legend_param = list(title_gp = gpar(fontsize = 7, fontface = "bold"), labels_gp = gpar(fontsize = 6), grid_width = unit(0.3, "cm"), grid_height = unit(0.3, "cm"), nrow = 2),
                                     simple_anno_size = unit(3, "mm"),
                                     show_annotation_name = .grp_filter == "ADUResist"
                                   )
                                   ) |> `attr<-`("cor_res", mat))()
      return(hm)
    })
  })
})

hm_split_all_only = map(hm_ls_split$all, ~ reduce(.x[1], `+`) |> draw(heatmap_legend_side = "right", annotation_legend_side = "top", align_heatmap_legend = "heatmap_top"))
saveFigures("hm_filtered_all-sample-pca_protect-exercise", draw(hm_split_all_only$exercise), 3.5, 6.5)
```

```{r, fig.width=3.2, fig.height=2, dpi=300}
pc1_nuclei_isolation_batch = pca_ls$all$exercise$`~M0`$x |> 
  mutate(Timepoint = fct_relabel(Timepoint, function(.label) prettyLabels(.label, TRUE))) |> 
  ggplot(aes(x = factor(nuclei_isolation_batch), y = PC1)) + ylab(paste0("~M0 ", pca_ls$all$exercise$`~M0`$variance[1])) + xlab("Nuclei Isolation Batch") +
  geom_point(aes(color = randomGroupCode, shape = Timepoint), size = 2, alpha = 0.5, stroke = 0) +
  guides(color = guide_legend(order = 1, override.aes = list(alpha = 1)),
         shape = guide_legend(override.aes = list(alpha = 1))) + 
  gg_add + 
  scale_shape_manual(values = c(15, 17, 18, 16)) +
  theme(legend.position = "none") +
  geom_smooth(method = "lm", se = FALSE, aes(x = as.numeric(factor(nuclei_isolation_batch))), color = "#E5C494",  data = ~ .x |> filter(nuclei_isolation_batch < 50)) +
  geom_smooth(method = "lm", se = FALSE, aes(x = as.numeric(factor(nuclei_isolation_batch)), color = randomGroupCode), data = ~ .x |> filter(nuclei_isolation_batch < 50)) + 
  ggpubr::stat_cor(method = "spearman", aes(x = as.numeric(factor(nuclei_isolation_batch))), cor.coef.name = "rho", color = "#E5C494", label.y.npc = 0.27, label.x.npc = 0.72, p.digits = 2, p.accuracy = 1e-10) +
  ggpubr::stat_cor(method = "spearman", aes(x = as.numeric(factor(nuclei_isolation_batch)), color = randomGroupCode), cor.coef.name = "rho", label.y.npc = 0.2, label.x.npc = 0.72, p.accuracy = 0.001); pc1_nuclei_isolation_batch
saveFigures("pc1-vs-nuclei-isolation-batch", plot(pc1_nuclei_isolation_batch), 3, 2)
```

```{r}
pc2_frozen_nuclei = pca_ls$all$exercise$`~M0`$x |> 
  ggplot(aes(x = randomGroupCode, y = PC2, color = nuclei2atac_lag, shape = Timepoint)) |> 
  (\(.p) list(
    .p + geom_boxplot(outliers = FALSE, show.legend = FALSE, aes(shape = NULL), linewidth = 0.25),
    .p + geom_boxplot(outliers = FALSE, show.legend = FALSE, linewidth = 0.25)
  ))() |> 
  map(function(.p) {
    .p + 
      geom_point(position = position_jitterdodge(), size = 1.25, stroke = 0) + 
      guides(color = guide_legend(title = "Use of Frozen Nuclei", override.aes = list(size = 2)),
             shape = "none") +
      theme(legend.position = "bottom", legend.margin = margin(), 
            legend.key.width = unit(2, "mm"), legend.key.height = unit(1, "mm")) +
      labs(x = NULL, y = paste0("~M0 ", pca_ls$all$exercise$`~M0`$variance[2])) + 
      scale_shape_manual(values = c(15, 17, 18, 16)) + 
      scale_x_discrete(labels = prettyLabels)
  }) |> 
  ggpubr::ggarrange(plotlist = _, nrow = 1, common.legend = TRUE, widths = c(0.7, 1), legend = "bottom")
saveFigures("pc2-vs-frozen-nuclei", plot(pc2_frozen_nuclei), 2.8, 2)
```

```{r, fig.height=2, fig.asp=1}
p_pca_grid1 = imap(pca_ls$all$exercise[2], function(.x, .additional) {
  model_anno = paste0("In Model: ", paste0(attr(.x, "in_model"), collapse = ", ") |> str_wrap(45), "\n",
                      "Protected: ", paste0(attr(.x, "protect"), collapse = ", "))
  .x$x = .x$x |> 
    mutate(Timepoint = fct_relabel(Timepoint, function(.label) prettyLabels(.label, TRUE)))
  p = ggplot(.x$x, aes(x = PC1, y = PC2, color = randomGroupCode,shape = Timepoint, linetype = Timepoint, linewidth = Timepoint))
  p = p +
    geom_point(size = 1.5, stroke = 0, alpha = 0.8) + 
    stat_ellipse(alpha = 0.8) +
    theme(aspect.ratio = 1) +
    theme(legend.direction = "horizontal", legend.position = "bottom") +
    guides(color = guide_legend(override.aes = list(linetype = 0), order = 1), 
           shape = guide_legend(override.aes = list(size = 1))) +
    labs(x = paste0("~M0 ", .x$variance[["PC1"]]), y = paste0("~M0 ", .x$variance[["PC2"]]), caption = model_anno) +
    scale_linewidth_manual(values = c(0.5, 0.5, 0.15, 0.15)) +
    scale_linetype_manual(values = c("solid", "dotted", "dotted", "solid")) +
    gg_add + 
    theme(legend.key.width = unit(3, "mm")) +
    scale_shape_manual(values = c(15, 17, 18, 16))
  
  p = p |> 
    `attr<-`("legend", ggpubr::get_legend(p))
  p = p +
    theme(legend.position = "none")
  return(p)
}) 

m0_pca = p_pca_grid1$`~M0` + labs(caption = NULL)
saveFigures("m0-pca-plot", plot(m0_pca), 2, 2)
```
```{r}
p_legend = (p_pca_grid1$`~M0` + 
  theme(legend.position = "right") +
  guides(linetype = guide_legend(order = 1, override.aes = list(alpha = 1)),
         linewidth = guide_legend(order = 1),
         shape = guide_legend(order = 2, override.aes = list(size = 2)),
         color = guide_legend(order = 3, override.aes = list(linetype = 0, size = 2, alpha = 1)))) |> 
  ggpubr::get_legend() 
saveFigures("legend", plot(p_legend), 2.5, 0.5)

grDevices::png(file.path("tmp", "figures", paste0("legend", ".png")), width = 2.5, height = 0.5, units = "in", res = 300, family = "ArialMT"); print(plot(p_legend)); dev.off()
```

# Numbers
```{r}
metadata |> 
  group_by(randomGroupCode, nuclei2atac_lag) |> tally() |> pivot_wider(names_from = nuclei2atac_lag, values_from = n) |> mutate(sum = `FALSE` + `TRUE`)

metadata |> 
  group_by(pid) |> 
  mutate(any_frozen = any(nuclei2atac_lag)) |> 
  distinct(pid, any_frozen, randomGroupCode) |> 
  group_by(randomGroupCode, any_frozen) |> 
  tally() |> 
  pivot_wider(names_from = any_frozen, values_from = n) |> mutate(pct = scales::label_percent()(`TRUE` / (`FALSE` + `TRUE`)))

metadata |> 
  group_by(pid) |> 
  mutate(multi_seq = n_distinct(Seq_batch_short) > 1, .before = 1) |> 
  distinct(pid, multi_seq, randomGroupCode) |> 
  group_by(randomGroupCode, multi_seq) |> 
  tally() |> 
  pivot_wider(names_from = multi_seq, values_from = n) |> mutate(pct = scales::label_percent()(`TRUE` / (`FALSE` + `TRUE`)))
```

## Peak Counts
```{r}
og_res = MotrpacBicQC::dl_read_gcp("gs://motrpac-data-hub/analysis/human-precovid-sed-adu/c2.0/epigenomics/da/human-precovid-sed-adu_t06-muscle_epigen-atac-seq_da_dream-acute_v2.1.txt", tmpdir = "tmp")
# og_res = MotrpacHumanPreSuspensionAnalysis::load_differential_analysis("epigen-atac-seq", "muscle", epigen = TRUE, repo_local_dir = "tmp")

results_category = og_res |> 
  # filter(contrast_type != "Endur_vs_Resist") |> 
  filter(adj_p_value < 0.05) |> 
  distinct(contrast) |> 
  mutate(contrast_tmp = str_remove_all(contrast, "group_timepoint")) |> 
  separate_wider_delim(contrast_tmp, delim = regex(" -|\\+ "), names = c("ex_post", "ex_pre", "con_post", "con_pre"), too_few = "align_start", cols_remove = FALSE) |> 
  mutate(randomGroupCode = str_remove_all(ex_post, "\\..*$"),
         Timepoint = str_remove_all(ex_post, "^.*\\.(?=post)"),
         model_type = paste0(randomGroupCode, "-", ifelse(is.na(con_post), randomGroupCode, str_remove_all(con_post, "\\..*$"))) |> str_remove_all(" ") |> prettyLabels(), .before = 1) |> 
  filter(model_type != "EE-RE") |> 
  mutate(contrast_type = ifelse(str_detect(model_type, "-CON"), "exercise_with_controls", "exercise_no_controls"), .before = 1) |> 
  mutate(contrast_type = fct_rev(contrast_type),
         randomGroupCode = factor(randomGroupCode),
         Timepoint = fct_relevel(Timepoint, "post_24_hr", after = 3)) |> 
  select(-c(ex_post, ex_pre, con_post, con_pre))

peak_compare = results_category |> 
  inner_join(og_res |> filter(adj_p_value < 0.05), by = "contrast") |> 
  # droplevels() |> 
  group_by(contrast_type, Timepoint, randomGroupCode) |> 
  summarise(peaks = list(feature_id)) |> 
  pivot_wider(names_from = "contrast_type", values_from = "peaks", values_fill = list()) |> 
  mutate(dd_only = map2_dbl(exercise_with_controls, exercise_no_controls, ~ length(setdiff(.x, .y))), 
         common = map2_dbl(exercise_with_controls, exercise_no_controls, ~ length(intersect(.x, .y))), 
         within_group_only = map2_dbl(exercise_with_controls, exercise_no_controls, ~ length(setdiff(.y, .x))), .after = 2) |> 
  bind_rows(data.frame(Timepoint = factor("post_24_hr"), randomGroupCode = "ADUEndur", dd_only = 0, common = 0, within_group_only = 0)) |> 
  arrange(randomGroupCode, Timepoint) 

p_count_compare2 = peak_compare |> 
  select(-where(is.list)) |> 
  pivot_longer(-c(Timepoint, randomGroupCode), names_to = "category", values_to = "n") |> 
  mutate(x_val = c("dd_only" = 0.125, "common" = 0.5, "within_group_only" = 0.875)[category]) |> 
  mutate(fill_val = ifelse(x_val == 0.5, 0.5, 0.25)) |> 
  mutate(category = fct_inorder(category)) |> 
  ggplot(aes(x = x_val, y = fct_rev(Timepoint), label = n, fill = category)) + 
  geom_tile(color = "white", linewidth = 1) +
  # geom_col(aes(x = fill_val), position = position_stack(reverse = TRUE)) +
  geom_text(fontface = "bold") + 
  facet_wrap(~ randomGroupCode) + 
  labs(y = NULL, x = NULL) + 
  scale_y_discrete(labels = ~ prettyLabels(.x, TRUE), expand = expansion(0)) + 
  scale_x_continuous(breaks = c(0.125, 0.5, 0.875), labels = c("difference-in-\nchanges only", "common", "within-group\nonly"), expand = expansion(0)) +
  scale_fill_manual(values = aesthetically_pleasing_colors$colorBlindFriendly2[-c(1:3)]) +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + 
  ggtitle("Distinct & Overlapping Significant Peaks")

p_count_compare1 = results_category |> 
  inner_join(og_res |> filter(adj_p_value < 0.05), by = "contrast") |> 
  group_by(contrast_type, Timepoint, randomGroupCode) |> tally() |> 
  ungroup() |> 
  bind_rows(data.frame(contrast_type = c(rep("exercise_with_controls", 2), "exercise_no_controls"), Timepoint = factor("post_24_hr"), randomGroupCode = c("ADUResist", rep("ADUEndur", 2)), n = 0)) |> 
  mutate(contrast_type = fct_inorder(contrast_type)) |> 
  ggplot(aes(y = fct_rev(Timepoint), x = contrast_type, fill = n, label = n)) +
  geom_tile(color = "white", linewidth = 1) + 
  geom_text(fontface = "bold") + 
  facet_wrap(~ randomGroupCode) + 
  # coord_fixed() + 
  scale_fill_continuous(palette = c("#EFEDF5", "#BCBDDC", "#756BB1")) +
  scale_x_discrete(labels = c("difference-in-\nchanges", "within-\ngroup"), expand = expansion(0)) + 
  scale_y_discrete(labels = ~ prettyLabels(.x, TRUE), expand = expansion(0)) + 
  labs(y = NULL, x = NULL) + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + 
  ggtitle("Count of Significant Peaks")

p_count_compare_grid = ggpubr::ggarrange(p_count_compare1, p_count_compare2, nrow = 1, align = "h", widths = c(0.6, 1))
saveFigures("peak-count-plots", plot(p_count_compare_grid), 6, 1.3)

# percent corroboration
sum(peak_compare$common) / (sum(peak_compare$common) + sum(peak_compare$dd_only))
```
