---
title: "pca_variance_exp_plots"
output: html_document
date: "2025-07-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacHumanPreSuspensionAnalysis)
library(MotrpacHumanPreSuspensionData)
library(dplyr)
library(tidyr)
library(here)
library(tibble)
library(purrr)
library(variancePartition)
library(BiocParallel)
library(Rtsne)

figure_out_path = file.path(here(), "data", "tmp", "figures")
```

# Note: 

This figure(s) requires access to `MotrpacHumanPreSuspensionData` to fully recreate. 

# Figure Info
This code is used to generate: 
Figure S1C,D
Table S3. 
Supplemental figures for some other purpose. 

```{r}
non_methyl = ome_available_list()
non_methyl = non_methyl[non_methyl != "epigen-methylcap-seq"]

qc_norm_list = load_qc(selected_omes = non_methyl,
                       epigen = TRUE,
                       repo_local_dir = file.path(here(), "data", "tmp"))
```

First - PCAs, CCAs, TSNE

So for these figures, we are trying to collapse the sets of metabolomics platforms
and just make one figure/panel block for metabolomics platforms intead of having
one PCA/CCA/TSNE per metab. 
```{r make metab_qc_norm}
for(tissue in names(qc_norm_list)){
  #so we re-load qc_norm metab platforms to use `combine_qc_matrixes`
  #and then we're going to plug in this combined one in place of all the single matrixes
  metab_platforms_only = load_qc(selected_tissues = tissue,
                                 selected_omes = metab_only_list())
  
  combined_metab_matrixes = combine_qc_matrixes(metab_platforms_only)
  #we select any of the metadata, because the resulting combined matrix should have
  #only shared participants, so the metadata i need should be in every sample_meta
  metab_single_metadata = metab_platforms_only[[tissue]][[1]][["sample_metadata"]]
  #now we convert back from the format of `combine_qc_matrix` to map back to vialLabels
  #in reality ea column of the metab consists of mixes of metab, but for the figures
  #we're making here, it shouldn't need any of the other metadata.
  name_mapping = metab_single_metadata %>%
    mutate(pid_timepoint = interaction(pid, Timepoint, sep = "..")) %>%
    filter(pid_timepoint %in% colnames(combined_metab_matrixes)) %>%
    dplyr::select(pid_timepoint, vialLabel) %>%
    deframe()
  colnames(combined_metab_matrixes) = name_mapping[colnames(combined_metab_matrixes)]
  
  combined_metab_replacement = list("qc_norm" = combined_metab_matrixes,
                                    "sample_metadata" = metab_single_metadata)
  
  #and just get rid of the individual platforms
  qc_norm_list[[tissue]][grep("metab", names(qc_norm_list[[tissue]]))] = NULL
  qc_norm_list[[tissue]][["metab"]] = combined_metab_replacement
}
```

```{r}
cca_plot_list = list()
for(tissue in names(qc_norm_list)){
  for(assay in names(qc_norm_list[[tissue]])){
    data_to_pca = qc_norm_list[[tissue]][[assay]][["qc_norm"]]
    if(assay == "prot-ph" || assay == "prot-pr")  
      data_to_pca = data_to_pca %>% tidyr::drop_na()
    
    if(nrow(data_to_pca) < 10) next
    sample_metadata = qc_norm_list[[tissue]][[assay]][["sample_metadata"]]
    rownames(sample_metadata) = sample_metadata$vialLabel
    pdf(file = file.path(figure_out_path, "pcas", paste0(tissue,"_", assay, "_pca.pdf")))
    p = plot_precovid_pca(data_to_pca,
                          sample_metadata,
                          custom_title = paste(tissue, assay))
    print(p)
    dev.off()
    
    #-----new: cca plot
    pdf(file = file.path(figure_out_path, "ccas", paste0(tissue,"_", assay, "_cca.pdf")))
    cca_plot = plot_precovid_cca(data_to_pca,
                                 sample_metadata,
                                 custom_title = paste(tissue, assay))
    cca_plot_list[[tissue]][[assay]] = cca_plot
    dev.off()
    
    #------new: tSNE plots
    if(nrow(data_to_pca) < 100 | ncol(data_to_pca) < 50) next
    tSNE_counts = Rtsne(t(data_to_pca), 
                        check_duplicates = TRUE, 
                        num_threads = 4)
    unique_pids = unique(sample_metadata$pid)
    pid_colors = hsv(seq(0, 1, length.out = length(unique_pids)), v = 1, s = 0.8)
    pid_colors = pid_colors[as.numeric(factor(sample_metadata$pid))]
    
    pdf(file = file.path(figure_out_path, "tsnes", paste0(tissue,"_", assay, "_tsnes.pdf")))
    plot(tSNE_counts$Y, 
         col = pid_colors[match(colnames(data_to_pca), sample_metadata$vialLabel)], 
         pch = 20, 
         xlab = "Each color represents one participant", 
         ylab = "", 
         xaxt = "n", 
         yaxt = "n", 
         main = paste(assay, "tSNE", tissue))
    dev.off()
  }
}
```


```{r}
library(gridExtra)
#Note: we also are modifying this to make each tissue a column for the grid 
grobs_by_tissue = list()
for(tissue in names(cca_plot_list)){
  sublist = list()
  for(assay in names(cca_plot_list[[tissue]])){
    pheat = cca_plot_list[[tissue]][[assay]]
    if (inherits(pheat, "pheatmap")) {
      sublist[[assay]] = pheat$gtable
    }
  }
  if(length(sublist) > 0){
    grobs_by_tissue[[tissue]] = sublist
  }
}

# Convert each tissue's set of grobs into a vertically stacked gTree
tissue_panels = lapply(grobs_by_tissue, function(grob_list) {
  do.call(arrangeGrob, c(grob_list, ncol = 1))
})

# Arrange the grobs in a grid. Make sure you either combine metab or 
# do something else because the figure looks terrible with all the metabs separate
# because there's basically infinite rows

pdf(file = file.path(figure_out_path,"ccas", "all_tissue_ome_cca.pdf"),
    width = 15, height = 15)
do.call(grid.arrange, c(tissue_panels, ncol = length(tissue_panels)))
dev.off()



```

next we want to work on variance explained plots

note: actually fitting the models
```{r, eval = TRUE}
var_part_results = list()

for(tissue in names(qc_norm_list)){
  for(assay in names(qc_norm_list[[tissue]])){
    message(paste(tissue, assay))
    qc_norm = qc_norm_list[[tissue]][[assay]][["qc_norm"]] %>% drop_na()
    if(nrow(qc_norm) < 10) next
    sample_metadata = qc_norm_list[[tissue]][[assay]][["sample_metadata"]]
    rownames(sample_metadata) = sample_metadata$vialLabel 
    sample_metadata$group_timepoint = interaction(sample_metadata$randomGroupCode, sample_metadata$Timepoint)
    sample_metadata = sample_metadata[match(colnames(qc_norm), rownames(sample_metadata)), ] 
    #reorder so they're in the same order as colnames of raw_counts
    
    # param = SnowParam(8, "SOCK", progressbar = TRUE)
    
    cov_info = COVARIATES_FILE %>% 
      dplyr::filter(tech_or_design == "Design",
                    ome == assay,
                    tissue %in% c(tissue, "all"),
                    covariate != "visitcode") #this is for training only
    
    num_cov = cov_info %>% filter(data_type == "numerical") %>% pull(covariate)
    #we manually add metab covariate inclusion, because normally DA is done
    #per platform but here we're doing for all metabolites. So there's no
    #metab in the covariates file formally. 
    if(assay == "metab") num_cov = c("BMI", "calculatedAge")
    num_part = paste(num_cov, collapse = " + ")
    
    factor_cov = cov_info %>% filter(data_type == "factor") %>% pull(covariate)
    #so we are fitting things according to the group timepoint framework as
    #a cell means model, just like the DA. 
    factor_cov = factor_cov[!factor_cov %in%  c("randomGroupCode", "Timepoint")]
    factor_cov = c(factor_cov, "group_timepoint")
    
    #we manually add metab covariate inclusion, because normally DA is done
    #per platform but here we're doing for all metabolites. So there's no
    #metab in the covariates file formally. 
    if(assay == "metab") factor_cov = c("pid", "Sex", "group_timepoint")
    
    factor_part = factor_cov %>%
      paste0("(1 | ", ., ")") %>%
      paste(collapse = " + ")
    
    var_part_formula = paste("~", num_part, "+", factor_part)
    var_part_model = fitExtractVarPartModel(exprObj = qc_norm,
                                            formula = var_part_formula,
                                            data = sample_metadata)
    var_part_results[[tissue]][[assay]] = var_part_model
  }
}
saveRDS(var_part_results, file = file.path(here(), "data", "tmp", "var_exp_results.RDS"))
```

```{r}
var_part_results = readRDS(file.path(here(), "data", "tmp", "var_exp_results.RDS"))
for(tissue in names(var_part_results)){
  for(assay in names(var_part_results[[tissue]])){
    pdf(file = file.path(figure_out_path,"var_explained", paste0(tissue,"_", assay, "_var_exp.pdf")))
    
    figure = plotVarPart(var_part_results[[tissue]][[assay]], 
                         main = paste(tissue, assay, "variance explained"))
    print(figure)
    dev.off()
  }
}
```

```{r}
#structure is [[tissue]][[ome]] for ea df (well it can be converted to df),
#just rbind my way into an actual big df
#also i just realized that i used assay for everything above but ome here. oops
var_part_df = bind_rows(
  lapply(names(var_part_results), function(tissue) {
    lapply(names(var_part_results[[tissue]]), function(ome) {
      df = as.data.frame(var_part_results[[tissue]][[ome]])
      df %>%
        tibble::rownames_to_column("feature") %>%
        mutate(tissue = tissue, ome = ome)
    }) %>% bind_rows()
  })
)
#we also include this information in a supplementary table. 
saveRDS(var_part_df, file.path(repo_local_dir, "figures", "table_S3.RDS"))


plot_df = var_part_df %>%
  pivot_longer(
    cols = c("pid", "group_timepoint", "Sex", "BMI", "calculatedAge"),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(ome = case_when(
    grepl("metab", ome) ~ "Metabolomics",
    grepl("prot-ol|prot-pr", ome) ~ "Proteomics", #collapse into prot
    ome == "prot-ph" ~ "Phosproteomics", #collapse into prot
    ome == "epigen-atac-seq" ~ "ATAC",
    ome == "transcript-rna-seq" ~ "RNA",
    TRUE ~ ome
  ),
  variable = case_when(
    variable == "pid" ~ "participant",
    variable == "randomGroupCode" ~ "Group",
    variable == "calculatedAge" ~ "Age",
    TRUE ~ variable
  )) %>%
  mutate(variable = factor(variable,
                           levels = c("participant", "group_timepoint", "Sex", "BMI", "Age")))

pdf(file = file.path(figure_out_path, "var_explained", paste0("variance_exp_facetted.pdf")))
p = plot_df %>%
  ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_violin(scale = "width", trim = TRUE, color = NA) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 1.5, color = "black") +
  geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", fill = NA) +
  facet_grid(ome ~ tissue) +
  labs(
    x = "Covariate",
    y = "Variance Explained",
    title = "Distribution of Variance Explained per covariate"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 7),  
    panel.spacing = unit(1, "lines"),
    legend.position = "none"
  )

print(p)
dev.off()

```
