---
title: "deconvolution_diagnosis"
output: html_document
date: "2026-01-22"
---


```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(tibble)
library(ggplot2)
library(MotrpacHumanPreSuspensionData)
library(MotrpacHumanPreSuspensionAnalysis)

config = jsonlite::fromJSON("~/config.json")
repo_local_dir = config$precovid_repo_path

source("deconv_helpers.R")
```

## Methods

To evaluate whether shifts in cellular composition confounded the differential expression results, a sensitivity analysis was performed using deconvolution-derived cell-type proportions in blood. Cellular composition was first estimated from whole-blood transcriptomes by deconvolving the bulk RNA-seq profiles with CIBERSORTx (cibersortx.stanford.edu) using the LM22 leukocyte signature matrix in 'Batch Correct' mode. Because LM22 resolves 22 closely related leukocyte subsets whose marker profiles are highly correlated, the subsets were collapsed by summing their fractions into broader populations: T cells, B cells, NK cells, monocytes, macrophages, dendritic cells, neutrophils, and mast cells. The reliability of these estimates was then assessed by correlating CIBERSORTx proportions at pre-exercise with matched complete blood count values for the same participants, where available. Finally, the blood RNA-seq differential analysis was refit with the coarse cell-type fractions (T cells, B cells, NK cells, monocytes, and neutrophils) added as numeric covariates. Proportions were z-scored before inclusion, and the model otherwise matched the original differential analysis pipeline. Comparing the refit results against the original model indicated whether accounting for cellular composition altered the estimated exercise effects.

## Results

After collapsing the LM22 subsets, blood transcriptomes were deconvolved into seven leukocyte populations across 807 samples from 173 participants spanning the endurance, resistance, and control arms. Neutrophils were the dominant estimated population (mean 50.7% of cells), followed by T cells (21.0%), NK cells (8.4%), B cells (5.9%), macrophages (4.9%), monocytes (4.6%), and mast cells (4.2%).

Estimated neutrophil proportions at pre-exercise were moderately correlated with matched complete blood count neutrophil values (Pearson R = 0.40 across all participants; R = 0.53 after excluding two samples with implausibly low clinical neutrophil counts).

Adding the coarse cell-type fractions as covariates had little effect on the blood RNA-seq differential analysis. Across exercise arms and timepoints, log fold-changes from the covariate-adjusted model were nearly identical to the original estimates (Pearson R = 0.88-0.98). 

## Algorithm choice

We use `cibersortx.stanford.edu` 's webpage for deconvolution which doesn't have a direct R package implementation that I've found.

### Reference resolution

Another important consideration is the resolution the reference is at—some references will try and be exceptionally precise in terms of the cell type (e.g. try to differentiation signal between Tregs and Gamma T cells), but this level of precision is (1) Not that useful for our study, (2) leads to instability in some models, depending on the algorithm used. The cibersort results document gives some information for correlation to marker genes and the correlation is not that great in many circumstances. I am worried that the low representation + high variance is not appropriate. 

For (1) - Any clinical measurements or decision making is not going to have cells profiled at an extremely high specificity, so, even if the algorithm is exactly correct, knowing differences between resting and activated Dendritic cells doesn't help with the research question that we're approaching.

For (2) - This is partially because of co-linearity in some very similar cell types, e.g. if activated and resting dendritic cells are 99% similar for the markers in our dataset (just hypothetically), any methods using linear model approaches can suffer from co-linearity issues.


### Note/Data loading. 

These results were previously generated by Greg for the blood paper using the direct qc-norm implementation. The instructions suggest that just directly using transcripts per million with "batch correction mode" in cibersort could be better. This analysis is preliminary and for any manuscript submissions I will need to review and do additional sanity checks on differences due to inputs. - Chris 04/21/2026. 

```{r}
cibersort_results = file.path(repo_local_dir,
                              "figures",
                              "blood",
                              "final_figure_input_data_files",
                              "CIBERSORTx_PrecovidBlood_RNADecon_Results.csv")
cibersort_results = read.csv(cibersort_results)
```


```{r, include = FALSE}
signature_matrix_LM22 = read.csv(file = "LM22_source_GEPs.txt",
                                 sep = "\t") %>%
  tibble::column_to_rownames("genesinput") %>%
  as.matrix()

row_vars = apply(signature_matrix_LM22, 1, var, na.rm = TRUE)
top5000 = order(row_vars, decreasing = TRUE)[1:5000]

# subset to top n most variable markers for plotting purposes.
p = signature_matrix_LM22[top5000, ] %>%
  t() %>%
  pheatmap::pheatmap(
    scale = "row",
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    show_rownames = TRUE,
    show_colnames = FALSE
  )

print(p)
```

This plot just generally visualizes the reference dataset `LM22` and shows how specific types of cells relate to each other. The reference itself has a TON (\~11k) of genes, and in most cases, the cell types are quite similar for most genes. Overall, the hierarchical clustering ends up looking how we'd expect based on upstream cell types (e.g. Macrophage M0, m2 are similar) though.

For a more direct comparison, I just directly draw a scatterplot for the correlation of 2 specific (similar-ish) cell types

```{r}
signature_matrix_LM22_plot = signature_matrix_LM22 %>%
  as.data.frame() %>%
  mutate(log_counts_dendritic_resting = log2(Dendritic.cells.resting),
         log_counts_dendritic_activated = log2(Dendritic.cells.activated))

r2_value = cor(signature_matrix_LM22_plot$log_counts_dendritic_resting,
               signature_matrix_LM22_plot$log_counts_dendritic_activated)^2

p = signature_matrix_LM22_plot %>%
  ggplot(aes(x = log_counts_dendritic_resting,
             y = log_counts_dendritic_activated)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  annotate("text", 
           x = max(signature_matrix_LM22_plot$log_counts_dendritic_resting, na.rm = TRUE), 
           y = max(signature_matrix_LM22_plot$log_counts_dendritic_activated, na.rm = TRUE), 
           label = paste0("R^2 = ", round(r2_value, 3)), 
           hjust = 5, vjust = 1)

print(p)
```

### Quick detour for attempting to fix this:

Here, I had a strategy to test what deconvolution results would look like if we were to condense cell types by expected cell fraction. For example, if we knew 20% of all Macrophages were 'Macrophage.M0', 20% were 'Macrophage.M1', and 60% were 'Macrophage.M2', we could find a weighted average that represents 'Macrophage'. This could improve deconvolution model stability and keep results at a resolution that is appropriate for our dataset. This could be better than trying to estimate specific cell type fractions and then adding up values after the deconvolution is performed. Note: supposedly the CIBERSORT algorithm does a lot better at digesting this colinearity, and is well suited for the type of data that we have in our project.


### Actual coding implementation


To address multicolinearity and different resolutions of cell types, we combine things
into different categories like this

```{r}
cibersort_merge_df = cibersort_results %>%
  dplyr::select(-c(P.value, Correlation, RMSE)) %>%
  squish_proportions() %>%
  dplyr::rename(vialLabel = Mixture) %>%
  dplyr::mutate(vialLabel = as.character(vialLabel))

head(cibersort_merge_df %>% dplyr::select(-vialLabel))
```

```{r}
blood_qc = load_qc(selected_tissues = "blood",
                   selected_omes = "transcript-rna-seq")

blood_metadata = blood_qc$blood$`transcript-rna-seq`$sample_metadata %>%
  dplyr::mutate(vialLabel = as.character(vialLabel)) %>%
  dplyr::select(vialLabel, pid, Timepoint, randomGroupCode) %>%
  inner_join(cibersort_merge_df, by = "vialLabel") %>%
  dplyr::mutate(group_timepoint = interaction(randomGroupCode,Timepoint, sep = ".")) %>%
  dplyr::select(-Eosinophils)

write.csv(blood_metadata, 
          file = "cibersort_with_metadata.csv",
          row.names = FALSE)

desired_comparisons = .generate_contrasts_acute(blood_metadata)
```

```{r}
.parse_contrast_vec = function(contrast_str, coef_names) {
  contrast_vec = setNames(rep(0.0, length(coef_names)), coef_names)
  tokens = strsplit(trimws(contrast_str), "\\s*[+-]\\s*")[[1]]
  signs = c(1L, ifelse(regmatches(contrast_str, gregexpr("[+-]", contrast_str))[[1]] == "-", -1L, 1L))
  for (i in seq_along(tokens)) {
    nm = trimws(tokens[i])
    if (nm %in% coef_names) contrast_vec[nm] = contrast_vec[nm] + signs[i]
  }
  contrast_vec
}

meta_cols = c("vialLabel", "pid", "Timepoint", "randomGroupCode", "group_timepoint")
cell_cols = setdiff(colnames(blood_metadata), meta_cols)

lmer_results = purrr::map_dfr(cell_cols, function(cell_type) {
  df = blood_metadata %>%
    dplyr::select(dplyr::all_of(c("pid", "group_timepoint", cell_type))) %>%
    dplyr::rename(y = dplyr::all_of(cell_type)) %>%
    dplyr::mutate(y = pmax(pmin(qlogis(y / 100), 5), -5))

  fit = lmerTest::lmer(y ~ 0 + group_timepoint + (1 | pid), data = df)
  coef_names = names(lme4::fixef(fit))
  
  purrr::map_dfr(desired_comparisons, function(contrast_str) {
    L = .parse_contrast_vec(contrast_str, coef_names)
    beta = lme4::fixef(fit)
    V = as.matrix(vcov(fit))
    estimate = as.numeric(L %*% beta)
    se = as.numeric(sqrt(L %*% V %*% L))
    tstat = estimate / se
    df_satt = as.numeric(lmerTest::contest(fit, L = L, rhs = 0, joint = FALSE)[["df"]])
    p_value = 2 * pt(abs(tstat), df = df_satt, lower.tail = FALSE)
    tibble::tibble(
      cell_type = cell_type,
      contrast = contrast_str,
      p_value = p_value
    )
  })
})

lmer_results = lmer_results %>%
  dplyr::group_by(contrast) %>%
  dplyr::mutate(adj_p_value = p.adjust(p_value, method = "BH")) %>%
  dplyr::ungroup()
```



# Update: Comparing to actual clinical measurements

Motrpac has some CBC values 

```{r}
cbc_raw = MotrpacHumanPreSuspensionData::cln_raw_local_lab_results$data
# There are 2 samples with mistaken neutrophil values (≤5). The figures below show results with and without excluding them.
```

```{r}
# Excluding CBC neutrophil values ≤5 (2 suspected erroneous samples)
p_filtered = plot_cbc_vs_cibersortx(blood_metadata, cbc_raw, filter_lt5 = TRUE)
print(p_filtered)
ggsave("cbc_vs_cibersortx_filtered.pdf", plot = p_filtered, width = 7, height = 5)
```

```{r}
# All samples included (no filter on CBC neutrophil value)
p_unfiltered = plot_cbc_vs_cibersortx(blood_metadata, cbc_raw, filter_lt5 = FALSE)
print(p_unfiltered)
```

**Figure legend.** *Concordance between CIBERSORTx-estimated and clinically measured neutrophil
proportions.* Each point is a participant, comparing the clinical complete blood count (CBC)
neutrophil value (x-axis) with the CIBERSORTx-deconvolved neutrophil percentage (y-axis), both
measured at the pre-exercise timepoint. The blue line is the ordinary least-squares fit and the
Pearson correlation (*r*) is annotated. Two versions are shown: one excluding two samples with
implausibly low CBC neutrophil values (≤ 5; *r* = 0.53) and one including all samples
(*r* = 0.40). The moderate positive correlation indicates that the deconvolution estimates track
clinically measured neutrophil abundance.



## Adding cell types as covariates
See `DA_with_cell_types.R` to see how I'm implementing DA models with additional covariates. 

The DA is not staged anywhere but was put in the repo_local_dir part. 
```{r}
finished_da_with_cell_types = read.csv(file.path(repo_local_dir, "data", "tmp", "freeze_DA",
                                                 "human-precovid-sed-adu_t04-blood-rna_transcript-rna-seq_da_dream-acute-cell_types_v1.3.txt"),
                                       sep = "\t")
existing_motrpac_da = load_differential_analysis(selected_omes = "transcript-rna-seq",
                                                 selected_tissues = "blood",
                                                 single_matrix = TRUE) %>%
  dplyr::select(feature_id, z.std, logFC, contrast)

combined_da = finished_da_with_cell_types %>%
  dplyr::select(feature_id, z.std, logFC, contrast) %>%
  dplyr::left_join(
    existing_motrpac_da %>% dplyr::select(feature_id, z.std, logFC, contrast),
    by = c("feature_id", "contrast"),
    suffix = c(".cell_type", ".no_cell_type")
  ) %>%
  dplyr::left_join(MotrpacHumanPreSuspensionAnalysis::CONTRAST_CONVERTER, by = "contrast") %>%
  dplyr::filter(contrast_type == "exercise_with_controls")

```

```{r}
timepoint_short = c(
  "pre_exercise"      = "Pre",
  "during_20_min"     = "D20",
  "during_40_min"     = "D40",
  "post_10_min"       = "P10",
  "post_15_30_45_min" = "P15-45",
  "post_3.5_4_hr"     = "P35",
  "post_24_hr"        = "P24"
)

group_colors = MotrpacHumanPreSuspensionAnalysis::HUMAN_EXERCISE_GROUP_COLORS
names(group_colors) = dplyr::recode(names(group_colors),
                                    "ADUEndur" = "EE", "ADUResist" = "RE", "ADUControl" = "CON")

timepoint_colors = MotrpacHumanPreSuspensionAnalysis::HUMAN_ACUTE_TIMEPOINT_COLORS
names(timepoint_colors) = dplyr::recode(names(timepoint_colors), !!!as.list(timepoint_short))

# White strip text on dark fills, black on light fills
.contrast_text = function(fills) {
  rgb_mat = grDevices::col2rgb(fills) / 255
  luminance = 0.299 * rgb_mat[1, ] + 0.587 * rgb_mat[2, ] + 0.114 * rgb_mat[3, ]
  ifelse(luminance < 0.55, "white", "black")
}

.plot_covariate_corr = function(da, metric, outfile) {
  x_col = paste0(metric, ".no_cell_type")
  y_col = paste0(metric, ".cell_type")

  plot_df = da %>%
    dplyr::filter(!is.na(.data[[x_col]]), !is.na(.data[[y_col]]),
                  randomGroupCode %in% c("ADUEndur", "ADUResist")) %>%
    dplyr::mutate(
      group = dplyr::recode(as.character(randomGroupCode),
                            "ADUEndur" = "EE", "ADUResist" = "RE", "ADUControl" = "CON"),
      timepoint = dplyr::recode(as.character(Timepoint), !!!as.list(timepoint_short))
    ) %>%
    dplyr::mutate(
      group = factor(group, levels = intersect(c("EE", "RE", "CON"), group)),
      timepoint = factor(timepoint, levels = unname(timepoint_short[timepoint_short %in% timepoint]))
    )

  corr_labels = plot_df %>%
    dplyr::group_by(group, timepoint) %>%
    dplyr::summarise(
      r = cor(.data[[x_col]], .data[[y_col]], use = "complete.obs"),
      .groups = "drop"
    ) %>%
    dplyr::mutate(label = paste0("R = ", round(r, 2)))

  col_fills = unname(timepoint_colors[levels(plot_df$timepoint)])
  row_fills = unname(group_colors[levels(plot_df$group)])

  p = plot_df %>%
    ggplot(aes(x = .data[[x_col]], y = .data[[y_col]])) +
    geom_point(size = 0.4, alpha = 0.25) +
    geom_abline(slope = 1, intercept = 0, color = "firebrick", linetype = "dashed", linewidth = 0.6) +
    geom_text(data = corr_labels,
              aes(x = -Inf, y = Inf, label = label),
              hjust = -0.15, vjust = 1.5,
              size = 3.5, inherit.aes = FALSE) +
    ggh4x::facet_grid2(
      group ~ timepoint,
      strip = ggh4x::strip_themed(
        background_x = lapply(col_fills, function(f) element_rect(fill = f, color = "grey20")),
        background_y = lapply(row_fills, function(f) element_rect(fill = f, color = "grey20")),
        text_x = lapply(.contrast_text(col_fills), function(c) element_text(color = c, size = 11, face = "bold")),
        text_y = lapply(.contrast_text(row_fills), function(c) element_text(color = c, size = 11, face = "bold"))
      )
    ) +
    labs(
      x = paste0(metric, " (no cell type covariate)"),
      y = paste0(metric, " (with cell type covariate)"),
      title = paste0("Effect of adding cell type covariates on blood RNA-seq DA (", metric, ")")
    ) +
    theme_bw() +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      title = element_text(size = 13)
    )
  ggsave(outfile, plot = p, width = 8, height = 4)
  p
}

print(.plot_covariate_corr(combined_da, "z.std", "cell_type_covariate_inclusion.pdf"))
print(.plot_covariate_corr(combined_da, "logFC", "cell_type_covariate_inclusion_logFC.pdf"))
```

**Figure legend.** *Effect of adding deconvolution-derived cell-type covariates on the blood
RNA-seq differential analysis.* Per-transcript exercise effect estimates (logFC) from the
original model (x-axis) are plotted against estimates from the model that additionally includes
the z-scored coarse cell-type fractions (T cells, B cells, NK cells, monocytes, and neutrophils)
as numeric covariates (y-axis). Each point is a transcript; panels are faceted by exercise group
(rows: EE, endurance exercise; RE, resistance exercise) and timepoint (columns: D20/D40, 20/40 min
during exercise; P10/P1545/P35/P24, 10 min, 15-45 min, 3.5-4 hr, and 24 hr post exercise). The
dashed red line is the line of identity (identical estimates from both models) and the Pearson
correlation (*R*) is annotated in each panel. Estimates falling tightly along the identity line
indicate that accounting for cellular composition did not substantively alter the estimated
exercise effects.


