---
title: "re_vs_ee_visualizations"
output: html_document
date: "2025-07-07"
---

```{r setup, include=FALSE}
library(MotrpacHumanPreSuspensionAnalysis)
library(dplyr)
library(tidyr)

# ----complexupset is giving problems with ggplot 4.0.0----
if (!requireNamespace("ggplot2", quietly = TRUE) ||
    packageVersion("ggplot2") != "3.5.2") {
  stop("ggplot2 3.5.2 is required. Please install with:
       remotes::install_version('ggplot2', version = '3.5.2')")
}
library(ggplot2)
library(ggrepel)
library(stringr)
library(ComplexUpset)
library(patchwork)
library(purrr)
library(here)
library(ComplexHeatmap)
library(limma)
knitr::opts_chunk$set(echo = TRUE)
```

# General Outline

Ideas that we're trying to present in 
```{r, width = 15, include = F}
ee_vs_re = load_differential_analysis(selected_omes = "all",
                                      selected_tissues = "all",
                                      single_matrix = TRUE) 
ee_re_sig = ee_vs_re %>%
  dplyr::filter(contrast_type == "Endur_vs_Resist") %>%
  dplyr::select(tissue, assay, feature_id, Timepoint, adj_p_value, logFC)

ee_re_wide = ee_vs_re %>%
  dplyr::filter(contrast_type == "exercise_with_controls") %>%
  dplyr::select(tissue, assay, feature_id, Timepoint, randomGroupCode, logFC, adj_p_value) %>%
  pivot_wider(names_from = c("randomGroupCode"),
              values_from = c("logFC", "adj_p_value")) %>%
  merge(., ee_re_sig, by = c("tissue", "assay", "feature_id", "Timepoint")) %>%
  dplyr::mutate(sig_contrast = ifelse(adj_p_value < 0.05, "Sig", "Non-Sig"),
         sig_ee_single = ifelse(adj_p_value_ADUEndur < 0.05, "Sig", "Non-Sig"),
         sig_re_single = ifelse(adj_p_value_ADUResist < 0.05, "Sig", "Non-Sig"))

categorized = ee_re_wide %>%
  dplyr::mutate(
    Significant_in = dplyr::case_when(
      sig_contrast == "Sig" & sig_ee_single == "Sig" & sig_re_single == "Non-Sig" ~ "Only EE",
      sig_contrast == "Sig" & sig_ee_single == "Non-Sig" & sig_re_single == "Sig" ~ "Only RE",
      sig_contrast == "Sig" & sig_ee_single == "Sig" & sig_re_single == "Sig" ~ "Both Groups",
      sig_contrast == "Sig" & sig_ee_single == "Non-Sig" & sig_re_single == "Non-Sig" ~ "Neither",
      TRUE ~ "Not RE vs EE Hit"
    ),
    border_color = dplyr::case_when(
      sig_contrast == "Sig" ~ "black",
      Significant_in == "Only EE" ~ HUMAN_EXERCISE_GROUP_COLORS[["ADUEndur"]],
      Significant_in == "Only RE" ~ HUMAN_EXERCISE_GROUP_COLORS[["ADUResist"]],
      Significant_in == "Both Groups" ~ "brown",
      Significant_in == "Neither" ~ "#86C5D8",
      TRUE ~ "grey"
    )
  ) %>%
  dplyr::group_by(tissue, feature_id) %>%
  #so for a given tissue, we just squash timepoints by picking the most significant one. 
  dplyr::slice_min(adj_p_value) %>%
  dplyr::ungroup()

ggplot_re_ee = ggplot() +
  #adding axis plots to clearly identify quadrants
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black") +
  geom_hline(yintercept = 0, color = "black") +
  geom_point(
    data = categorized %>% dplyr::filter(sig_contrast != "Sig"),
    aes(x = logFC_ADUEndur, y = logFC_ADUResist),
    color = "grey", size = 0.5, shape = 20
  ) +
  geom_point(
    data = categorized %>% dplyr::filter(sig_contrast == "Sig"),
    aes(x = logFC_ADUEndur, y = logFC_ADUResist, fill = Significant_in),
    color = "black", size = 1.5, shape = 21, stroke = 0.4
  ) +
  #we could add a different line in red that describes how the "opposite" 
  #effect could look like. personally the red line looks a little bad
  # geom_abline(slope = -1, intercept = 0, linetype = "dashed", color = "red") +
  scale_fill_manual(
    values = c(
      "Only EE" = HUMAN_EXERCISE_GROUP_COLORS[["ADUEndur"]],
      "Only RE" = HUMAN_EXERCISE_GROUP_COLORS[["ADUResist"]],
      "Both Groups" = "brown",
      "Neither" = "#86C5D8"
    )
  ) +
  labs(
    x = "logFC (EE vs Control)", 
    y = "logFC (RE vs Control)",
    title = NULL
    # caption = "Small grey dots represent features not significant in EE vs RE comparison,
    # colored dots represent RE vs EE significant hits"
  ) +
  facet_grid(~ tissue) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 12))
print(ggplot_re_ee)
```

This chunk of code is a quick detour for text in the manuscript - corresponding
to what is now figure 3E
```{r}
oppositional = ee_re_wide %>% 
  dplyr::filter(adj_p_value_ADUEndur < 0.05 & adj_p_value_ADUResist < 0.05) %>%
  dplyr::filter(sign(logFC_ADUEndur) != sign(logFC_ADUResist))
```

```{r}
upset_list = lapply(unique(categorized$tissue), function(tissue) {
  #getting this into a format ready for an upset plot
  df_tiss = categorized %>%
    dplyr::filter(tissue == !!tissue) %>%
    #we're also newly adding assay to try to make stacked barplots
    dplyr::select(feature_id, assay, sig_contrast, sig_ee_single, sig_re_single) %>%
    dplyr::distinct() %>%
    dplyr::mutate(dplyr::across(c(sig_contrast, sig_ee_single, sig_re_single), ~ . == "Sig")) %>%
    dplyr::rename(
      EE_vs_RE = sig_contrast,
      EE_vs_Ctrl = sig_ee_single,
      RE_vs_Ctrl = sig_re_single
    ) %>%
    #so we notably remove the triple negative--those not significant in any 
    #modality specific, and also the re vs ee. 
    dplyr::filter(EE_vs_RE | EE_vs_Ctrl | RE_vs_Ctrl)
  
  #custom colors. Notably, if theres an intersection that one tissue has
  #but not another, it doesn't run at all. so i have to add thing specially
  #bc adipose is missing a "both" feature. 
  custom_colors = list(
    upset_query(
      intersect = c('EE_vs_RE', 'EE_vs_Ctrl'),
      color = HUMAN_EXERCISE_GROUP_COLORS[["ADUEndur"]],
      fill = HUMAN_EXERCISE_GROUP_COLORS[["ADUEndur"]],
      only_components = c('intersections_matrix', 'Intersection size')
    ),
    upset_query(
      intersect = c('EE_vs_RE', 'RE_vs_Ctrl'),
      color = HUMAN_EXERCISE_GROUP_COLORS[["ADUResist"]],
      fill = HUMAN_EXERCISE_GROUP_COLORS[["ADUResist"]],
      only_components = c('intersections_matrix', 'Intersection size')
    ),
    upset_query(
      intersect = c('EE_vs_RE'),
      color = '#86C5D8',
      fill = '#86C5D8',
      only_components = c('intersections_matrix', 'Intersection size')
    )
  )
  
  
  if(tissue == "muscle" | tissue == "blood"){    
    custom_colors[[4]] = upset_query(
      intersect = c('EE_vs_RE', 'EE_vs_Ctrl', 'RE_vs_Ctrl'),
      color = 'brown',
      fill = 'brown',
      only_components = c('intersections_matrix', 'Intersection size')
    )
  }
  
  ComplexUpset::upset(
    df_tiss,
    intersect = c("EE_vs_RE", "EE_vs_Ctrl", "RE_vs_Ctrl"),
    #so here we don't want to just go in the same order every time
    sort_intersections = FALSE,
    sort_sets = FALSE,
    intersections = list(
      c("EE_vs_RE", "EE_vs_Ctrl", "RE_vs_Ctrl"),
      c("RE_vs_Ctrl", "EE_vs_RE"),
      c("EE_vs_Ctrl", "EE_vs_RE"),
      "EE_vs_RE",
      "RE_vs_Ctrl",
      "EE_vs_Ctrl"
    ),
    name = tissue,
    encode_sets = FALSE,
    min_size = 3,
    set_sizes = FALSE,
    base_annotations = list(
      'Intersection size' = intersection_size(
        counts = TRUE) +
        #remove assay legend bc itll get printed 3 times
        guides(fill = "none")
    ),
    matrix = intersection_matrix(
      geom = geom_point(shape = 'square', size = 3.5),
      segment = geom_segment(linetype = 'dotted')
    ),
    queries = custom_colors
  )
})

combined_upset_plot = patchwork::wrap_plots(upset_list, nrow = 1)
print(combined_upset_plot)
```


```{r}
final_plot = ggplot_re_ee / combined_upset_plot + plot_layout(heights = c(1.5, 1.2))

pdf(file = file.path("figure3a_re_vs_ee.pdf"),
    width = 12, height = 8)
print(final_plot)
dev.off()
```


Okay now lets do some ORA here
```{r}
all_ora_results = data.frame()
for(tissue in tissue_available_list()){
  unique_assays = categorized %>% dplyr::filter(tissue == !!tissue) %>% dplyr::pull(assay) %>% unique()
  for(assay in unique_assays){
    #we're not going to do refmet or ptm-sea type stuff here. 
    if(assay == "metab" | assay == "prot-ph") next
    
    for(groups in c("Only EE", "Only RE")){
      #filler contrast is to make sure the heatmap has the right colors
      if(groups == "Only EE") filler_contrast = "EE"
      if(groups == "Only RE") filler_contrast = "RE"
      
      #to make the background for the ORA. 
      all_results = categorized %>% 
        dplyr::filter(tissue == !!tissue,
               assay == !!assay) %>%
        dplyr::left_join(HUMAN_FEATURE_TO_GENE, by = c("feature_id", "assay")) %>%
        dplyr::filter(!is.na(gene_symbol))
      
      #significant in the certain group we want. 
      sig_results = all_results %>% 
        dplyr::filter(Significant_in == groups,
               adj_p_value < 0.1)
      
      message(paste(tissue, groups, assay, nrow(sig_results)))
      #if its <50 samples, we dont even do ORA. 
      if(nrow(sig_results) < 50) next
      # write.csv(sig_results, 
      #           file = file.path(here(), "data", "tmp", "figures", paste0(filler_contrast,tissue,"only_ORA_enrichments.csv")),
      #           row.names = FALSE)
      #bc its a factor otherwise
      significant = as.character(unique(sig_results$gene_symbol))
      all_feats = as.character(unique(all_results$gene_symbol))
      
      ora_results = run_ORA(input = significant,
                            background = all_feats) 
      ora_formatted = ora_results %>%
        dplyr::mutate(
          tissue = !!tissue,
          assay = assay,
          contrast = filler_contrast,
          contrast_type = "exercise_with_controls",
          #basically just qnorm normal to find the zscore
          z.std = qnorm(1 - p_value / 2)
        )
      all_ora_results = rbind(all_ora_results, ora_formatted)
    }
  }
}

all_ora_results = all_ora_results %>%
  dplyr::mutate(z.std = dplyr::case_when(
    z.std == "Inf" ~ 15,
    TRUE ~ z.std
  ),
  contrast = as.factor(contrast)
  )
```

So here we make the actual figure and set colors and stuff
```{r}
# write.csv(all_ora_results, 
#           file = file.path(here(), "data", "tmp", "figures", "all_muscle_re_ee_only_ORA_results.csv"),
#           row.names = FALSE)
#okay just take n from ea group in a really brute force way
all_ora_results_figure = all_ora_results %>%
  dplyr::filter(tissue == "muscle") %>%
  dplyr::group_by(contrast) %>%
  dplyr::arrange(adj_p_value, .by_group = TRUE) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::ungroup() %>%
  dplyr::filter(adj_p_value < 0.05) %>%
  #strip the database prefix (e.g. GOCC_, HALLMARK_) off the set name, then cap at
  #35 characters with a trailing ... if longer
  dplyr::mutate(neg_log10_p = -log10(p_value),
                set_short = stringr::str_remove(set, paste0("^", database, "_")),
                set_short = ifelse(nchar(set_short) > 35,
                                   paste0(substr(set_short, 1, 35), "..."),
                                   set_short))

#by default these arrange themselves alphabetically so instead we just 
#force them to get arranged by contrast like they were before. 
ordered_sets = all_ora_results_figure %>%
  dplyr::pull(set_short) 

contrast_colors = c(
  "EE" = HUMAN_EXERCISE_GROUP_COLORS[["ADUEndur"]],
  "RE" = HUMAN_EXERCISE_GROUP_COLORS[["ADUResist"]]
)

enrichmap(all_ora_results_figure,
          n_top = 30L,
          set_column = "set_short",
          statistic_column = "neg_log10_p",
          contrast_column = "contrast", 
          padj_column = "adj_p_value",
          colors = c("white", "#543483"),
          filename = file.path("Fig3b_re_ee_only_enrichments.pdf"),
          width = 7.5, 
          height = 7.5,
          heatmap_args = list(
            column_title = "Muscle \n Transcriptomics",
            row_order = ordered_sets,
            "cluster_columns" = FALSE,
            "cluster_rows"= FALSE,
            na_col = "grey55",
            top_annotation = HeatmapAnnotation(
              Modality = c("EE", "RE"),
              col = list(Modality = contrast_colors),
              show_annotation_name = TRUE
            ),
            heatmap_legend_param = list(
              title = latex2exp::TeX("$\\bf{$-log$_{10}($P-Value$)}$")
            )
          )
)
```


The goal of this is:

1. To visualize how things look compared to the previous plot of feature level results

2. To find interesting pathways to present using a triangle type plot (see below)


```{r}
# triangle_size_scale = 1 #easier parameter to change for testing iterations

camera_tp_group = CAMERA_RESULTS %>%
  dplyr::mutate(contrast_left = stringr::str_split_fixed(contrast_short, " - ", 2)[,1],
                randomGroupCode = dplyr::case_when(
                  contrast_type == "Endur_vs_Resist" ~ "ADUEndur - ADUResist",
                  TRUE ~ paste0("ADU", stringr::str_split_fixed(contrast_left, "\\.", 2)[,1])
                ),
                Timepoint = as.factor(stringr::str_split_fixed(contrast_left, "\\.", 2)[,2])) %>%
  dplyr::select(-contrast_left) 

enrich_results = camera_tp_group %>%
  dplyr::filter(contrast_type == "Endur_vs_Resist") %>%
  dplyr::select(tissue, assay, set_id, set_short, adj_p_value, contrast_short, t, randomGroupCode, Timepoint)
#-----------------------------------------------------------------------------------------------------
#ok so i made this code to become wide -> merge -> pull in overall RE-EE sig for each group
#for a different thing, but the wide -> long code can definitely be improved/trimmed. 
#not really worth worrying about too much yet. 
enrich_wide = camera_tp_group %>%
  dplyr::filter(contrast_type == "exercise_with_controls") %>%
  dplyr::select(tissue, assay, set_id, set_short,  Timepoint, randomGroupCode, t, adj_p_value) %>%
  pivot_wider(names_from = c("randomGroupCode"),
              values_from = c("t", "adj_p_value")) %>%
  merge(., enrich_results, by = c("tissue", "assay", "set_id", "set_short", "Timepoint")) 

# potential_pathways = 
interesting_pathways = readxl::read_excel(path = file.path(here(), "figures", "landscape", "curated_pathways.xlsx"),
                                          sheet = "5") %>% dplyr::pull(set_id)

enrich_wide$set_short = as.character(enrich_wide$set_short)
selected_pathways_long = enrich_wide %>%
  #filter to desired pathways
  #currently we're only keep transcript bc trying to do multiple assays is too messy
  dplyr::filter(set_id %in% interesting_pathways, assay %in% c("transcript-rna-seq")) %>%
  #i need to have a better way to organize set_short bc this is not it 
  dplyr::mutate(set_short = ifelse(
    nchar(set_short) > 35,
    paste0(substr(set_short, 1, 35), "..."),
    set_short
  ))%>%  
  dplyr::rename(t_overall = t, adj_p_value_overall = adj_p_value) %>%
  pivot_longer(
    cols = c(t_ADUResist, t_ADUEndur, adj_p_value_ADUResist, adj_p_value_ADUEndur),
    names_to = c(".value", "condition"),
    names_pattern = "(t|adj_p_value)_(ADUResist|ADUEndur)"
  ) %>%
  dplyr::mutate(outline = ifelse(adj_p_value < 0.05, "signif", "nonsignif"), #significant within single group
         shape = ifelse(t > 0, "Upregulated", "Downregulated"), #up or down within single group
         tile_filt = ifelse(adj_p_value_overall < 0.05, "Adj P <= 0.05", "Adj P > 0.05"), #significant across groups
         size = abs(t), #size based on absolute 
         condition = dplyr::recode(condition,
                            "t_ADUResist" = "ADUResist",
                            "t_ADUEndur" = "ADUEndur")) %>%
  dplyr::arrange(dplyr::desc(size)) %>%
  dplyr::mutate(Timepoint_labels = dplyr::case_when(
    Timepoint == "post_10_min" ~ "P10M",
    Timepoint == "post_15_30_45_min" ~ "P15-45M",
    Timepoint == "post_3.5_4_hr" ~ "P3.5/4H",
    Timepoint == "post_24_hr" ~ "P24H",
    TRUE ~ Timepoint
  ),
  Timepoint_labels = factor(Timepoint_labels, 
                            levels = c("P10M", "P15-45M", "P3.5/4H", "P24H"))) %>%
  droplevels()

# write.csv(selected_pathways_long, 
#           file = file.path(here(), "data", "tmp", "figures", "Fig3C_triangle_selected_pathways.csv"),
#           row.names = FALSE)

```

So now we make the actual plots themselves
```{r}
triangle_plots = ggplot(selected_pathways_long,
                        aes(x = Timepoint_labels, y = set_short)) +
  # Background tile for adjusted P significance
  geom_tile(aes(alpha = tile_filt),
            fill = "grey60",
            colour = NA,
            height = 0.95, width = 0.6) +
  
  # Separate points for endurance and resistance (nudged for clarity)
  geom_point(
    data = subset(selected_pathways_long, condition == "ADUEndur"),
    aes(fill = condition, shape = shape, size = size),
    stroke = 0.8,
    position = position_nudge(y = 0.2)
  ) +
  geom_point(
    data = subset(selected_pathways_long, condition == "ADUResist"),
    aes(fill = condition, shape = shape, size = size),
    stroke = 0.8,
    position = position_nudge(y = -0.2)
  ) +
  
  # Legend for significance tiles
  scale_alpha_manual(
    values = c("Adj P <= 0.05" = 0.3, "Adj P > 0.05" = 0),
    name = "RE-EE"
  ) +
  
  # Legend for fill (Group)
  scale_fill_manual(
    values = HUMAN_EXERCISE_GROUP_COLORS,
    name = "Group",
    guide = guide_legend(
      nrow = 2, byrow = TRUE,
      override.aes = list(shape = 22, size = 5, colour = "black")
    )
  ) +
  
  # Legend for up/down regulation
  scale_shape_manual(
    values = c("Upregulated" = 24, "Downregulated" = 25),
    name = "Direction"
  ) +
  
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(linewidth = 1),
    
    # Axis adjustments
    axis.text.y.right = element_text(colour = "black", size = 8),
    axis.text.y.left = element_blank(),
    axis.ticks.y.left = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(colour = "black", angle = 45, hjust = 1, vjust = 1, size = 10),
    
    # Legend layout
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.justification = c(0, 0),
    legend.box.just = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    plot.margin = margin(t = 5, r = 50, b = 5, l = 8)
  ) +
  
  # Unified legend configuration
  guides(
    fill = guide_legend(
      order = 1,
      nrow = 2, 
      byrow = TRUE,
      override.aes = list(
        shape = 22,          # square shape with fill (not circle)
        size = 5,            # larger for visibility
        colour = "black"     # outline around fill
      )
    ),
    shape = guide_legend(order = 2, nrow = 2, byrow = TRUE),
    size  = guide_legend(
      title = "z-std",
      order = 3,,
      nrow = 3,
      override.aes = list(
        shape = 24,       # triangles for size legend
        fill  = "grey60",
        color = "black"   # add border outline
      )
    ),
    alpha = guide_legend(
      order = 4,
      nrow = 2,
      override.aes = list(
        fill  = "grey60",
        color = "black",   # outline for RE-EE tiles
        shape = 22,
        size  = 4
      )
    )
  ) +
  scale_y_discrete(position = "right") +
  facet_grid(~ tissue, scales = "free_x", space = "free_x") +
  ggtitle("Group-level pathway enrichment")



pdf(file = file.path(here(), "data", "tmp", "figures", "Fig3C_triangle_re_ee_plots.pdf"),
    width = 10, height = 12)
print(triangle_plots)
dev.off()
```







