---
title: "DA_hit_type_exploration"
author: "Dan Katz, Malene Lindholm, Christopher Jin"
date: "2025-06-03"
output: html_document
---

```{r setup}
session_packages = c('pheatmap', 'limma', 'UpSetR', 'ComplexHeatmap', 'gprofiler2',
                     'plier', 'cluster', 'readxl','qvalue', 'reshape2', 'ggh4x','ggVennDiagram','ggpattern',
                     'MASS', 'splitstackshape', 'MotrpacBicQC', 'tidyverse', 
                     'powerjoin','MotrpacHumanPreSuspensionAnalysis','conflicted','patchwork','ggbeeswarm',
                     'circlize', 'grid','gridExtra')


for(i in session_packages) {
  tryCatch(
    {
      library(i, character.only = T)
    }, error = function(e) {
      print(paste("cannot load package:", i,", please install", i))
    }
  )
}

conflict_prefer('select','dplyr')
conflict_prefer('filter','dplyr')

sessionInfo()

```

```{r}
all_omes_except_methyl = ome_available_list()
all_omes_except_methyl = all_omes_except_methyl[all_omes_except_methyl != "epigen-methylcap-seq"]


full_da_results_all_tissues = MotrpacHumanPreSuspensionAnalysis::load_differential_analysis(
  selected_omes = all_omes_except_methyl,
  selected_tissues = "all",
  epigen = TRUE,
  single_matrix = TRUE
) %>%
  mutate(assay = case_when(
    grepl("metab", assay) ~ "Metabolomics", #stick all the metab together
    grepl("prot-ol|prot-pr", assay) ~ "Proteomics",  #stick ol into MS/general Prot
    TRUE ~ assay)
  )  %>%
  mutate(assay = recode(assay, 
                        "transcript-rna-seq" = "Transcriptomics",
                        "prot-ph" = "Phosphoproteomics",
                        "epigen-atac-seq" = "ATAC"))

```


```{r}
wide_by_contrasts = full_da_results_all_tissues %>%   
  dplyr::filter(!contrast_type %in% c("Endur_vs_Resist", "baseline")) %>%
  #so we want to be able to flip some results up and down dependning on if 
  #the logFC is up or down
  dplyr::mutate(direction = sign(logFC)) %>%
  dplyr::select(tissue, assay, randomGroupCode, Timepoint, contrast_type, adj_p_value, feature_id, direction) %>%
  pivot_wider(., names_from = c("contrast_type"), #aaaa i could also just split by group here? will do both
              values_from = adj_p_value) %>%
  group_by(tissue, assay, Timepoint, feature_id) %>%
  #so because its grouped by feature_id/timepoint, we can just use first for
  #finding control_only
  mutate(control_only = ifelse(is.na(control_only), 
                               dplyr::first(na.omit(control_only)), 
                               control_only)) %>% 
  #^ just filling in control values to compare with either group
  ungroup() %>%
  #because we have control_only already, we get rid of this part below, the values
  #are already filled into the other group codes
  dplyr::filter(!randomGroupCode == "Control") %>%
  mutate(Timepoint = factor(Timepoint,
                            levels = c("during_20_min", "during_40_min", "post_10_min", "post_15_30_45_min", "post_3.5_4_hr", "post_24_hr")))
#remove timepoint for blood samples
#now we actually categorize things similar to before. 

```

General categorization thought process here:
Control effect - we want to profile features that are stable in a given exercise
group, but the comparison is significant because of some movement in the controls.
so an experiment without controls would not be able to detect these.

Exercise effect - our general positive control section, the types of effects that
most people would categorize for exercise effect. We see some stimulus in the exercise group,
and the overall comparison to the controls is significant. We don't really care
what direction the control group is going in this section. 

Circadian/Non Exercise effect - these are NOT significant differences between
exercise and control groups. These are features that could be considered
statistically significant in a different study, but because we have controls,
we know that those would be an experimental "false positive". 

Note:: we don't do much looking into directions, magnitudes here. Theoretically, 
this could cause an issue, in cases where certain groups have patterns that 
just dont reach statistical significance, but we are more or less just making
some elaborate upset plots. 
```{r}
wide_by_contrasts_merge = wide_by_contrasts %>%
  mutate(
    exercise_with_controls_cat = ifelse(exercise_with_controls <= 0.05, "Sig", "Non-sig"),
    exercise_no_controls_cat = ifelse(exercise_no_controls <= 0.05, "Sig", "Non-sig"),
    control_only_cat = ifelse(control_only <= 0.05, "Sig", "Non-sig")
  ) %>%
  mutate(
    category = case_when(
      (exercise_no_controls_cat == "Non-sig" & exercise_with_controls_cat == "Sig") ~ "Ctrl-driven exercise effects",
      (exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Sig") ~ "Exercise effects",
      (control_only_cat == "Sig" & exercise_no_controls_cat == "Sig" & exercise_with_controls_cat == "Non-sig") ~ "Non-exercise induced changes",
      TRUE ~ "Other"
    )
  ) %>%
  dplyr::filter(category != "Other")

```


```{r}
# Combine all color mappings into one vector
combined_colors <- c(
  "Non-exercise induced changes" = "#7570b3",
  "Exercise effects" = "brown4",        
  "Ctrl-driven exercise effects" = "brown1",
  HUMAN_OME_COLORS,  
  HUMAN_TISSUE_COLORS,  
  HUMAN_EXERCISE_GROUP_COLORS 
)

#This part is needed because the current data objects
#only list the "fuller" names e.g. ADUResist instead of EE/RE. 
tissue_colors <- setNames(
  HUMAN_TISSUE_COLORS[levels(wide_by_contrasts_merge$tissue)],
  levels(wide_by_contrasts_merge$tissue)
)
group_colors <- setNames(
  HUMAN_EXERCISE_GROUP_COLORS[c("ADUEndur","ADUResist")],
  levels(wide_by_contrasts_merge$randomGroupCode)
)
assay_colors <- setNames(
  HUMAN_OME_COLORS[levels(wide_by_contrasts_merge$assay)],
  levels(wide_by_contrasts_merge$assay)
)
#-----------------------------------

#Make variables into factors first
wide_by_contrasts_merge$assay <- as.factor(wide_by_contrasts_merge$assay)
wide_by_contrasts_merge$tissue <- as.factor(wide_by_contrasts_merge$tissue)
wide_by_contrasts_merge$randomGroupCode <- as.factor(wide_by_contrasts_merge$randomGroupCode)


```


```{r}
#replace names to make things ready for figures
filtered_data <- wide_by_contrasts_merge %>%
  mutate(randomGroupCode = recode(randomGroupCode, 
                                  "ADUEndur" = "EE", 
                                  "ADUResist" = "RE"),
         Timepoint = case_when(
           Timepoint == "during_20_min" ~ "D20M",
           Timepoint == "during_40_min" ~ "D40M",
           Timepoint == "post_10_min" ~ "P10M",
           Timepoint == "post_15_30_45_min" ~ "P15-45M",
           Timepoint == "post_3.5_4_hr" ~ "P3.5/4H",
           Timepoint == "post_24_hr" ~ "P24H",
           TRUE ~ Timepoint
         ),
         Timepoint = factor(Timepoint, 
                            levels = c("D20M", "D40M", "P10M", "P15-45M", "P3.5/4H", "P24H"))) %>%
  group_by(assay, tissue, randomGroupCode, Timepoint, category, direction) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(y_value = ifelse(direction == 1, count, -count))
#so now we're just flipping across the axis in cases where its
#up vs downregulated to add extra info

# Calculate the maximum count for each assay using filtered data
# to inform figure dimensions for each row 
max_counts_by_assay = filtered_data %>%
  summarise(max_count = max(abs(y_value)), .by = "assay")

#--------------------------------------------------------------------------
#so here we try and find the combinations of omic platforms that were NOT run 
#Important--make sure that an assay is NOT run instead of not having any sig results
#because if we start with 'filtered data' that already trims sections without 
#any interesting results. 
existing_combinations = full_da_results_all_tissues %>%
  distinct(tissue, assay, randomGroupCode, Timepoint) %>%
  mutate(Timepoint = case_when(
    Timepoint == "during_20_min" ~ "D20M",
    Timepoint == "during_40_min" ~ "D40M",
    Timepoint == "post_10_min" ~ "P10M",
    Timepoint == "post_15_30_45_min" ~ "P15-45M",
    Timepoint == "post_3.5_4_hr" ~ "P3.5/4H",
    Timepoint == "post_24_hr" ~ "P24H",
    TRUE ~ Timepoint
  ),
  randomGroupCode = recode(randomGroupCode, 
                           "ADUEndur" = "EE", 
                           "ADUResist" = "RE"))
#probably should fix this syntax stuff so i dont have to recode timepoint/group again

# Create complete grid only for valid combinations
# Blood gets all timepoints, muscle and adipose exclude the three specified
all_combinations <- rbind(
  expand.grid(
    Timepoint = unique(filtered_data$Timepoint[filtered_data$tissue == "blood"]),
    assay = unique(filtered_data$assay),
    tissue = "blood",
    randomGroupCode = unique(filtered_data$randomGroupCode)
  ),
  expand.grid(
    Timepoint =  unique(filtered_data$Timepoint[filtered_data$tissue %in% c("muscle", "adipose")]),
    assay = unique(filtered_data$assay),
    tissue = c("muscle", "adipose"),
    randomGroupCode = unique(filtered_data$randomGroupCode)
  )
)

# Identify which combinations are missing from filtered data
# the reason I try to make this grid is to set up which squares need grey backgrounds
missing_combinations <- dplyr::anti_join(all_combinations, existing_combinations,
                                         by = c("Timepoint", "assay", "tissue", "randomGroupCode")) %>%
  left_join(max_counts_by_assay, by = "assay")  %>% # Add the max count for each assay to the missing combinations
  mutate(category = "Not assayed") %>%
  #filter these out because there's no bars here in the first place
  filter(!(tissue %in% c("adipose", "muscle") & Timepoint %in% c("D20M", "D40M", "P10M"))) %>%
  mutate(y_value = max_count) %>% #so we are now also adding up vs down reg so i make 2
  tidyr::uncount(2, .id = "sign") %>%     # duplicate each row twice
  mutate(y_value = ifelse(sign == 1, y_value, -y_value))
#--------------------------------------------------------------------------------
```

```{r}
p = ggplot(filtered_data, aes(x = Timepoint)) +
  # Grey bars for missing assay combinations (background)
  #now with up/down stuff for counting up and down 
  geom_col(
    data = missing_combinations,
    aes(y = y_value),
    fill = "grey80", width = 0.8, alpha = 0.7
  ) +
  geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
  # Stacked exercise-induced bars (signed)
  geom_col(
    data = subset(filtered_data, category != "Non-exercise induced changes"),
    aes(y = y_value, fill = category),
    width = 0.3, position = "stack"
  ) +
  
  # Nudged non-exercise-induced bars (signed)
  geom_col(
    data = subset(filtered_data, category == "Non-exercise induced changes"),
    aes(y = y_value, fill = category),
    width = 0.3, position = position_nudge(x = 0.3)
  ) +
  scale_fill_manual(values = c(combined_colors, "Not assayed" = "grey80")) +
  coord_cartesian(clip = "off") +
  facet_grid2(
    assay ~ tissue + randomGroupCode,
    scales = "free",
    space = "free_x",
    strip = strip_nested(
      background_y = elem_list_rect(fill = assay_colors),
      background_x = elem_list_rect(fill = c(tissue_colors, group_colors, group_colors, group_colors)),
      text_x = list(element_text(colour = "black"), element_text(colour = "black")),
      by_layer_x = FALSE
    )
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.text = element_text(size = 12),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.justification = "top",
    legend.background = element_rect(fill = "white", colour = "black", linewidth = 0.5),
    legend.margin = margin(3, 3, 3, 3)
  )

pdf(file = file.path(here(), "figures", "landscape", "figure_S2a_signed.pdf"),
    height = 8, width= 9)
print(p)
dev.off()
```





